---
title: "Spatial Evaluation - Full RF model of raw CSE, alternative km projections"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/knitted_mds/" 
)
```

RStudio Configuration:  
   - **R version:** R 4.4.0 (Geospatial packages)  
   - **Number of cores:** 4 (up to 32 available)   
   - **Account:** saarman-np  
   - **Partition:** saarman-np (now auto allows multiple simultaneous jobs)  
   - **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)
  
  

## Load libraries
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(randomForest)
```
## Inputs
  - `../input/Gff_cse_envCostPaths.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2), same input used for building model.
  - `../results_dir/fullRF_CSE.tif` - Final full model projected estimates of CSE at 1 pixel geodist.
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
  
## Outputs   
Major outputs from this script are saved in two locations:  
**Small text outputs saved to results_dir within git repo:**
 - `spatial_eval_rawCSE_LCPpred.csv` Pairwise least-cost path summaries for each site pair: LCP_mean, LCP_sum.
 - `spatial_eval_rawCSE_cber.csv` Pairwise circuit-based effective resistance (CBER) values for each site pair.
 - `spatial_eval_rawCSE_predictions.csv` Combined pairwise evaluation table containing observed CSE and spatial predictions for each site pair: CSE, LCP_mean, LCP_sum, CBER.
 - `spatial_eval_rawCSE_eval_metrics.csv` Summary table of evaluation metrics for each spatial summary method (LCP_mean, LCP_sum, CBER): RMSE, R², MSE, MAE, Correlation.

**Large or reusable outputs saved to results_dir_big:**
 - `least_cost_paths/LC_paths_fullRF_rawCSE.shp` Saved least-cost-path spatial lines for all evaluated site pairs. This is a reusable spatial output and does not need to be recomputed unless the resistance surface changes.
 - `spatial_eval_rawCSE_LCPpred.csv` Backup copy of least-cost path summaries.
 - `spatial_eval_rawCSE_cber.csv` Backup copy of circuit-based effective resistance (CBER) results.
 - `spatial_eval_rawCSE_predictions.rds` R object version of the combined evaluation table (eval_results) for quick reloading in later scripts.
 - `spatial_eval_rawCSE_predictions.csv` Backup copy of the combined evaluation table saved in the large-results directory.
 - `spatial_eval_rawCSE_predictions_calibrated.rds` R object version of the evaluation table with calibrated predictions (eval_df), including LCP_mean_cal, LCP_sum_cal, and CBER_cal.
 - `spatial_eval_rawCSE_eval_metrics.rds` R object version of the evaluation metrics table (metrics_df) for quick reloading in later scripts.
 - `spatial_eval_rawCSE_eval_metrics.csv` Backup copy of the evaluation metrics table saved in the large-results directory.

# Code outline

Estimate evaluation metrics for multiple projection   
Sum across least-cost path (sum LCP)  

Metrics to calculate for this screening step: 

Primary metrics:  
1. correlation (Correlation here answers: does the method recover relative ordering?)
2. R² (1 - SSE/SST, after correcting for scale mismatch) 

Secondary evaluation metrics:  
3. RMSE (after correcting for scale mismatch)
4. MSE (after correcting for scale mismatch) 
5. MAE(after correcting for scale mismatch) 

NOTE: RMSE, MSE, MAE, and R2 after scaling answer: after correcting for simple scale mismatch, how well does the method recover observed CSE?). These screening metrics are used only to decide which response formulation and spatial summary to carry forward into the final LOPOCV validation framework and permutation-based significance testing.

# Step 1: Inputs
```{r directories-inputs}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

######################################
# What surface are you using?
surf_file  <- file.path(results_dir_big, "fullRF_CSE.tif")
########################################

lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)

# Load env stack with named layers
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Neutralize sampling layer to average
env$samp_20km <- 1.027064e-11 #neutralize sampling bias

# Load rdf of final model
rf_predicted <- readRDS(file.path(results_dir_big, "rf_mean_full_tuned.rds"))
```

# Step 2. Functions

```{r kdist}
# --------------------------------------------------
# Screening: compare projection distances using LCP_sum only
# --------------------------------------------------

# projection distances to test
proj_km <- c(1, 2, 5, 10, 20, 50)

# distance layer name in model
dist_layer <- "geo_dist"

# keep only within-cluster comparisons
cse_df_eval <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates
sites_coords <- bind_rows(
  cse_df_eval %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df_eval %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# pair table
site_pairs <- cse_df_eval %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# calibration function
calibrate_pred <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  df <- data.frame(obs = obs[keep], pred = pred[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(obs))
  out[keep] <- predict(fit, newdata = data.frame(pred = pred[keep]))
  out
}

# metrics function
# correlation uses raw predictions
# RSQ and error metrics use calibrated predictions
calc_metrics <- function(obs, pred_raw, pred_cal) {
  keep <- complete.cases(obs, pred_raw, pred_cal)

  obs      <- obs[keep]
  pred_raw <- pred_raw[keep]
  pred_cal <- pred_cal[keep]

  sse  <- sum((obs - pred_cal)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred_cal)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred_cal))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred_raw, method="spearman")

  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Pearson = corr
  )
}

# output directory
screen_dir <- file.path(results_dir_big, "projection_screening_LCPsum")
dir.create(screen_dir, showWarnings = FALSE, recursive = TRUE)

all_metrics <- list()
```

# Step 3. Test 1 km versus 5 km
```{r test}
kdist <- 1
env_k1 <- env
env_k1[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k1[["pix_dist"]]))
pred1 <- predict(env_k1, rf_predicted, type = "response")

kdist <- 5
env_k5 <- env
env_k5[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k5[["pix_dist"]]))
pred5 <- predict(env_k5, rf_predicted, type = "response")

cellStats(abs(pred1 - pred5), stat = "max", na.rm = TRUE)
cellStats(abs(pred1 - pred5), stat = "mean", na.rm = TRUE)

kdist <- 150
env_k5 <- env
env_k5[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k5[["pix_dist"]]))
pred5 <- predict(env_k5, rf_predicted, type = "response")

cellStats(abs(pred1 - pred5), stat = "max", na.rm = TRUE)
cellStats(abs(pred1 - pred5), stat = "mean", na.rm = TRUE)
```

1 vs 5 are identical. 1 vs 150 km are not. Where is the split?

# Step 4. Find the Split
```{r find-split, eval = FALSE}
kvals <- c(1, 2, 5, 10, 20, 50, 100, 150)

pred_list <- lapply(kvals, function(k) {
  env_k <- env
  env_k[["pix_dist"]] <- setValues(env[[1]], k)
  predict(env_k, rf_predicted, type = "response")
})

names(pred_list) <- paste0("k", kvals)

expand.grid(i = seq_along(kvals), j = seq_along(kvals)) |>
  dplyr::filter(i < j) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    k1 = kvals[i],
    k2 = kvals[j],
    mean_diff = cellStats(abs(pred_list[[i]] - pred_list[[j]]), stat = "mean", na.rm = TRUE),
    max_diff  = cellStats(abs(pred_list[[i]] - pred_list[[j]]), stat = "max", na.rm = TRUE)
  ) |>
  dplyr::select(k1, k2, mean_diff, max_diff)
```

1, 2, 5 km are identical
mean_diff = 0
max_diff = 0
→ these produce exactly the same raster
10 km is barely different
mean diff ~ 4e-05
→ essentially negligible
20 km = small change
mean diff ~ 3e-04
→ still tiny
50+ km = real change
mean diff jumps to ~ 4e-03 → 1e-02 → 4e-02
→ now you’re actually moving into different RF partitions

# Step 5. Loop through 1-100 km

```{r loop, eval = FALSE}
proj_km <- c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)

n_cores <- 16
cl <- makeCluster(n_cores)
registerDoParallel(cl)

for (kdist in proj_km) {

  cat("\n----------------------------\n")
  cat("Running projection at", kdist, "km\n")
  cat("----------------------------\n")

  # -----------------------------------
  # Step 1: project surface at kdist
  # -----------------------------------
  env_k <- env

  # assign constant raster for projection distance
  env_k[["pix_dist"]] <- setValues(env[[1]], kdist)

  cse_surface <- predict(env_k, rf_predicted, type = "response")

  # align lake mask if needed
  if (!compareRaster(cse_surface, lake_mask,
                     extent = TRUE, rowcol = TRUE,
                     crs = TRUE, res = TRUE,
                     stopiffalse = FALSE)) {
    lake_mask_use <- projectRaster(lake_mask, cse_surface, method = "ngb")
  } else {
    lake_mask_use <- lake_mask
  }

  # assign lakes high cost
  cse_surface[lake_mask_use[] == 1] <- max(values(cse_surface), na.rm = TRUE)

  # save projected surface
  writeRaster(
    cse_surface,
    filename = file.path(screen_dir, paste0("fullRF_CSE_", kdist, "km.tif")),
    format = "GTiff",
    overwrite = TRUE
  )

  # -----------------------------------
  # Step 2: prepare sites in raster CRS
  # -----------------------------------
  sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
    st_transform(crs(cse_surface))

  sites_sp <- as(sites_df, "Spatial")
  site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

  # -----------------------------------
  # Step 3: build transition object
  # -----------------------------------
  resistance_rast <- cse_surface
  min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
  resistance_rast[resistance_rast <= 0] <- min_pos

  conductance_rast <- 1 / resistance_rast
  tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)
  tr_lcp <- geoCorrection(tr, type = "c")

  # -----------------------------------
  # Step 4: recompute least-cost paths
  # -----------------------------------

  paths_list <- foreach(ii = seq_len(nrow(site_pairs)),
                        .packages = c("gdistance", "sp", "sf")) %dopar% {
    i <- site_index[site_pairs$Var1[ii]]
    j <- site_index[site_pairs$Var2[ii]]

    path <- tryCatch({
      shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
    }, error = function(e) NULL)

    if (!is.null(path)) {
      path_sf <- st_as_sf(path)
      path_sf$id <- site_pairs$id[ii]
      return(path_sf)
    } else {
      return(NULL)
    }
  }

  paths_list <- paths_list[!sapply(paths_list, is.null)]
  paths_sf <- do.call(rbind, paths_list)
  st_crs(paths_sf) <- st_crs(cse_surface)

  # extract raster values along each path
  path_vals <- raster::extract(cse_surface, as(paths_sf, "Spatial"))

  lcp_summary <- data.frame(
    id = paths_sf$id,
    LCP_sum = sapply(path_vals, function(x) {
      if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
    })
  )

  # -----------------------------------
  # Step 5: combine and calibrate
  # -----------------------------------
  eval_results <- site_pairs %>%
    left_join(lcp_summary, by = "id")

  eval_results$LCP_sum_cal <- calibrate_pred(eval_results$CSE, eval_results$LCP_sum)

  # -----------------------------------
  # Step 6: calculate metrics
  # -----------------------------------
  metrics_k <- calc_metrics(
    obs = eval_results$CSE,
    pred_raw = eval_results$LCP_sum,
    pred_cal = eval_results$LCP_sum_cal
  ) %>%
    mutate(
      projection_km = kdist,
      method = "LCP_sum"
    ) %>%
    dplyr::select(projection_km, method, everything())

  # save outputs
  write.csv(
    eval_results,
    file.path(screen_dir, paste0("eval_results_LCPsum_", kdist, "km.csv")),
    row.names = FALSE
  )

  write.csv(
    metrics_k,
    file.path(screen_dir, paste0("eval_metrics_LCPsum_", kdist, "km.csv")),
    row.names = FALSE
  )

  all_metrics[[paste0("proj_", kdist, "km")]] <- metrics_k
}

stopCluster(cl)

# bind all metrics
metrics_all <- bind_rows(all_metrics)

write.csv(
  metrics_all,
  file.path(screen_dir, "projection_distance_comparison_metrics_LCPsum.csv"),
  row.names = FALSE
)

metrics_all
```

# Step 6. Visualize
```{r review-km-projections}
km_summary<- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/projection_screening_LCPsum/projection_distance_comparison_metrics_LCPsum.csv")
print(km_summary)

plot(
  km_summary$projection_km,
  km_summary$RSQ,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "% Variance Explained"
)

plot(
  km_summary$projection_km,
  km_summary$MSE,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "OOB MSE"
)

plot(
  km_summary$projection_km,
  km_summary$Pearson,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "Mean absolute raster change"
)
```

```{r plot-to-pdf, eval = FALSE}
# Open PDF device
pdf("/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/projections_km_fig.pdf", width = 7, height = 3)

# Set up 1 row, 3 columns
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))  # adjust margins if needed

# (a) % Variance Explained
plot(
  km_summary$projection_km,
  km_summary$RSQ,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "% Variance Explained"
)
mtext("(a)", side = 3, adj = 0, line = 0.5, font = 2)

# (b) OOB MSE
plot(
  km_summary$projection_km,
  km_summary$MSE,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "OOB MSE"
)
mtext("(b)", side = 3, adj = 0, line = 0.5, font = 2)

# (c) Pearson
plot(
  km_summary$projection_km,
  km_summary$Pearson,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "Pearson's Correlation"
)
mtext("(c)", side = 3, adj = 0, line = 0.5, font = 2)

# Close PDF device
dev.off()

# Reset plotting layout (optional)
par(mfrow = c(1, 1))
```
  
# Step 7. Response formulations and spatial evaluation tests

## CSE isolation-by-distance residuals  

This is for justification of choice of one response formulation (raw CSE) and spatial summary (LCP sum).

This is a screening step (full model only). Use the fitted full model to compare candidate response formulations and spatial summaries.  

Response formulations to compare:  
- raw CSE (above)  
- residuals from simple IBD model  
- CSE/km 

Run spatial evaluation across the residuals resistance surface (projected estimates of residuals) to estimate evaluation metrics by 3 methods:   
1. Mean across least-cost path (mean LCP)   
2. Sum across least-cost path (sum LCP)  
3. Circuit-based effective resistance (CBER)  

Metrics to calculate for this screening step: 

Primary metrics:  
1. correlation (Correlation here answers: does the method recover relative ordering?)
2. R² (1 − SSE/SST, after correcting for scale mismatch) 

Secondary evaluation metrics:  
3. RMSE (after correcting for scale mismatch)
4. MSE (after correcting for scale mismatch) 
5. MAE(after correcting for scale mismatch) 

NOTE: RMSE, MSE, MAE, and R2 after scaling answer: after correcting for simple scale mismatch, how well does the method recover observed CSE?). These screening metrics are used only to decide which response formulation and spatial summary to carry forward into the final LOPOCV validation framework and permutation-based significance testing.


### Load libraries
```{r libraries-ibd, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

### 1: Inputs
```{r directories-inputs-ibd}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

######################################
# What surface are you using?
surf_file  <- file.path(results_dir_big, "fullRF_residuals.tif")
########################################

lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")


# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)
```
  
### 2: Check raster alignment and assign high cost to water

```{r check-alignment-ibd, eval = FALSE}
if (!compareRaster(cse_surface, lake_mask,
                   extent = TRUE, rowcol = TRUE,
                   crs = TRUE, res = TRUE,
                   stopiffalse = FALSE)) {
  lake_mask <- projectRaster(lake_mask, cse_surface, method = "ngb")
}

# Inspect projected raw CSE values before altering water cells
summary(values(cse_surface))

# Keep water traversible but very costly (max value from model)
cse_surface[lake_mask[] == 1] <- max(values(cse_surface))

# Check again after assigning lakes
summary(values(cse_surface))

# Quick plot
plot(cse_surface, main = "Projected raw CSE surface with water = high cost")

# Check column names of cse_df
names(cse_df)
```

### 3: Prepare pair table, site objects, and transition object  

Diagonal movement is allowed in the 16-way transition matrix, and conductance is corrected for geographic distance, with geoCorrection(type = "c"), so diagonal steps are appropriately penalized relative to orthogonal steps.

```{r prep-transition-ibd, eval = FALSE}
# keep only within-cluster comparisons, matching your earlier workflow
cse_df <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates from both columns
sites_coords <- bind_rows(
  cse_df %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# convert to sf and then to SpatialPoints in raster CRS
sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(cse_surface))

sites_sp <- as(sites_df, "Spatial")

# lookup index for site names
site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

# pair table for evaluation
site_pairs <- cse_df %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# treat projected raw CSE as local resistance
resistance_rast <- cse_surface

# protect against zero or negative values before inversion
min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
resistance_rast[resistance_rast <= 0] <- min_pos

# convert resistance to conductance for gdistance
conductance_rast <- 1 / resistance_rast

# transition object:
# directions = 16 allows diagonal/extended neighborhood movement
tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)

# correction for least-cost paths
# geoCorrection(type = "c") corrects with longer diagonal
tr_lcp <- geoCorrection(tr, type = "c")

# correction for circuit / random-walk distances
# geoCorrection(type = "r") corrects for random walk
tr_cber <- geoCorrection(tr, type = "r")
```

### 4: Sanity checks
```{r sanity-ibd, eval = FALSE}
nrow(cse_df)
nrow(site_pairs)
nrow(sites_coords)

head(site_pairs)

plot(cse_surface, main = "Projected raw CSE surface")
plot(sites_sp, add = TRUE, pch = 16, cex = 0.5)
```  

### 5: Compute least-cost paths in parallel (once) and save output  
Test one path first:
```{r test-lcp-ibd, eval = FALSE}
# Test one least-cost path before parallel run
i <- site_index[site_pairs$Var1[100]]
j <- site_index[site_pairs$Var2[100]]

test_path <- shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")

plot(cse_surface, main = "Test least-cost path")
lines(test_path, col = "red", lwd = 2)
points(sites_sp[c(i, j), ], pch = 16)
```  
Run full loop:
```{r save-lcp-ibd, eval = FALSE}
output_dir <- file.path(results_dir_big, "least_cost_paths")
dir.create(output_dir, showWarnings = FALSE)

# Set number of cores and register cluster
n_cores <- 16
cl <- makeCluster(n_cores)
registerDoParallel(cl)

paths_list <- foreach(k = 1:nrow(site_pairs),
                      .packages = c("gdistance", "sp", "sf")) %dopar% {

  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]

  path <- tryCatch({
    shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

stopCluster(cl)

# drop NULLs and combine
paths_list <- paths_list[!sapply(paths_list, is.null)]
paths_sf <- do.call(rbind, paths_list)

# assign CRS and transform back to geographic coordinates for saving
st_crs(paths_sf) <- st_crs(cse_surface)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# join observed CSE values
paths_sf <- left_join(paths_sf, site_pairs, by = "id")

# save paths
st_write(paths_sf,
         file.path(output_dir, "LC_paths_fullRF_residuals.shp"),
         delete_layer = TRUE)
```

### 7: Extract predicted CSE values along LCPs and calculate mean and sum
```{r extract-mean-sum-ibd, eval = FALSE}
# if needed later, reload:
paths_sf <- st_read(file.path(results_dir_big, "least_cost_paths", "LC_paths_fullRF_residuals.shp"), quiet = TRUE)

# transform paths to raster CRS for extraction
paths_extract <- st_transform(paths_sf, crs = crs(cse_surface))

# extract raster values along each path
path_vals <- raster::extract(cse_surface, as(paths_extract, "Spatial"))

# summarize values along each least-cost path
lcp_summary <- data.frame(
  id = paths_extract$id,
  LCP_mean = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
  }),
  LCP_sum = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
  })
)

# join back to pair table
LCP_results <- site_pairs %>%
  left_join(lcp_summary, by = "id")

head(LCP_results)
summary(LCP_results$LCP_mean)
summary(LCP_results$LCP_sum)

# Save LCP output as text file in results directories
write.csv(LCP_results,file.path(results_dir, "spatial_eval_residuals_LCPpred.csv"), row.names = FALSE)

write.csv(LCP_results,file.path(results_dir_big, "spatial_eval_residuals_LCPpred.csv"), row.names = FALSE)

```  

### 8: Compute circuit-based effective resistance (CBER)  
Test one CBER calculation first:
```{r cber-test-ibd, eval = FALSE}
# test one CBER calculation first
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])

as.numeric(dmat_test)[1] # returns the single number for pair
```
Run as a loop sequentially, also save output from steps 6-8 to file again...

```{r cber_calculations-ibd, eval = FALSE}
library(doParallel)
library(foreach)
library(gdistance)
library(sp)

# number of cores
n_cores <- 16

# on Linux/CHPC, FORK is usually more efficient for large objects like tr_cber
cl <- parallel::makeForkCluster(n_cores)
doParallel::registerDoParallel(cl)

# optional: quick test outside parallel
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
as.numeric(dmat_test)[1]

# parallel loop across all 1091 pairs
cber_results <- foreach(
  k = seq_len(nrow(site_pairs)),
  .combine = rbind,
  .packages = c("gdistance", "sp")
) %dopar% {
  
  site1 <- site_pairs$Var1[k]
  site2 <- site_pairs$Var2[k]
  pair_id <- site_pairs$id[k]
  
  i <- site_index[site1]
  j <- site_index[site2]
  
  # compute commute distance for the pair
  dmat <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
  
  # extract the single numeric value from the dist object
  cber_val <- as.numeric(dmat)[1]
  
  data.frame(
    Var1 = site1,
    Var2 = site2,
    id   = pair_id,
    CBER = cber_val
  )
}

# stop cluster
stopCluster(cl)

# inspect
head(cber_results)
str(cber_results)

# save output in results directories
write.csv(cber_results,file.path(results_dir, "spatial_eval_residuals_cber.csv"), row.names = FALSE)

write.csv(cber_results,file.path(results_dir_big, "spatial_eval_residuals_cber.csv"), row.names = FALSE)

```

### 9: Combine LCP and CBER results
```{r combine-ibd, eval = FALSE}
# load and bind your existing evaluation data
library(dplyr)

# load the LCP sum and mean results you just computed
LCP_results <- read.csv(file.path(results_dir, "spatial_eval_residuals_LCPpred.csv"))

# bind CBER and LCP results
eval_results <- LCP_results %>%
  left_join(
    cber_results %>% select(id, CBER),
    by = "id"
  )

# Save results output as RDS for later ease of loading
saveRDS(
  eval_results,
  file.path(results_dir_big, "spatial_eval_residuals_predictions.rds")
)
```

### 10: Calculate evaluation metrics from eval_results
```{r calibrate-and-metrics-ibd, eval = FALSE}
library(dplyr)
library(ggplot2)

# Start eval_df from eval_results
eval_df <- eval_results

# Fit linear calibration and return calibrated predictions
calibrate_pred <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  df <- data.frame(obs = obs[keep], pred = pred[keep])
  
  fit <- lm(obs ~ pred, data = df)
  
  out <- rep(NA_real_, length(obs))
  out[keep] <- predict(fit, newdata = data.frame(pred = pred[keep]))
  out
}

# Create calibrated predictions
eval_df$LCP_sum_cal  <- calibrate_pred(eval_df$CSE, eval_df$LCP_sum)
eval_df$LCP_mean_cal <- calibrate_pred(eval_df$CSE, eval_df$LCP_mean)
eval_df$CBER_cal     <- calibrate_pred(eval_df$CSE, eval_df$CBER)

# Plots: raw predicted vs observed, with fitted line
plot1 <- ggplot(eval_df, aes(x = LCP_sum, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP sum, raw scale)",
    y = "Observed CSE"
  )
plot1

plot2 <- ggplot(eval_df, aes(x = LCP_mean, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP mean, raw scale)",
    y = "Observed CSE"
  )
plot2

plot3 <- ggplot(eval_df, aes(x = CBER, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (CBER, raw scale)",
    y = "Observed CSE"
  )
plot3

# Calculate formal metrics
# Correlation is calculated on raw predictions because it is scale-invariant
# Error metrics and R2 are calculated on calibrated predictions
calc_metrics <- function(obs, pred_raw, pred_cal) {
  keep <- complete.cases(obs, pred_raw, pred_cal)
  obs      <- obs[keep]
  pred_raw <- pred_raw[keep]
  pred_cal <- pred_cal[keep]
  
  sse  <- sum((obs - pred_cal)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred_cal)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred_cal))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred_raw)
  
  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Correlation = corr
  )
}

metrics_df <- bind_rows(
  calc_metrics(eval_df$CSE, eval_df$LCP_mean, eval_df$LCP_mean_cal) %>%
    mutate(method = "LCP_mean"),
  
  calc_metrics(eval_df$CSE, eval_df$LCP_sum, eval_df$LCP_sum_cal) %>%
    mutate(method = "LCP_sum"),
  
  calc_metrics(eval_df$CSE, eval_df$CBER, eval_df$CBER_cal) %>%
    mutate(method = "CBER")
) %>%
  dplyr::select(method, everything())

# Print metrics
print(metrics_df)

# Save outputs
saveRDS(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_residuals_eval_metrics.rds")
)

write.csv(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_residuals_eval_metrics.csv"),
  row.names = FALSE
)

write.csv(
  metrics_df,
  file.path(results_dir, "spatial_eval_residuals_eval_metrics.csv"),
  row.names = FALSE
)
```

Print to knitted file: 

```{r print-ibd, eval = TRUE}
ibd_metrics_df <- read.csv(file.path(results_dir_big, "spatial_eval_residuals_eval_metrics.csv"))

# Print metrics
print(ibd_metrics_df)

```
## CSE per kilometer (CSE/km) for comparison

Spatial Evaluation - CSE per km (rate, LC lakes paths)

### Load libraries
```{r libraries-rate, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

### 1: Inputs
```{r directories-inputs-rate}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

######################################
# What surface are you using?
surf_file  <- file.path(results_dir_big, "fullRF_rate.tif")
########################################

lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")


# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)
```
  
### 2: Check raster alignment and assign high cost to water

```{r check-alignment-rate, eval = FALSE}
if (!compareRaster(cse_surface, lake_mask,
                   extent = TRUE, rowcol = TRUE,
                   crs = TRUE, res = TRUE,
                   stopiffalse = FALSE)) {
  lake_mask <- projectRaster(lake_mask, cse_surface, method = "ngb")
}

# Inspect projected raw CSE values before altering water cells
summary(values(cse_surface))

# Keep water traversible but very costly (max value from model)
cse_surface[lake_mask[] == 1] <- max(values(cse_surface))

# Check again after assigning lakes
summary(values(cse_surface))

# Quick plot
plot(cse_surface, main = "Projected raw CSE surface with water = high cost")

# Check column names of cse_df
names(cse_df)
```

### 3: Prepare pair table, site objects, and transition object  

Diagonal movement is allowed in the 16-way transition matrix, and conductance is corrected for geographic distance, with geoCorrection(type = "c"), so diagonal steps are appropriately penalized relative to orthogonal steps.

```{r prep-transition-rate, eval = FALSE}
# keep only within-cluster comparisons, matching your earlier workflow
cse_df <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates from both columns
sites_coords <- bind_rows(
  cse_df %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# convert to sf and then to SpatialPoints in raster CRS
sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(cse_surface))

sites_sp <- as(sites_df, "Spatial")

# lookup index for site names
site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

# pair table for evaluation
site_pairs <- cse_df %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# treat projected raw CSE as local resistance
resistance_rast <- cse_surface

# protect against zero or negative values before inversion
min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
resistance_rast[resistance_rast <= 0] <- min_pos

# convert resistance to conductance for gdistance
conductance_rast <- 1 / resistance_rast

# transition object:
# directions = 16 allows diagonal/extended neighborhood movement
tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)

# correction for least-cost paths
# geoCorrection(type = "c") corrects with longer diagonal
tr_lcp <- geoCorrection(tr, type = "c")

# correction for circuit / random-walk distances
# geoCorrection(type = "r") corrects for random walk
tr_cber <- geoCorrection(tr, type = "r")
```

### 4: Sanity checks
```{r sanity-rate, eval = FALSE}
nrow(cse_df)
nrow(site_pairs)
nrow(sites_coords)

head(site_pairs)

plot(cse_surface, main = "Projected raw CSE surface")
plot(sites_sp, add = TRUE, pch = 16, cex = 0.5)
```  

### 5: Compute least-cost paths in parallel (once) and save output  
Test one path first:
```{r test-lcp-rate, eval = FALSE}
# Test one least-cost path before parallel run
i <- site_index[site_pairs$Var1[100]]
j <- site_index[site_pairs$Var2[100]]

test_path <- shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")

plot(cse_surface, main = "Test least-cost path")
lines(test_path, col = "red", lwd = 2)
points(sites_sp[c(i, j), ], pch = 16)
```  
Run full loop:
```{r save-lcp}
output_dir <- file.path(results_dir_big, "least_cost_paths")
dir.create(output_dir, showWarnings = FALSE)

# Set number of cores and register cluster
n_cores <- 16
cl <- makeCluster(n_cores)
registerDoParallel(cl)

paths_list <- foreach(k = 1:nrow(site_pairs),
                      .packages = c("gdistance", "sp", "sf")) %dopar% {

  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]

  path <- tryCatch({
    shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

stopCluster(cl)

# drop NULLs and combine
paths_list <- paths_list[!sapply(paths_list, is.null)]
paths_sf <- do.call(rbind, paths_list)

# assign CRS and transform back to geographic coordinates for saving
st_crs(paths_sf) <- st_crs(cse_surface)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# join observed CSE values
paths_sf <- left_join(paths_sf, site_pairs, by = "id")

# save paths
st_write(paths_sf,
         file.path(output_dir, "LC_paths_fullRF_rate.shp"),
         delete_layer = TRUE)
```

### 7: Extract predicted CSE values along LCPs and calculate mean and sum
```{r extract-mean-sum-rate, eval = FALSE}
# if needed later, reload:
paths_sf <- st_read(file.path(results_dir_big, "least_cost_paths", "LC_paths_fullRF_rate.shp"), quiet = TRUE)

# transform paths to raster CRS for extraction
paths_extract <- st_transform(paths_sf, crs = crs(cse_surface))

# extract raster values along each path
path_vals <- raster::extract(cse_surface, as(paths_extract, "Spatial"))

# summarize values along each least-cost path
lcp_summary <- data.frame(
  id = paths_extract$id,
  LCP_mean = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
  }),
  LCP_sum = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
  })
)

# join back to pair table
LCP_results <- site_pairs %>%
  left_join(lcp_summary, by = "id")

head(LCP_results)
summary(LCP_results$LCP_mean)
summary(LCP_results$LCP_sum)

# Save LCP output as text file in results directories
write.csv(LCP_results,file.path(results_dir, "spatial_eval_rate_LCPpred.csv"), row.names = FALSE)

write.csv(LCP_results,file.path(results_dir_big, "spatial_eval_rate_LCPpred.csv"), row.names = FALSE)

```  

### 8: Compute circuit-based effective resistance (CBER)  
Test one CBER calculation first:
```{r cber-test-rate, eval = FALSE}
# test one CBER calculation first
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])

as.numeric(dmat_test)[1] # returns the single number for pair
```
Run as a loop sequentially, also save output from steps 6-8 to file again...

```{r cber_calculations-rate, eval = FALSE}
library(doParallel)
library(foreach)
library(gdistance)
library(sp)

# number of cores
n_cores <- 16

# on Linux/CHPC, FORK is usually more efficient for large objects like tr_cber
cl <- parallel::makeForkCluster(n_cores)
doParallel::registerDoParallel(cl)

# optional: quick test outside parallel
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
as.numeric(dmat_test)[1]

# parallel loop across all 1091 pairs
cber_results <- foreach(
  k = seq_len(nrow(site_pairs)),
  .combine = rbind,
  .packages = c("gdistance", "sp")
) %dopar% {
  
  site1 <- site_pairs$Var1[k]
  site2 <- site_pairs$Var2[k]
  pair_id <- site_pairs$id[k]
  
  i <- site_index[site1]
  j <- site_index[site2]
  
  # compute commute distance for the pair
  dmat <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
  
  # extract the single numeric value from the dist object
  cber_val <- as.numeric(dmat)[1]
  
  data.frame(
    Var1 = site1,
    Var2 = site2,
    id   = pair_id,
    CBER = cber_val
  )
}

# stop cluster
stopCluster(cl)

# inspect
head(cber_results)
str(cber_results)

# save output in results directories
write.csv(cber_results,file.path(results_dir, "spatial_eval_rate_cber.csv"), row.names = FALSE)

write.csv(cber_results,file.path(results_dir_big, "spatial_eval_rate_cber.csv"), row.names = FALSE)

```

### 9: Combine LCP and CBER results
```{r combine-rate, eval = FALSE}
# load and bind your existing evaluation data
library(dplyr)

# load the LCP sum and mean results you just computed
LCP_results <- read.csv(file.path(results_dir, "spatial_eval_rate_LCPpred.csv"))

# bind CBER and LCP results
eval_results <- LCP_results %>%
  left_join(
    cber_results %>% select(id, CBER),
    by = "id"
  )

# Save results output as RDS for later ease of loading
saveRDS(
  eval_results,
  file.path(results_dir_big, "spatial_eval_rate_predictions.rds")
)
```

### 10: Calculate evaluation metrics from eval_results
```{r calibrate-and-metrics-rate, eval = FALSE}
library(dplyr)
library(ggplot2)

# Start eval_df from eval_results
eval_df <- eval_results

# Fit linear calibration and return calibrated predictions
calibrate_pred <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  df <- data.frame(obs = obs[keep], pred = pred[keep])
  
  fit <- lm(obs ~ pred, data = df)
  
  out <- rep(NA_real_, length(obs))
  out[keep] <- predict(fit, newdata = data.frame(pred = pred[keep]))
  out
}

# Create calibrated predictions
eval_df$LCP_sum_cal  <- calibrate_pred(eval_df$CSE, eval_df$LCP_sum)
eval_df$LCP_mean_cal <- calibrate_pred(eval_df$CSE, eval_df$LCP_mean)
eval_df$CBER_cal     <- calibrate_pred(eval_df$CSE, eval_df$CBER)

# Plots: raw predicted vs observed, with fitted line
plot1 <- ggplot(eval_df, aes(x = LCP_sum, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP sum, raw scale)",
    y = "Observed CSE"
  )
plot1

plot2 <- ggplot(eval_df, aes(x = LCP_mean, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP mean, raw scale)",
    y = "Observed CSE"
  )
plot2

plot3 <- ggplot(eval_df, aes(x = CBER, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (CBER, raw scale)",
    y = "Observed CSE"
  )
plot3

# Calculate formal metrics
# Correlation is calculated on raw predictions because it is scale-invariant
# Error metrics and R2 are calculated on calibrated predictions
calc_metrics <- function(obs, pred_raw, pred_cal) {
  keep <- complete.cases(obs, pred_raw, pred_cal)
  obs      <- obs[keep]
  pred_raw <- pred_raw[keep]
  pred_cal <- pred_cal[keep]
  
  sse  <- sum((obs - pred_cal)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred_cal)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred_cal))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred_raw)
  
  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Correlation = corr
  )
}

metrics_df <- bind_rows(
  calc_metrics(eval_df$CSE, eval_df$LCP_mean, eval_df$LCP_mean_cal) %>%
    mutate(method = "LCP_mean"),
  
  calc_metrics(eval_df$CSE, eval_df$LCP_sum, eval_df$LCP_sum_cal) %>%
    mutate(method = "LCP_sum"),
  
  calc_metrics(eval_df$CSE, eval_df$CBER, eval_df$CBER_cal) %>%
    mutate(method = "CBER")
) %>%
  dplyr::select(method, everything())
 
print(rate_metrics_df)

# Save outputs
saveRDS(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_rate_eval_metrics.rds")
)

write.csv(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_rate_eval_metrics.csv"),
  row.names = FALSE
)

write.csv(
  metrics_df,
  file.path(results_dir, "spatial_eval_rate_eval_metrics.csv"),
  row.names = FALSE
)
```

# Print to knitted file:

```{r print-rate, eval = TRUE}
rate_metrics_df <- read.csv(file.path(results_dir_big, "spatial_eval_rate_eval_metrics.csv"))

# Print metrics
print(rate_metrics_df)

```

