---
title: "Uganda tsetse PopGen: PCA, diversity, CSE"
author: "Norah Saarman"
date: "2025-06-03"
output: pdf_document
---

RStudio Configuration:
   - **R version:** R 4.4.0 (Geospatial packages)  
   - **Number of cores:** 4 (up to 32 available)   
   - **Account:** saarman-np  
   - **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
   - **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
library(adegenet)
library(geosphere)
library(igraph)
library(scatterplot3d)
library(ggplot2)
library(dplyr)
library(hierfstat)
library(pegas)
library(PopGenReport)
library(poppr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(cowplot)
library(viridis)
```
# Importing data as genind object from .gen
```{r genind}
# 1. Genepop .gen file "./input/Gff_11loci_allsites_genepop.gen" 
Gff.genind <- read.genepop("../input/Gff_11loci_allsites_genepop.gen", ncode = 3)

# Metadata with base R function read.csv()
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", header=TRUE, sep = "\t")

# Add site code to @pop slot
Gff.genind@pop <- as.factor(Gff$SiteCode)

# Check genind object
Gff.genind
summary(Gff.genind)
```
# Visualize genetic variation with PCA  

All samples:
```{r pca-all}
# replace NAs with locus means for PCA
gen_tab <- tab(Gff.genind, NA.method = "mean")  

# Compute PCA on the subset
pca_result3 <- dudi.pca(gen_tab, cent = FALSE, scale = TRUE, scannf = FALSE, nf = 4)

# Extract percent variance explained for PC1 and PC2
var_explained <- round(100 * pca_result3$eig / sum(pca_result3$eig), 1)

# Generate 86 visually distinct colors
set.seed(123)  # for reproducibility
n_pops <- nlevels(pop(Gff.genind))
color_palette <- colorRampPalette(c("red", "orange", "yellow", "green", "cyan", "blue", "purple"))(n_pops)

# Map population levels to colors
site_levels <- levels(pop(Gff.genind))
site_colors <- setNames(color_palette, site_levels)

# Get color for each individual
pop_colors <- site_colors[as.character(pop(Gff.genind))]

# Layout: 2 panels (main plot + legend)
layout(matrix(c(1, 2), nrow = 1), widths = c(6, 2.5))

# PCA plot
#pdf("../figures/PCA-all-2D.pdf", width = 14, height = 10)
par(mar = c(5, 4, 4, 2))
plot(pca_result3$li[, 1:2],
     col = pop_colors,
     pch = 19,
     xlab = paste0("PC1 (", var_explained[1], "%)"),
     ylab = paste0("PC2 (", var_explained[2], "%)"))
# Legend panel
par(mar = c(1,1,1,1))  # increase space to avoid clipping
plot.new()
legend("center", 
       legend = site_levels,
       col = site_colors,
       pch = 19,
       cex = 0.6,
       title = "Populations", 
       ncol = 3,
       bty = "n")  # <- no border
#dev.off()

# 3D PCA plot
#pdf("../figures/PCA-all-3D.pdf", width = 14, height = 10)
par(mar = c(5, 4, 4, 2))
s3d <- scatterplot3d(pca_result3$li[, c(1,3,2)],
                     color = pop_colors,
                     pch = 19,
                     xlab = paste0("PC1 (", var_explained[1], "%)"), # horizontal 
                     ylab = paste0("PC3 (", var_explained[3], "%)"), # depth
                     zlab = paste0("PC2 (", var_explained[2], "%)"), # vertical
                     angle = 55,
                     box = TRUE)
# Legend panel
par(mar = c(1, 1, 1, 1))
plot.new()
legend("center",
       legend = site_levels,
       col = site_colors,
       pch = 19,
       cex = 0.6,
       title = "Populations",
       ncol = 3,
       bty = "n")
#dev.off()
```
# Considering K=2 by Evano method
```{r k2}
# Add K2 cluster combining "south" and "west" into "south"
Gff$K2 <- ifelse(Gff$SiteMajCluster %in% c("south", "west"), "south", Gff$SiteMajCluster)
```
PCA for each K2 clusters separately
```{r pca-k2}
# PCA for each K2 cluster separately
for (region in c("north","south")) {
  genind_sub <- Gff.genind[pop(Gff.genind) %in% Gff$SiteCode[Gff$K2 == region]]
  # Replace NAs with locus means for PCA
  gen_tab <- tab(genind_sub, NA.method = "mean")
  pca_result <- dudi.pca(gen_tab, cent = FALSE, scale = TRUE, scannf = FALSE, nf = 4)
  var_explained <- round(100 * pca_result$eig / sum(pca_result$eig), 1)
  # Color palette
  set.seed(123)
  n_pops <- nlevels(pop(genind_sub))
  palette <- colorRampPalette(c("red", "orange", "yellow", "green", "cyan", "blue", "purple"))(n_pops)
  site_levels <- levels(pop(genind_sub))
  site_colors <- setNames(palette, site_levels)
  pop_colors <- site_colors[as.character(pop(genind_sub))]
  layout(matrix(c(1, 2), nrow = 1), widths = c(6, 2.5))
  
  # 2D PCA plot
  #pdf(paste0("../figures/PCA-", region, "-2D.pdf"), width = 14, height = 10)
  par(mar = c(5, 4, 4, 2))
  plot(pca_result$li[, 1:2],
       col = pop_colors,
       pch = 19,
       xlab = paste0("PC1 (", var_explained[1], "%)"),
       ylab = paste0("PC2 (", var_explained[2], "%)"))
  # Legend
  par(mar = c(1, 1, 1, 1))
  plot.new()
  legend("center",
         legend = site_levels,
         col = site_colors,
         pch = 19,
         cex = 0.6,
         title = "Populations",
         ncol = 2,
         bty = "n")
  #dev.off()
  
  # 3D PCA plot
  #pdf(paste0("../figures/PCA-", region, "-3D.pdf"), width = 14, height = 10)
  par(mar = c(5, 1, 4, 2))
  scatterplot3d(x = pca_result$li[, 1], # horizontal
                y = pca_result$li[, 3], # depth
                z = pca_result$li[, 2], # vertical
                color = pop_colors,
                pch = 19,
                xlab = paste0("PC1 (", var_explained[1], "%)"),
                ylab = paste0("PC3 (", var_explained[3], "%)"),
                zlab = paste0("PC2 (", var_explained[2], "%)"),
                angle = 55,
                box = TRUE) 
  par(mar = c(1, 1, 1, 1))
  
  plot.new()
  legend("center",
         legend = site_levels,
         col = site_colors,
         pch = 19,
         cex = 0.6,
         title ="Populations",
         ncol = 2,
         bty = "n")
  
  #dev.off()
}
```

# Choosing sites for Random Forest (K=2)  

Filter out any Gff$Cluster == nogroup to avoid admixed individuals, which will inflate the genetic distance within each geographic region.  

Filtering for only largest/most complete genetic sample from each 5 km radius for gps within each Gff$SiteMajCluster independently so that we get best samples from both north and south of the river/Lake Kyoga dividing north and south. 

```{r distm}
# Create filtered copy of Gff excluding 'nogroup'
Gff_filt <- Gff[Gff$Cluster != "nogroup", ]

# Ensure Gff$SiteCode matches pop(Gff.genind)
Gff_filt$SiteCode <- as.character(Gff_filt$SiteCode)
pop_names <- as.character(pop(Gff.genind))
if (!all(pop_names == Gff_filt$SiteCode)) {
  Gff_filt <- Gff_filt[match(pop_names, Gff_filt$SiteCode), ]
  stopifnot(all(as.character(pop(Gff.genind)) == Gff_filt$SiteCode))
}

# Initialize storage
selected_sites <- character()
report_list <- list()

# Loop through each K2 cluster separately (north/south)
for (region in c("north","south")) {
  Gff_sub <- Gff_filt[Gff_filt$K2 == region, ]
  # Build coordinate matrix
  xy <- cbind(lon = Gff_sub$SiteLong, lat = Gff_sub$SiteLat)
  dist_mat <- distm(xy, fun = distHaversine)
  # Define adjacency for <2 km
  threshold_m <- 2000
  adj_mat <- dist_mat < threshold_m
  # Create undirected graph of close sites
  g <- graph_from_adjacency_matrix(adj_mat, mode = "undirected", diag = FALSE)
  clust_ids <- components(g)$membership
  # For each 5km cluster, pick the SiteCode with the most individuals in Gff.genind
  for (cid in unique(clust_ids)) {
    site_group <- Gff_sub$SiteCode[clust_ids == cid]
    counts <- table(pop(Gff.genind)[pop(Gff.genind) %in% site_group])
    best_site <- names(which.max(counts))
    selected_sites <- c(selected_sites, best_site)
    report_list[[best_site]] <- site_group
  }
}

# Subset Gff and Gff.genind to selected sites
Gff_rf <- Gff_filt[Gff_filt$SiteCode %in% selected_sites, ]
Gff.genind_rf <- Gff.genind[pop(Gff.genind) %in% selected_sites]

# Report selected SiteCodes by region
print(split(Gff_rf$SiteCode, Gff_rf$K2))

# Report groupings and selected representative from each 5km cluster
report_df <- data.frame(
  selected = names(report_list),
  group = sapply(report_list, function(x) paste(sort(unique(x)), collapse = ", ")),
  row.names = NULL
)
print(report_df)
```
Remove smaller sample size(s) of any 2-3 sites within 2 km of each other: 04-OM, 29-KO, 13-GOR, 34-AMI, 35-DK, 39-OC, 53-UWA, 77-NB, 80-SA

Brings total site count down to 75.
```{r count}
Gff.genind_rf # group size 8-236
table(pop(Gff.genind_rf))
```
Too imbalanced... 
  - Downsample 55-KAF, 78-OK, 79-BU to N individuals (e.g., 50)
  - Remove sites with fewer than n=15
  
```{r balance}
# Set thresholds
min_n <- 15
max_n <- 50

# Compute sample sizes per population
pop_sizes <- table(pop(Gff.genind_rf))

# Keep populations with n >= min_n
pops_to_keep <- names(pop_sizes[pop_sizes >= min_n])

# Indices to retain
inds_keep <- unlist(lapply(pops_to_keep, function(p) {
  inds <- which(pop(Gff.genind_rf) == p)
  if (length(inds) > max_n) {
    sample(inds, max_n)  # downsample
  } else {
    inds  # keep as-is
  }
}))

#Subset both genind and metadata
Gff.genind_rf <- Gff.genind_rf[inds_keep]
Gff_rf <- Gff_rf[inds_keep, ]

# Re-check population sizes
Gff.genind_rf
summary(Gff.genind_rf)
table(pop(Gff.genind_rf))
length(unique(Gff_rf$SiteCode))
```
Now 68 pops with group size ranges from 15-50.  

# Genetic diversity  

Estimate allelic richness (standardized/rarefied) with pegas  
```{r ar}
genind_hf <- genind2hierfstat(Gff.genind_rf)

# This function rarefies allelic richness to the smallest sample size
allelic_richness <- allelic.richness(genind_hf)

# Extract mean rarefied allelic richness per population
mean_ar <- colMeans(allelic_richness$Ar, na.rm = TRUE)
mean_ar
``` 
Most diversity stats use poppr  
```{r poppr}
# Compute per-population stats
div_stats <- poppr(Gff.genind_rf)

# View a summary table
div_stats
```
Hierfstat, separating by pop and rarefying to min samples size with 100 replications with own code:
```{r rarefy}
# Set rarefaction parameters
# Minimum number of individuals per population to include
min_n <- 15

# Number of random resampling repetitions per population
n_reps <- 100

# Split the genind object by population
# seppop() returns a list of genind objects, one for each population
pop_list <- seppop(Gff.genind_rf)

# For each pop, repeat random sampling n_reps times
# For each replicate:
# - Randomly sample min_n individuals (with no replacement)
# - Convert to hierfstat format
# - Run basic.stats()
# - Extract and average FIS, Ho (observed heterozygosity), and He (expected heterozygosity)
results <- lapply(names(pop_list), function(pname) {
  g <- pop_list[[pname]]
  # Skip populations with fewer than min_n individuals
  if (nInd(g) < min_n) return(NULL)
  # Repeat rarefied sampling n_reps times
  replicate(n_reps, {
    # Randomly sample min_n individuals
    inds <- sample(1:nInd(g), min_n)
    g_sub <- g[inds, ]
    # Convert to hierfstat format and compute basic stats
    hf <- genind2hierfstat(g_sub)
    bs <- basic.stats(hf)
    # Return vector of average values across loci
    c(
      FIS = mean(bs$Fis, na.rm = TRUE),
      Ho  = mean(bs$Ho, na.rm = TRUE),
      He  = mean(bs$Hs, na.rm = TRUE)
    )
  }) |>
    t() |>                  # transpose so each row = one replicate
    as.data.frame()         # convert to data frame for easy averaging
})

# Average results across replicates for each population
# Keep only populations where stats were successfully computed
names(results) <- names(pop_list)[sapply(results, is.data.frame)]

# Compute mean FIS, Ho, He across replicates
summary_stats <- do.call(rbind, lapply(results, function(df) {
  colMeans(df, na.rm = TRUE)
}))

# Format as a final summary table
summary_df <- data.frame(
  Population = rownames(summary_stats),
  round(summary_stats, 3),  # round to 3 decimal places
  row.names = NULL
)
```
Compile results from the three code chunks into one data frame  
```{r compile}
# Convert allelic richness (mean_ar) to a data frame
ar_df <- data.frame(
  Population = names(mean_ar),
  Allelic_Richness = as.numeric(mean_ar),
  row.names = NULL
)

# Convert poppr results (div_stats) to a data frame
#    (assumes div_stats has rownames corresponding to populations)
div_stats_df <- data.frame(
  Population = div_stats$Pop,
  div_stats[ , setdiff(names(div_stats), "Pop") ],
  row.names = NULL
)

# Merge all three data frames by "Population"
#    Use Reduce() for a sequential full join
summary_df <- Reduce(function(x, y) {
  merge(x, y, by = "Population", all = TRUE)
}, list(ar_df, div_stats_df, summary_df))

# Save the combined summary to a CSV file
#write.csv(summary_df, "../input/summary_df.csv", row.names = FALSE)

# Print the final table
print(summary_df)
```
# Cavalli-Sforza and Edwards' Distance (CSE) 
Pairwise within cluster "north"
```{r north}
# Subset genind and metadata
meta_sub <- Gff_rf[Gff_rf$K2 == "north", ]
gen_sub <- Gff.genind_rf[pop(Gff.genind_rf) %in% meta_sub$SiteCode]

# Convert to genpop (aggregates by population)
genpop_sub <- genind2genpop(gen_sub)

# Compute Cavalli-Sforza and Edwards’ distance
cse_dist <- dist.genpop(genpop_sub, method = 2)  # method = 2 (CSE distance

# Display as a matrix
cse_north <- as.matrix(cse_dist)
```
  
Pairwise within cluster "south"  
```{r south}
# Subset genind and metadata
meta_sub <- Gff_rf[Gff_rf$K2 == "south", ]
gen_sub <- Gff.genind_rf[pop(Gff.genind_rf) %in% meta_sub$SiteCode]

# Convert to genpop (aggregates by population)
genpop_sub <- genind2genpop(gen_sub)

# Compute Cavalli-Sforza and Edwards’ distance
cse_dist <- dist.genpop(genpop_sub, method = 2)  # method = 2 (CSE distance)

# Display as a matrix
cse_south <- as.matrix(cse_dist)
```
  
Combine CSE output into a single dataframe for downstream analysis:  
```{r cse}
# Create a metadata data frame for coordinates and cluster
meta_coords <- unique(Gff_rf[, c("SiteCode", "SiteLat", "SiteLong", "K2")])
names(meta_coords) <- c("SiteCode", "lat", "long", "cluster")

# Process the "north" CSE matrix
north_df <- as.data.frame(as.table(cse_north), stringsAsFactors = FALSE)
names(north_df) <- c("Var1", "Var2", "CSEdistance")

# Keep only one triangle (Var1 < Var2 alphabetically)
north_df <- subset(north_df, Var1 < Var2)

# Merge to get lat1, long1, and Pop1_cluster
north_df <- merge(
  north_df,
  meta_coords,
  by.x = "Var1",
  by.y = "SiteCode",
  all.x = TRUE,
  suffixes = c("", ".ignore")
)
names(north_df)[names(north_df) == "lat"]       <- "lat1"
names(north_df)[names(north_df) == "long"]      <- "long1"
names(north_df)[names(north_df) == "cluster"]   <- "Pop1_cluster"

# Merge to get lat2, long2, and Pop2_cluster
north_df <- merge(
  north_df,
  meta_coords,
  by.x = "Var2",
  by.y = "SiteCode",
  all.x = TRUE,
  suffixes = c("", ".ignore")
)
names(north_df)[names(north_df) == "lat"]       <- "lat2"
names(north_df)[names(north_df) == "long"]      <- "long2"
names(north_df)[names(north_df) == "cluster"]   <- "Pop2_cluster"
# Select and order columns
north_df <- north_df[, c(
  "Var1", "Var2", "CSEdistance",
  "lat1", "long1", "lat2", "long2",
  "Pop1_cluster", "Pop2_cluster"
)]

# Process the "south" CSE matrix (analogous steps)
south_df <- as.data.frame(as.table(cse_south), stringsAsFactors = FALSE)
names(south_df) <- c("Var1", "Var2", "CSEdistance")
south_df <- subset(south_df, Var1 < Var2)
south_df <- merge(
  south_df,
  meta_coords,
  by.x = "Var1",
  by.y = "SiteCode",
  all.x = TRUE,
  suffixes = c("", ".ignore")
)
names(south_df)[names(south_df) == "lat"]       <- "lat1"
names(south_df)[names(south_df) == "long"]      <- "long1"
names(south_df)[names(south_df) == "cluster"]   <- "Pop1_cluster"
south_df <- merge(
  south_df,
  meta_coords,
  by.x = "Var2",
  by.y = "SiteCode",
  all.x = TRUE,
  suffixes = c("", ".ignore")
)
names(south_df)[names(south_df) == "lat"]       <- "lat2"
names(south_df)[names(south_df) == "long"]      <- "long2"
names(south_df)[names(south_df) == "cluster"]   <- "Pop2_cluster"
south_df <- south_df[, c(
  "Var1", "Var2", "CSEdistance",
  "lat1", "long1", "lat2", "long2",
  "Pop1_cluster", "Pop2_cluster"
)]
# Combine "north" and "south" results
cse_combined <- rbind(north_df, south_df)

# Save to CSV
#write.csv(cse_combined, "../input/Gff_11loci_68sites_cse.csv", row.names = FALSE)

# View the first few rows
head(cse_combined)
```
Create density plots of CSEdistance for each cluster  
```{r density}
# Capitalize or rename cluster labels (“Northern”/“Southern”)
cse_combined <- cse_combined %>%
  mutate(
    Pop1_cluster = case_when(
      Pop1_cluster == "north" ~ "Northern",
      Pop1_cluster == "south" ~ "Southern",
      TRUE ~ Pop1_cluster
    ),
    Pop2_cluster = case_when(
      Pop2_cluster == "north" ~ "Northern",
      Pop2_cluster == "south" ~ "Southern",
      TRUE ~ Pop2_cluster
    )
  )
# Create two density plots (one for “Northern”, one for “Southern”)
cluster_names <- unique(
  cse_combined$Pop1_cluster[cse_combined$Pop1_cluster == cse_combined$Pop2_cluster]
)
density_plots <- lapply(cluster_names, function(cl) {
  df <- filter(cse_combined, Pop1_cluster == cl & Pop2_cluster == cl)
  ggplot(df, aes(x = CSEdistance)) +
    geom_density(fill = "grey80", color = "black") +
    labs(
      title = paste0(cl, " Genetic Cluster"),
      x = "CSE Distance",
      y = "Density"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9)
    )
})
# Turn each row into an sf LINESTRING, colored by CSE
lines_list <- lapply(seq_len(nrow(cse_combined)), function(i) {
  coords <- matrix(
    c(
      cse_combined$long1[i], cse_combined$lat1[i],
      cse_combined$long2[i], cse_combined$lat2[i]
    ),
    ncol = 2,
    byrow = TRUE
  )
  st_linestring(coords)
})
lines_sf <- st_sf(
  geometry = st_sfc(lines_list, crs = 4326),
  CSE = cse_combined$CSEdistance
)
# 5. Load Uganda outline as an sf object
uganda <- ne_countries(scale = "medium", country = "Uganda", returnclass = "sf")

# 6. Build the map with reversed “magma” (yellow→purple) and
#    move legend to the right, vertically centered
map_plot <- ggplot() +
  geom_sf(data = uganda, fill = "white", color = "black") +
  geom_sf(
    data = lines_sf,
    aes(color = CSE),
    size = 0.6
  ) +
  scale_color_viridis_c(
    name = "CSE",
    option = "magma",   # use the “magma” palette
    direction = -1      # reverse so low = yellow, high = purple
  ) +
  theme_void() +
  theme(
    # place legend at (x=0.95, y=0.50) relative to entire panel
    legend.position = c(1.3, 0.50),
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
# 7. Combine the two density plots (top) and the map (bottom)
top_row <- plot_grid(
  plotlist = density_plots,
  nrow = 1,
  rel_widths = rep(1, length(density_plots))
)
final_figure <- plot_grid(
  top_row,
  map_plot,
  ncol = 1,
  rel_heights = c(1, 2)
)

# Display CSE density plots
#pdf("../figures/CSE_density.pdf", width = 14, height = 10)
final_figure
#dev.off()  
```