---
title: "Uganda tsetse population genetics"
author: "Norah Saarman"
date: "2025-06-03"
output: pdf_document
---

RStudio Configuration:
   - **R version:** R 4.4.0 (Geospatial packages)  
   - **Number of cores:** 4 (up to 32 available)   
   - **Account:** saarman-np  
   - **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
   - **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(adegenet)
library(hierfstat)
library(ggplot2)
library(scatterplot3d)
library(vcfR)
library(LEA)
library(bgchm)
require(pegas)       
require(PopGenReport)
require(dplyr)
require(poppr) 
```
# Importing data as genind object
Using adegenet and base R

```{R, import}
# 1. CSV file "./data/uganda_genotypes.csv" --> data frame 
# with base R function read.csv()
Flr <- read.csv("./data/uganda_genotypes.csv", header=TRUE)

# 2. Select only adults with base R indexing of data frame 
# rows where OffID==0, all columns
Flr <- Flr[Flr$OffID==0,]

# 3. Combine columns with base R function paste()
Flr <- data.frame(Flr[,1:5],loc1 = paste(Flr[,6],  Flr[,7], sep=":"), 
                            loc2 = paste(Flr[,8],  Flr[,9], sep=":"), 
                            loc3 = paste(Flr[,10], Flr[,11], sep=":"), 
                            loc4 = paste(Flr[,12], Flr[,13], sep=":"), 
                            loc5 = paste(Flr[,14], Flr[,15], sep=":"),  
                            loc6 = paste(Flr[,16], Flr[,17], sep=":"),  
                            loc7 = paste(Flr[,18], Flr[,19], sep=":"))
# 4. Create genind object with "adegenet" function df2genind() 
# using NA.char = "NA"
Flr.genind <- df2genind(X=Flr[,c(6:12)], sep=":", ncode=NULL,
                        ind.names= Flr$ID, loc.names=names(Flr[,c(6:12)]), pop=Flr$Population,
                        NA.char="NA", ploidy=2, type="codom", strata=NULL, hierarchy=NULL)

# 5. Check genind object
Flr.genind
summary(Flr.genind)
```

# Count the number of individuals in each pop
```{r}
table(Flr$Population)
```


## Extract population labels:

```
# - regexpr("^[A-Za-z]+", ...) finds the match of 
# one or more letters at the beginning of each name
# - regmatches() extracts that matching substring.
pop_labels <- regmatches(indNames(genind_obj), regexpr("^[^-_]+", indNames(genind_obj)))

#replace Ps01-Ps08 with Ps
pop_labels <- gsub("^Ps\\d+", "Ps", pop_labels)

# Assign the extracted labels as pop names
pop(genind_obj) <- pop_labels

# Which 17 pops should be included? 
# Only confident about 13 of them...
genind_sub <- genind_obj[genind_obj@pop %in% c("ZXPd02",
  "ZXPd03", "ZXPd04", "XCPd03", "DQPd02","BCPy03", 
  "GSPy04", "JLPy02", "LJPy02", "WSPy01", "WXPy01", 
  "ZDPy06","ZDPy07")]
```

# Visualize genetic variation with PCA
```{r pca}
# Compute PCA on the subset
pca_result3 <- dudi.pca(genind_sub, cent = FALSE, 
    scale = TRUE, scannf = FALSE, nf = 4)

# Choose colors by site/pop
colors <- c(colorRampPalette(c("red", "orange",
  "yellow"))(5),colorRampPalette(c("blue", "cyan",
  "purple"))(10))
site_colors <- c("ZXPd02"=colors[1], 
                 "ZXPd03"=colors[2], 
                 "ZXPd04"=colors[3], 
                 "XCPd03"=colors[4], 
                 "DQPd02"=colors[5], 
                 "BCPy03"=colors[6], 
                 "GSPy04"=colors[8], 
                 "JLPy02"=colors[9], 
                 "LJPy02"=colors[10], 
                 "WSPy01"=colors[11], 
                 "WXPy01"=colors[12], 
                 "ZDPy06"=colors[13], 
                 "ZDPy07"=colors[14])

# Extract population codes 
pop_codes <- as.character(pop(genind_sub))

# Index the named color vector by the pop_group
# by sampled site
pop_colors <- site_colors[pop_codes]

# Define layout: Two columns, the second column 
# is reserved for the legend
layout(matrix(c(1, 2), nrow = 1), 
       widths = c(3.5,0.5))  # 7:1 ratio

# First panel: PCA plot
par(mar = c(5, 4, 4, 2))  # Reset margins
plot(pca_result3$li[,1:2],
     col = pop_colors,
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA - Hybrid Cline")

# Second panel: Legend (ensures it appears on the right)
par(mar = c(5, 0, 4, 0))  # Remove margins
plot.new()  # Create an empty plot space
legend("center", 
       legend = names(site_colors),
       col = site_colors,
       pch = 19,
       cex = 0.8,
       title = "Populations")
```
