---
title: "Uganda tsetse pop gen: PCA, diversity, CSE"
author: "Norah Saarman"
date: "2025-06-03"
output: pdf_document
---

RStudio Configuration:
   - **R version:** R 4.4.0 (Geospatial packages)  
   - **Number of cores:** 4 (up to 32 available)   
   - **Account:** saarman-np  
   - **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
   - **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, results=FALSE, message=FALSE}
library(adegenet)
library(geosphere)
library(igraph)
library(scatterplot3d)
require(ggplot2)
require(dplyr)
library(hierfstat)
require(pegas)       
require(PopGenReport)
require(poppr) 
```

# Importing data as genind object from .gen
```{R, import}
# 1. Genepop .gen file "./input/Gff_11loci_allsites_genepop.gen" 
Gff.genind <- read.genepop("../input/Gff_11loci_allsites_genepop.gen", ncode = 3)

# Metadata with base R function read.csv()
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", header=TRUE, sep = "\t")

# Add site code to @pop slot
Gff.genind@pop <- as.factor(Gff$SiteCode)

# Check genind object
Gff.genind
summary(Gff.genind)
```

# Visualize genetic variation with PCA

All samples:
```{r pca-all}
# replace NAs with locus means for PCA
gen_tab <- tab(Gff.genind, NA.method = "mean")  

# Compute PCA on the subset
pca_result3 <- dudi.pca(gen_tab, cent = FALSE, scale = TRUE, scannf = FALSE, nf = 4)

# Extract percent variance explained for PC1 and PC2
var_explained <- round(100 * pca_result3$eig / sum(pca_result3$eig), 1)

# Generate ~86 visually distinct colors
set.seed(123)  # for reproducibility
n_pops <- nlevels(pop(Gff.genind))
color_palette <- colorRampPalette(c("red", "orange", "yellow", "green", "cyan", "blue", "purple"))(n_pops)

# Map population levels to colors
site_levels <- levels(pop(Gff.genind))
site_colors <- setNames(color_palette, site_levels)

# Get color for each individual
pop_colors <- site_colors[as.character(pop(Gff.genind))]

# Layout: 2 panels (main plot + legend)
layout(matrix(c(1, 2), nrow = 1), widths = c(6, 2.5))

# PCA plot
#pdf("../figures/PCA-all-2D.pdf", width = 14, height = 10)
par(mar = c(5, 4, 4, 2))
plot(pca_result3$li[, 1:2],
     col = pop_colors,
     pch = 19,
     xlab = paste0("PC1 (", var_explained[1], "%)"),
     ylab = paste0("PC2 (", var_explained[2], "%)"))

# Legend panel
par(mar = c(1,1,1,1))  # increase space to avoid clipping
plot.new()
legend("center", 
       legend = site_levels,
       col = site_colors,
       pch = 19,
       cex = 0.6,
       title = "Populations", 
       ncol = 3,
       bty = "n")  # <- no border
#dev.off()

# 3D PCA plot
#pdf("../figures/PCA-all-3D.pdf", width = 14, height = 10)
par(mar = c(5, 4, 4, 2))
s3d <- scatterplot3d(pca_result3$li[, c(1,3,2)],
                     color = pop_colors,
                     pch = 19,
                     xlab = paste0("PC1 (", var_explained[1], "%)"), # horizontal 
                     ylab = paste0("PC3 (", var_explained[3], "%)"), # depth
                     zlab = paste0("PC2 (", var_explained[2], "%)"), # vertical
                     angle = 55,
                     box = TRUE)

# Legend panel
par(mar = c(1, 1, 1, 1))
plot.new()
legend("center",
       legend = site_levels,
       col = site_colors,
       pch = 19,
       cex = 0.6,
       title = "Populations",
       ncol = 3,
       bty = "n")


#dev.off()

```

# Considering K=2 by Evano method
```{r}
# Add K2 cluster combining "south" and "west" into "south"
Gff$K2 <- ifelse(Gff$SiteMajCluster %in% c("south", "west"), "south", Gff$SiteMajCluster)
```
PCA for each K2 clusters separately
```{r}
# PCA for each K2 cluster separately
for (region in c("north","south")) {
  genind_sub <- Gff.genind[pop(Gff.genind) %in% Gff$SiteCode[Gff$K2 == region]]

  # Replace NAs with locus means for PCA
  gen_tab <- tab(genind_sub, NA.method = "mean")
  pca_result <- dudi.pca(gen_tab, cent = FALSE, scale = TRUE, scannf = FALSE, nf = 4)
  var_explained <- round(100 * pca_result$eig / sum(pca_result$eig), 1)

  # Color palette
  set.seed(123)
  n_pops <- nlevels(pop(genind_sub))
  palette <- colorRampPalette(c("red", "orange", "yellow", "green", "cyan", "blue", "purple"))(n_pops)
  site_levels <- levels(pop(genind_sub))
  site_colors <- setNames(palette, site_levels)
  pop_colors <- site_colors[as.character(pop(genind_sub))]

  layout(matrix(c(1, 2), nrow = 1), widths = c(6, 2.5))

  # 2D PCA plot
  #pdf(paste0("../figures/PCA-", region, "-2D.pdf"), width = 14, height = 10)
  par(mar = c(5, 4, 4, 2))
  plot(pca_result$li[, 1:2],
       col = pop_colors,
       pch = 19,
       xlab = paste0("PC1 (", var_explained[1], "%)"),
       ylab = paste0("PC2 (", var_explained[2], "%)"))

  # Legend
  par(mar = c(1, 1, 1, 1))
  plot.new()
  legend("center",
         legend = site_levels,
         col = site_colors,
         pch = 19,
         cex = 0.6,
         title = "Populations",
         ncol = 2,
         bty = "n")
  #dev.off()

  # 3D PCA plot
  #pdf(paste0("../figures/PCA-", region, "-3D.pdf"), width = 14, height = 10)
  par(mar = c(5, 1, 4, 2))
  scatterplot3d(x = pca_result$li[, 1], # horizontal
                y = pca_result$li[, 3], # depth
                z = pca_result$li[, 2], # vertical
                color = pop_colors,
                pch = 19,
                xlab = paste0("PC1 (", var_explained[1], "%)"),
                ylab = paste0("PC3 (", var_explained[3], "%)"),
                zlab = paste0("PC2 (", var_explained[2], "%)"),
                angle = 55,
                box = TRUE)
  par(mar = c(1, 1, 1, 1))
  plot.new()
  legend("center",
         legend = site_levels,
         col = site_colors,
         pch = 19,
         cex = 0.6,
         title ="Populations",
         ncol = 2,
         bty = "n")
  #dev.off()
}
```
# Choosing sites for Random Forest (K=2)

Filter out any Gff$Cluster == nogroup to avoid admixed individuals, which will inflate the genetic distance within each geographic region. 

Filtering for only largest/most complete genetic sample from each 5 km radius for gps within each Gff$SiteMajCluster independently so that we get best samples from both north and south of the river/Lake Kyoga dividing north and south

```{r distm}
# Load required packages
library(geosphere)
library(igraph)

# Create filtered copy of Gff excluding 'nogroup'
Gff_filt <- Gff[Gff$Cluster != "nogroup", ]

# Ensure Gff$SiteCode matches pop(Gff.genind)
Gff_filt$SiteCode <- as.character(Gff_filt$SiteCode)
pop_names <- as.character(pop(Gff.genind))

if (!all(pop_names == Gff_filt$SiteCode)) {
  Gff_filt <- Gff_filt[match(pop_names, Gff_filt$SiteCode), ]
  stopifnot(all(as.character(pop(Gff.genind)) == Gff_filt$SiteCode))
}

# Initialize storage
selected_sites <- character()
report_list <- list()

# Loop through each K2 cluster separately (north/south)
for (region in c("north","south")) {
  Gff_sub <- Gff_filt[Gff_filt$K2 == region, ]

  # Build coordinate matrix
  xy <- cbind(lon = Gff_sub$SiteLong, lat = Gff_sub$SiteLat)
  dist_mat <- distm(xy, fun = distHaversine)

  # Define adjacency for <2 km
  threshold_m <- 2000
  adj_mat <- dist_mat < threshold_m

  # Create undirected graph of close sites
  g <- graph_from_adjacency_matrix(adj_mat, mode = "undirected", diag = FALSE)
  clust_ids <- components(g)$membership

  # For each 5km cluster, pick the SiteCode with the most individuals in Gff.genind
  for (cid in unique(clust_ids)) {
    site_group <- Gff_sub$SiteCode[clust_ids == cid]
    counts <- table(pop(Gff.genind)[pop(Gff.genind) %in% site_group])
    best_site <- names(which.max(counts))
    selected_sites <- c(selected_sites, best_site)
    report_list[[best_site]] <- site_group
  }
}

# Subset Gff and Gff.genind to selected sites
Gff_rf <- Gff_filt[Gff_filt$SiteCode %in% selected_sites, ]
Gff.genind_rf <- Gff.genind[pop(Gff.genind) %in% selected_sites]

# Report selected SiteCodes by region
print(split(Gff_rf$SiteCode, Gff_rf$K2))

# Report groupings and selected representative from each 5km cluster
report_df <- data.frame(
  selected = names(report_list),
  group = sapply(report_list, function(x) paste(sort(unique(x)), collapse = ", ")),
  row.names = NULL
)
print(report_df)
```
Remove smaller sample size(s) of any 2-3 sites within 2 km of each other: 04-OM, 29-KO, 13-GOR, 34-AMI, 35-DK, 39-OC, 53-UWA, 77-NB, 80-SA

Brings total site count down to 75.
```{r}
Gff.genind_rf # group size 8-236
table(pop(Gff.genind_rf))
```

Too imbalanced... 

  - Downsample 55-KAF, 78-OK, 79-BU to N individuals (e.g., 50)
  - Remove sites with fewer than n=15

```{r}
# Set thresholds
min_n <- 15
max_n <- 50

# Compute sample sizes per population
pop_sizes <- table(pop(Gff.genind_rf))

# Keep populations with n ≥ min_n
pops_to_keep <- names(pop_sizes[pop_sizes >= min_n])

# Indices to retain
inds_keep <- unlist(lapply(pops_to_keep, function(p) {
  inds <- which(pop(Gff.genind_rf) == p)
  if (length(inds) > max_n) {
    sample(inds, max_n)  # downsample
  } else {
    inds  # keep as-is
  }
}))

# Subset both genind and metadata
Gff.genind_rf <- Gff.genind_rf[inds_keep]
Gff_rf <- Gff_rf[inds_keep, ]

# Re-check population sizes
Gff.genind_rf
summary(Gff.genind_rf)
table(pop(Gff.genind_rf))
length(unique(Gff_rf$SiteCode))
```
Now 68 pops with group size ranges from 15-50.

# Genetic diversity

Convert to hierfstat format
```{r}
library(hierfstat)

genind_hf <- genind2hierfstat(Gff.genind_rf)
```

Estimate allelic richness (standardized/rarefied)
```{r}
library(pegas)

# This function rarefies allelic richness to the smallest sample size
allelic_richness <- allelic.richness(genind_hf)

# Extract mean rarefied allelic richness per population
mean_ar <- colMeans(allelic_richness$Ar, na.rm = TRUE)

mean_ar
``` 

Most diversity stats use poppr
```{r}
# Load libraries
require(poppr)
require(adegenet)

# Compute per-population stats
div_stats <- poppr(Gff.genind_rf)

# View a summary table
div_stats
```
For F-statistics, you can use hierfstat::basic.stats() but need to split by population first... and basic.stats(genind_hf) will not do this for you automatically. Maybe you would need to run over a loop of pops, and subsample each population first, than basic.stats(@pop)

# Cavalli-Sforza and Edwards' Distance (CSE)

Pairwise within clusters "north" and "south"
```{r}
# Loop through K2 clusters
for (region in c("north", "south")) {

  # Subset genind and metadata
  meta_sub <- Gff_rf[Gff_rf$K2 == region, ]
  gen_sub <- Gff.genind_rf[pop(Gff.genind_rf) %in% meta_sub$SiteCode]

  # Convert to genpop (aggregates by population)
  genpop_sub <- genind2genpop(gen_sub)

  # Compute Cavalli-Sforza and Edwards’ distance
  cse_dist <- dist.genpop(genpop_sub, method = 2)  # method = 2 → CSE distance

  # Display as a matrix
  cse_matrix <- as.matrix(cse_dist)

  cat("\n--- Cavalli-Sforza and Edwards' distance for", region, "---\n")
  print(round(cse_matrix, 4))
}```
