---
title: "RF model full – LC lakes paths"
author: "Norah Saarman"
date: "2025-06-17"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 4 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
- **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)    

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(randomForest)
library(doParallel)
library(raster)
library(sf)
library(viridis)
library(dplyr)

# base directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
input_dir <- "../input"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results"

# read the combined CSE + coords table + pix_dist + Env variables
V.table <- read.csv(file.path(input_dir, "Gff_cse_envCostPaths.csv"),
                    header = TRUE)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
V.table <- V.table %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

# simple mode helper
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[ which.max(tabulate(match(x, ux))) ]
}

# setup running in parallel
cl <- makeCluster(4)
registerDoParallel(cl)
clusterExport(cl, "get_mode")
```

# Inputs  
  - `../input/Gff_cse_envCostPaths.csv`  - Combined CSE table with coordinates (long1, lat1, long2, lat2), pix_dist = geographic distance in sum of pixels, and mean, median, mode of each Env parameter  
# Outputs  
  - `../results_dir/rf_mean18_tuned.rds`  - Full RF model output, readRDS() to load  
  - `../results_dir/predicted_CSEdistance.tif` - Projection of full RF model, with pix_dist and samp_20km nuetralized by using uniform layers in projection
  
# 1. Prepare the data
```{r, prep}
# Assign input, checking for any rows with NA
sum(!complete.cases(V.table))  # should return 0
rf_data <- na.omit(V.table)    # should omit zero rows

# Confirm that CSEdistance is numeric
rf_data$CSEdistance <- as.numeric(rf_data$CSEdistance)

# Select variables: all predictors (mean, median, mode)  
predictor_vars <- c("pix_dist",                      # geo dist
  paste0("BIO", 1:7, "_mean"),                       # mean 
  paste0("BIO", 8:11, "S_mean"),                     # mean
  paste0("BIO", 12:15, "_mean"),                     # mean
  paste0("BIO", 16:19, "S_mean"),                    # mean
  "alt_mean", "slope_mean", "riv_3km_mean",          # mean
  "samp_20km_mean", "lakes_mean",                    # mean
  paste0("BIO", 1:7, "_median"),                     # median
  paste0("BIO", 8:11, "S_median"),                   # median
  paste0("BIO", 12:15, "_median"),                   # median
  paste0("BIO", 16:19, "S_median"),                  # median
  "alt_median", "slope_median", "riv_3km_median",    # median
  "samp_20km_median", "lakes_median",                # median
  paste0("BIO", 1:7, "_mode"),                       # mode
  paste0("BIO", 8:11, "S_mode"),                     # mode
  paste0("BIO", 12:15, "_mode"),                     # mode
  paste0("BIO", 16:19, "S_mode"),                    # mode
  "alt_mode", "slope_mode", "riv_3km_mode",          # mode
  "samp_20km_mode", "lakes_mode"                     # mode
)

# subset predictors that we want to use
rf_data <- rf_data[, c("CSEdistance", predictor_vars)]
```

# 2. Build full Random Forest model 
```{r rf}
# Build full RF model
set.seed(1234)  # ensures reproducibility
rf_full <- randomForest(
  CSEdistance ~ .,
  data = rf_data,
  importance = TRUE,
  ntree = 500
)

print(rf_full)

importance(rf_full)
```

# 2. Prune variables?

## Compare mean versus median versus mode:
```{r mean-v-median-v-mode}
# Extract groups of variables by suffix
mean_vars   <- grep("_mean$", names(rf_data), value = TRUE)
median_vars <- grep("_median$", names(rf_data), value = TRUE)
mode_vars   <- grep("_mode$", names(rf_data), value = TRUE)

# Always include geographic distance
common_var <- "pix_dist"

# Build and run each model
set.seed(123438972)  # ensures reproducibility
rf_mean <- randomForest(CSEdistance ~ ., data = rf_data[, c("CSEdistance", common_var, mean_vars)], ntree = 500, importance = TRUE)
rf_median <- randomForest(CSEdistance ~ ., data = rf_data[, c("CSEdistance", common_var, median_vars)], ntree = 500, importance = TRUE)
rf_mode <- randomForest(CSEdistance ~ ., data = rf_data[, c("CSEdistance", common_var, mode_vars)], ntree = 500, importance = TRUE)

# Compare performance
c(mean = rf_mean$rsq[500] * 100,
  median = rf_median$rsq[500] * 100,
  mode = rf_mode$rsq[500] * 100)
```

Including mean of env variable along least cost paths performs the best, adding median and mode does not greatly improve the model and increases risks of over fitting...

## Prune more variables after narrowing to mean only?

```{r mean}
# Plot variable importance
par(mar = c(5, 10, 2, 2))  # bottom, left, top, right
varImpPlot(rf_mean, main = "Mean Model Importance",cex = 0.6, pch = 19)

# Rank variables by %IncMSE (from tuned model)
imp <- importance(rf_mean)[, "%IncMSE"]
imp <- sort(imp, decreasing = TRUE)

# Multiple runs with N top predictors
# Store results
prune_results <- list()
n_list <- c(5:length(imp))

for (n in n_list) {
  top_vars <- names(imp)[1:n]
  formula_n <- as.formula(paste("CSEdistance ~", paste(top_vars, collapse = " + ")))
  
  set.seed(1234783645)
  rf_n <- randomForest(
    formula = formula_n,
    data = rf_data,
    ntree = 500,
    importance = TRUE
  )
  
  prune_results[[paste0("Top", n)]] <- rf_n
}

sapply(prune_results, function(mod) {
  c(OOB_MSE = mod$mse[500], VarExpl = mod$rsq[500] * 100)
})
```
% Variance Explained increases rapidly up to around 18 variables, after which it plateaus.

OOB MSE decreases quickly early on, with minimal gains beyond the top ~18 predictors.

## Top 18 predictors

```{r top18}
# Get variable importance
var_imp <- importance(rf_mean)[, "%IncMSE"]

# Sort and get names of top 18 predictors
top18_vars <- names(sort(var_imp, decreasing = TRUE))[1:18]

rf_top18_data <- rf_data[, c("CSEdistance", top18_vars)]
```

# 3. Tune random forest with chosen variables
```{r tune}

# Subset data for the mean-only model
#rf_data <- rf_data[, c("CSEdistance", common_var, mean_vars)]
rf_data <- rf_data[, c("CSEdistance", top18_vars)]

# Rename predictors by removing "_mean" for later projections
names(rf_mean_data) <- gsub("_mean$", "", names(rf_mean_data))

# Build full RF model
set.seed(10981234)  # ensures reproducibility
rf_mean18 <- randomForest(
  CSEdistance ~ .,
  data = rf_mean_data,
  importance = TRUE,
  ntree = 500
)

print(rf_mean18)
importance(rf_mean18)

# Tune mtry (number of variables tried at each split)
set.seed(92834567)
rf_mean18_tuned <- tuneRF(
  x = rf_mean_data[, -1],   # exclude response variable
  y = rf_mean_data$CSEdistance,
  ntreeTry = 500,
  stepFactor = 1.5,         # factor by which mtry is increased/decreased
  improve = 0.01,           # minimum improvement to continue search
  trace = TRUE,             # print progress
  plot = TRUE,              # plot OOB error vs mtry
  doBest = TRUE,             # return the model with lowest OOB error
  importance = TRUE
)


# Save the tuned random forest model to disk
saveRDS(rf_mean18_tuned, file = file.path(results_dir, "rf_mean18_tuned.rds"))

# Preserve as-is for projection
rf_final <- rf_mean18_tuned
```
FYI: Later, to load the model back into R:
`rf_mean18_tuned <- readRDS(file.path(results_dir, "rf_mean18_tuned.rds"))`


## Compare full and full tuned models (top 18 mean-only predictors)
```{r compare}
print(rf_mean18)
print(rf_mean18_tuned)

data.frame(
  Model = c("Full (default mtry)", paste("Tuned (mtry = ",rf_mean18_tuned$mtry,")")),
  MSE = c(rf_mean18$mse[rf_mean18$ntree], rf_mean18_tuned$mse[rf_mean18_tuned$ntree]),
  Rsq = c(rf_mean18$rsq[rf_mean18$ntree], rf_mean18_tuned$rsq[rf_mean18_tuned$ntree])
)

# pad names to trick varImpPlot
rownames(rf_mean18$importance) <- paste0("  ", rownames(rf_mean18$importance), "  ")
rownames(rf_mean18_tuned$importance) <- paste0("  ", rownames(rf_mean18_tuned$importance), "  ")

# plot with varImpPlot
par(mar = c(5, 30, 2,2))  # bottom, left, top, right
varImpPlot(rf_mean18, main = "Full Model Importance",cex = 0.6, pch = 19)
varImpPlot(rf_mean18_tuned, main = "Tuned Full Model Importance",cex = 0.6, pch = 19)
```

The tuned model performs slightly better, but the gain may not be meaningful... however, it does confirm that the model is stable and that the mean-only predictors carry strong signal.

Top 18 mean-based predictors retain nearly all the explanatory power of the original full model with 42 predictors.

# 4. Projection of model
```{r projection}
# Load env stack with named layers
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Replace samp_20km with neutral constant raster, 
# retains sampling bias in the model but neutralize 
# sampling bias during projection
samp_uniform <- env[["samp_20km"]]
values(samp_uniform) <- mean(rf_data$samp_20km_mean, na.rm = TRUE)
env[["samp_20km"]] <- samp_uniform # Replace in raster stack

prediction_raster <- predict(env, rf_final, type = "response")

# Write Prediction Raster to file
#writeRaster(prediction_raster, file.path(results_dir,"fullRF_CSE_resistance.tif"), format = "GTiff", overwrite = TRUE)
```

```{r plot-projection, warning=FALSE}
# Create base plot with viridis
plot(prediction_raster,
     col = viridis::magma(100),
     main = "Predicted CSE Distance",
     axes = FALSE,
     box = FALSE,
     legend.args = list(text = "CSE Distance", side = 2, line = 2.5, cex = 0.8))

# Overlay lakes in dark gray
lakes <- st_read(file.path(data_dir, "raw/ne_10m_lakes.shp"), quiet = TRUE)
lakes <- st_transform(lakes, crs = st_crs(prediction_raster))  # match CRS 
lakes <- st_make_valid(lakes) # fix geometries
r_ext <- st_as_sfc(st_bbox(prediction_raster)) # extent
st_crs(r_ext) <- st_crs(prediction_raster) # match CRS
lakes <- st_intersection(lakes, r_ext) # clip to extent
plot(st_geometry(lakes), col = "gray20", border = NA, add = TRUE)

# Overlay country outline
uganda <- rnaturalearth::ne_countries(continent = "Africa", scale = "medium", returnclass = "sf")
uganda <- st_intersection(uganda, r_ext) # clip to extent
plot(st_geometry(uganda), col = NA, border = "black", lwd = 1.2, add = TRUE)
```


# Repeat with scaled CSE

## Top 18 predictors
```{r top18}
# Load data
V.table_full <- read.csv(file.path(input_dir, "Gff_cse_envCostPaths.csv"))

# estimate mean sampling density
uniform_samp <- mean(V.table_full$samp_20km_mean, na.rm = TRUE)
message("The mean sampling density is: ", uniform_samp)
  
# Filter out western outlier "50-KB" 
V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# Create unique ID after filtering
V.table$id <- paste(V.table$Var1, V.table$Var2, sep = "_")

# Define top 18 mean predictors (adjust names if necessary)
predictor_vars <- c("pix_dist", "samp_20km_mean", "BIO3_mean", "BIO6_mean", "BIO15_mean","BIO13_mean", "BIO11S_mean", "BIO7_mean", "BIO1_mean", "BIO9S_mean","BIO12_mean", "alt_mean", "BIO18S_mean", "BIO4_mean", "BIO2_mean","slope_mean", "BIO10S_mean", "riv_3km_mean")

max_CSE <- max(V.table$CSEdistance, na.rm = TRUE)
V.table$CSE_scaled <- V.table$CSEdistance / max_CSE

# Filter to modeling-relevant columns only
rf_data <- V.table[, c("CSE_scaled", predictor_vars)]
```

# Tune random forest with chosen variables
```{r tune}

# Rename predictors by removing "_mean" for later projections
names(rf_data) <- gsub("_mean$", "", names(rf_data))

# Build full RF model
set.seed(10981234)  # ensures reproducibility
rf_scaled <- randomForest(
  CSE_scaled ~ .,
  data = rf_data,
  importance = TRUE,
  ntree = 500
)

print(rf_scaled)
importance(rf_scaled)

# Tune mtry (number of variables tried at each split)
set.seed(92834567)
rf_scaled_tuned <- tuneRF(
  x = rf_data[, -1],   # exclude response variable
  y = rf_data$CSE_scaled,
  ntreeTry = 500,
  stepFactor = 1.5,         # factor by which mtry is increased/decreased
  improve = 0.01,           # minimum improvement to continue search
  trace = TRUE,             # print progress
  plot = TRUE,              # plot OOB error vs mtry
  doBest = TRUE,             # return the model with lowest OOB error
  importance = TRUE
)


# Save the tuned random forest model to disk
saveRDS(rf_scaled_tuned, file = file.path(results_dir, "rf_scaled_tuned.rds"))

# Preserve as-is for projection
rf_final <- rf_scaled_tuned
```
FYI: Later, to load the model back into R:
`rf_scaled_tuned <- readRDS(file.path(results_dir, "rf_scaled_tuned.rds"))`


## Compare full and full tuned models (top 18 mean-only predictors)
```{r compare}
print(rf_scaled)
print(rf_scaled_tuned)

data.frame(
  Model = c("Full (default mtry)", paste("Tuned (mtry = ",rf_scaled_tuned$mtry,")")),
  MSE = c(rf_scaled$mse[rf_scaled$ntree], rf_scaled_tuned$mse[rf_scaled_tuned$ntree]),
  Rsq = c(rf_scaled$rsq[rf_scaled$ntree], rf_scaled_tuned$rsq[rf_scaled_tuned$ntree])
)

# pad names to trick varImpPlot
rownames(rf_scaled$importance) <- paste0("  ", rownames(rf_scaled$importance), "  ")
rownames(rf_scaled_tuned$importance) <- paste0("  ", rownames(rf_scaled_tuned$importance), "  ")

# plot with varImpPlot
par(mar = c(5, 30, 2,2))  # bottom, left, top, right
varImpPlot(rf_scaled, main = "Full Model Importance",cex = 0.6, pch = 19)
varImpPlot(rf_scaled_tuned, main = "Tuned Full Model Importance",cex = 0.6, pch = 19)
```

# 4. Projection of model
```{r projection-scaled}
# Load env stack with named layers
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Replace samp_20km with neutral constant raster, 
# retains sampling bias in the model but neutralize 
# sampling bias during projection
samp_uniform <- env[["samp_20km"]]
values(samp_uniform) <- mean(rf_data$samp_20km, na.rm = TRUE)
env[["samp_20km"]] <- samp_uniform # Replace in raster stack

prediction_raster <- predict(env, rf_final, type = "response")

# Write Prediction Raster to file
#writeRaster(prediction_raster, file.path(results_dir,"fullRF_CSE_scaled_resistance.tif"), format = "GTiff", overwrite = TRUE)
```

```{r plot-projection-scaled, warning=FALSE}
# Create base plot with viridis
plot(prediction_raster,
     col = viridis::magma(100),
     main = "Predicted Scaled CSE",
     axes = FALSE,
     box = FALSE,
     legend.args = list(text = "Scaled CSE", side = 2, line = 2.5, cex = 0.8))

# Overlay lakes in dark gray
lakes <- st_read(file.path(data_dir, "raw/ne_10m_lakes.shp"), quiet = TRUE)
lakes <- st_transform(lakes, crs = st_crs(prediction_raster))  # match CRS 
lakes <- st_make_valid(lakes) # fix geometries
r_ext <- st_as_sfc(st_bbox(prediction_raster)) # extent
st_crs(r_ext) <- st_crs(prediction_raster) # match CRS
lakes <- st_intersection(lakes, r_ext) # clip to extent
plot(st_geometry(lakes), col = "gray20", border = NA, add = TRUE)

# Overlay country outline
uganda <- rnaturalearth::ne_countries(continent = "Africa", scale = "medium", returnclass = "sf")
uganda <- st_intersection(uganda, r_ext) # clip to extent
plot(st_geometry(uganda), col = NA, border = "black", lwd = 1.2, add = TRUE)
```
 
# Unscaled projection (reversed scaling)

```{r plot-projection-unscaled, warning=FALSE}
prediction_raster <- (predict(env, rf_final, type = "response")) * max_CSE

# Write Prediction Raster to file
#writeRaster(prediction_raster, file.path(results_dir,"fullRF_CSE_unscaled_resistance.tif"), format = "GTiff", overwrite = TRUE)

# Create base plot with viridis
plot(prediction_raster,
     col = viridis::magma(100),
     main = "Predicted CSE",
     axes = FALSE,
     box = FALSE,
     legend.args = list(text = "CSE", side = 2, line = 2.5, cex = 0.8))

# Overlay lakes in dark gray
lakes <- st_read(file.path(data_dir, "raw/ne_10m_lakes.shp"), quiet = TRUE)
lakes <- st_transform(lakes, crs = st_crs(prediction_raster))  # match CRS 
lakes <- st_make_valid(lakes) # fix geometries
r_ext <- st_as_sfc(st_bbox(prediction_raster)) # extent
st_crs(r_ext) <- st_crs(prediction_raster) # match CRS
lakes <- st_intersection(lakes, r_ext) # clip to extent
plot(st_geometry(lakes), col = "gray20", border = NA, add = TRUE)

# Overlay country outline
uganda <- rnaturalearth::ne_countries(continent = "Africa", scale = "medium", returnclass = "sf")
uganda <- st_intersection(uganda, r_ext) # clip to extent
plot(st_geometry(uganda), col = NA, border = "black", lwd = 1.2, add = TRUE)
```


