---
title: "LOPOCV: Random Forest Cross-Validation"
author: "Norah Saarman"
date: "2025-06-18"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 8 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
- **Memory per job:** 200G (cluster limit: 1000G total; avoid exceeding half)    

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(dplyr)
library(readr)
library(raster)
library(randomForest)
library(sp)
library(sf)
library(spatstat)
library(foreach)
library(doParallel)

# Set number of cores and register cluster
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Define Paths to directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
output_dir <- paste0(results_dir,"lopocv")
dir.create(output_dir, showWarnings = FALSE)

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84
```

# Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../data_dir/processed/env_stack.grd` - Final raster stack for prediction including pix_dist
  
# Outputs  
  - `../results_dir/lopocv/` - 
  
# 1. Create Fold Files
```{r, fold-files}
# Load pairwise genetic distance table and add unique ID for each pair
G.table <- read.csv("../input/Gff_11loci_68sites_cse.csv", header = TRUE)
G.table$id <- paste(G.table$Var1, G.table$Var2, sep = "_")

# Get sorted list of unique sampling sites across all pairwise comparisons
sites <- sort(unique(c(G.table$Var1, G.table$Var2)))

# Load environmental raster stack and rename pixel density layer
envPlus <- stack(file.path(data_dir, "processed", "env_stack.grd"))
names(envPlus[["pix_dist"]]) <- "pixvals"

# Loop over each site to generate fold-specific train/test datasets
for (i in seq_along(sites)) {
  # Test set: all pairs where the site is involved
  test_rows  <- G.table %>% filter(Var1 == sites[i] | Var2 == sites[i])
  
  # Training set: all remaining pairs
  train_rows <- G.table %>% filter(Var1 != sites[i] & Var2 != sites[i])

  # Save to output directory
  write_csv(test_rows,  file.path(output_dir, sprintf("testData_%02d.csv", i)))
  write_csv(train_rows, file.path(output_dir, sprintf("trainData_%02d.csv", i)))
}
```

# 2. Run LOPOCV in Parallel 
```{r rf}
# Load least-cost path shapefile with attributes
lines_sf <- st_read(file.path(data_dir, "processed", "LC_paths.shp"), quiet = TRUE)

# Initialize output data frame to hold validation metrics
val_param <- data.frame(matrix(nrow = length(sites), ncol = 8))
colnames(val_param) <- c("RSQ", "RMSE", "RMSE2", "MAE", "MAE2", "MAE3", "COR1", "COR2")

# Run LOPOCV in parallel
foreach(i = seq_along(sites), .combine = rbind, .packages = c("dplyr", "readr", "raster", "randomForest", "sp", "sf")) %dopar% {

  # Identify site for this fold
  site <- sites[i]

  # Load train/test tables for this site
  test_table  <- read_csv(file.path(output_dir, sprintf("testData_%02d.csv", i)), show_col_types = FALSE)
  train_table <- read_csv(file.path(output_dir, sprintf("trainData_%02d.csv", i)), show_col_types = FALSE)

  # Filter spatial lines using precomputed ID column
  train_lines <- lines_sf %>% filter(id %in% train_table$id)
  test_lines  <- lines_sf %>% filter(id %in% test_table$id)

  # Convert to Spatial for raster extraction
  train_sp <- as(train_lines, "Spatial")
  test_sp  <- as(test_lines,  "Spatial")

  # Extract environmental predictors
  envvals_train <- raster::extract(envPlus, train_sp, fun = median, na.rm = TRUE)
  pixvals_train <- raster::extract(envPlus[["pixvals"]], train_sp, fun = sum, na.rm = TRUE)

  envvals_test  <- raster::extract(envPlus, test_sp, fun = median, na.rm = TRUE)
  pixvals_test  <- raster::extract(envPlus[["pixvals"]], test_sp, fun = sum, na.rm = TRUE)

  # Format data frames for training/testing
  train_df <- as.data.frame(envvals_train)
  train_df$pixvals <- pixvals_train
  train_df$Distance <- train_table$CSEdistance

  test_df <- as.data.frame(envvals_test)
  test_df$pixvals <- pixvals_test
  test_df$Distance <- test_table$CSEdistance

  # Tune mtry for Random Forest
  tune_x <- train_df[, setdiff(names(train_df), "Distance")]
  tune_y <- train_df$Distance
  bestmtry <- tuneRF(tune_x, tune_y, stepFactor = 1.5, improve = 1e-5, ntree = 500, trace = FALSE)
  mtry_opt <- bestmtry[,"mtry"][which.min(bestmtry[,"OOBError"])]

  # Train Random Forest model
  rf_model <- randomForest(Distance ~ ., data = train_df, importance = TRUE, mtry = mtry_opt, na.action = na.omit)

  # Predict full raster surface
  pred_raster <- predict(envPlus, rf_model)
  writeRaster(pred_raster,
              filename = file.path(output_dir, sprintf("LOPOCV_%02d.tif", i)),
              format = "GTiff", overwrite = TRUE)

  # Generate validation metrics
  pred_train <- predict(rf_model, newdata = train_df)
  pred_test  <- predict(rf_model, newdata = test_df)

  rsq   <- tail(rf_model$rsq, 1)
  rmse  <- sqrt(tail(rf_model$mse, 1))
  rmse2 <- sqrt(mean((pred_test - test_df$Distance)^2))
  mae   <- mean(abs(pred_train - train_df$Distance))
  mae2  <- mean(abs(predict(rf_model, train_df) - train_df$Distance))
  mae3  <- mean(abs(pred_test - test_df$Distance))
  cor1  <- cor(pred_train, train_df$Distance)
  cor2  <- cor(pred_test, test_df$Distance)

  # Return metrics for this fold
  c(rsq, rmse, rmse2, mae, mae2, mae3, cor1, cor2)
} -> val_param

# 5. Save validation metrics
write.csv(val_param, file.path(output_dir, "validation_metrics.csv"), row.names = FALSE)

# 6. Stop cluster
stopCluster(cl)
```
