---
title: "Iterative RF model and least-cost path updating"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../scripts_knitted_md", 
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/knitted_mds/"
)
```

RStudio Configuration:  
- R version: R 4.4.0 (Geospatial packages)  
- Number of cores: 4 (up to 32 available)   
- Account: saarman-np  
- Partition: saarman-np or saarman-shared-np  
- Memory per job: 100G  

# Setup

```{r libraries}
library(randomForest)
library(raster)
library(terra)
library(sf)
library(sp)
library(gdistance)
library(dplyr)
library(foreach)
library(doParallel)
```

## Directories and inputs

```{r input-dir}
# base directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
input_dir <- "../input"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results"

# main pairwise table with coordinates and CSE
G.table <- read.csv(
  file.path(input_dir, "Gff_11loci_68sites_cse.csv"),
  header = TRUE
)

# optional: remove western outlier now so it stays consistent through all iterations
G.table <- G.table %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# create pair id
G.table$id <- paste(G.table$Var1, G.table$Var2, sep = "_")

# define coordinate reference system
crs_geo <- "+proj=longlat +datum=WGS84 +no_defs"

# number of iterations
n_iter <- 10

# number of cores
n_cores <- 4
```

# 1. Load environmental rasters

```{r rasters}
# seasonal BioClim stack
envvars1 <- stack(
  file.path(data_dir, "processed", "UgandaBiovarsSeasonalAllYears.tif")
)
crs(envvars1) <- crs_geo
names(envvars1) <- c(paste0("BIO", 8:11, "S"), paste0("BIO", 16:19, "S"))

# additional BioClim layers
envvars2 <- stack(
  file.path(data_dir, "processed", "UgandaBiovarsQuartersAllYears.tif")
)
crs(envvars2) <- crs_geo
names(envvars2) <- paste0("BIO", c(1:19))

# combine in desired order
envvars <- raster::stack(
  envvars2[[1:7]],
  envvars1[[1:4]],
  envvars2[[12:15]],
  envvars1[[5:8]]
)

# load elevation and slope rasters
altitude <- raster(
  file.path(data_dir, "processed", "altitude_1KMmedian_MERIT_UgandaClip.tif")
)
crs(altitude) <- crs_geo
names(altitude) <- "alt"

slope <- raster(
  file.path(data_dir, "processed", "slope_1KMmedian_MERIT_UgandaClip.tif")
)
crs(slope) <- crs_geo
names(slope) <- "slope"

# load river and sampling kernels
rivers <- raster(
  file.path(data_dir, "processed", "river_kernel_density_3km.tif")
)
crs(rivers) <- crs_geo
names(rivers) <- "riv_3km"

kernel <- raster(
  file.path(data_dir, "processed", "sample_kernel_density_20km.tif")
)
crs(kernel) <- crs_geo
names(kernel) <- "samp_20km"

# load lake binary layer
lakeRaster <- raster(
  file.path(data_dir, "processed", "lake_binary.tif")
)
crs(lakeRaster) <- crs_geo
names(lakeRaster) <- "lakes"

# load uniform geographic distance raster
geo_dist <- raster(
  file.path(data_dir, "processed", "geo_dist_uniform.tif")
)
crs(geo_dist) <- crs_geo
names(geo_dist) <- "pix_dist"

# stack for extraction
env <- stack(envvars, altitude, slope, rivers, kernel, lakeRaster)

# stack for RF projection
envstack <- stack(envvars, altitude, slope, rivers, kernel, lakeRaster, geo_dist)

# optional: neutralize sampling density before projection if desired
# envstack$samp_20km <- 1.027064e-11
```

# 2. Starting paths

Iteration 1 starts from the existing least-cost paths that already avoid lakes.

```{r starting-paths}
current_lines_sf <- st_read(
  file.path(data_dir, "processed", "LC_paths.shp"),
  quiet = TRUE
)

# make sure paths are in geographic coordinates
current_lines_sf <- st_transform(current_lines_sf, crs = st_crs(crs_geo))
```

# 3. Helper functions

```{r helper-functions}
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

make_pair_points <- function(pair_table, crs_string) {
  p1 <- SpatialPoints(
    coords = pair_table[, c("long1", "lat1")],
    proj4string = CRS(crs_string)
  )
  p2 <- SpatialPoints(
    coords = pair_table[, c("long2", "lat2")],
    proj4string = CRS(crs_string)
  )
  list(p1 = p1, p2 = p2)
}

split_lines_list <- function(lines_sf) {
  lines_sp <- as(lines_sf, "Spatial")
  lapply(seq_len(length(lines_sp)), function(i) lines_sp[i, ])
}

extract_env_along_paths <- function(lines_sf, env, geo_dist, pair_table, n_cores = 4) {
  sp_lines_list <- split_lines_list(lines_sf)

  cl <- makeCluster(n_cores)
  registerDoParallel(cl)
  clusterExport(cl, "get_mode")

  env_summ <- foreach(
    i = seq_along(sp_lines_list),
    .combine = rbind,
    .packages = "raster"
  ) %dopar% {
    vals <- extract(env, sp_lines_list[[i]])[[1]]

    means <- colMeans(vals, na.rm = TRUE)
    medians <- apply(vals, 2, median, na.rm = TRUE)
    modes <- apply(vals, 2, get_mode)

    data.frame(t(c(means, medians, modes)))
  }

  geo_dist_sum <- foreach(
    i = seq_along(sp_lines_list),
    .combine = c,
    .packages = "raster"
  ) %dopar% {
    vals <- extract(geo_dist, sp_lines_list[[i]])[[1]]
    sum(vals, na.rm = TRUE)
  }

  stopCluster(cl)

  lay_names <- names(env)
  new_names <- c(
    paste0(lay_names, "_mean"),
    paste0(lay_names, "_median"),
    paste0(lay_names, "_mode")
  )
  colnames(env_summ) <- new_names

  res <- cbind(
    pair_table,
    data.frame(pix_dist = geo_dist_sum),
    env_summ
  )

  res
}

fit_rf_mean_model <- function(path_table, ntree = 500, seed = 1234) {
  predictor_vars <- c(
    "BIO1_mean", "BIO2_mean", "BIO3_mean", "BIO4_mean",
    "BIO5_mean", "BIO6_mean", "BIO7_mean",
    "BIO8S_mean", "BIO9S_mean", "BIO10S_mean", "BIO11S_mean",
    "BIO12_mean", "BIO13_mean", "BIO14_mean", "BIO15_mean",
    "BIO16S_mean", "BIO17S_mean", "BIO18S_mean", "BIO19S_mean",
    "slope_mean", "alt_mean", "lakes_mean", "riv_3km_mean",
    "samp_20km_mean", "pix_dist"
  )

  rf_mean_data <- path_table[, c("CSEdistance", predictor_vars)]
  names(rf_mean_data) <- gsub("_mean$", "", names(rf_mean_data))

  set.seed(seed)
  rf_model <- tuneRF(
    x = rf_mean_data[, -1],
    y = rf_mean_data$CSEdistance,
    ntreeTry = ntree,
    stepFactor = 1.5,
    improve = 0.01,
    trace = TRUE,
    plot = FALSE,
    doBest = TRUE,
    importance = TRUE
  )

  rf_model
}

predict_rf_surface <- function(envstack, rf_model, out_file = NULL) {
  pred <- predict(envstack, rf_model, type = "response")

  if (!is.null(out_file)) {
    writeRaster(pred, out_file, format = "GTiff", overwrite = TRUE)
  }

  pred
}

build_transition_from_cost <- function(cost_raster) {
  cost_raster[cost_raster <= 0] <- NA

  tr <- transition(
    x = 1 / cost_raster,
    transitionFunction = mean,
    directions = 8
  )

  geoCorrection(tr, type = "c")
}

shortest_path_one_pair <- function(tr, x1, y1, x2, y2, crs_string) {
  pt1 <- SpatialPoints(matrix(c(x1, y1), ncol = 2), proj4string = CRS(crs_string))
  pt2 <- SpatialPoints(matrix(c(x2, y2), ncol = 2), proj4string = CRS(crs_string))

  shortestPath(tr, pt1, pt2, output = "SpatialLines")
}

make_lcp_paths_from_surface <- function(pair_table, cost_raster, crs_string, n_cores = 4) {
  tr <- build_transition_from_cost(cost_raster)

  cl <- makeCluster(n_cores)
  registerDoParallel(cl)
  clusterExport(cl, c("tr", "crs_string", "shortest_path_one_pair"), envir = environment())

  path_list <- foreach(
    i = seq_len(nrow(pair_table)),
    .packages = c("gdistance", "sp")
  ) %dopar% {
    shortest_path_one_pair(
      tr = tr,
      x1 = pair_table$long1[i],
      y1 = pair_table$lat1[i],
      x2 = pair_table$long2[i],
      y2 = pair_table$lat2[i],
      crs_string = crs_string
    )
  }

  stopCluster(cl)

  path_sfc <- st_sfc(lapply(path_list, st_as_sfc), crs = st_crs(crs_string))
  path_sfc <- do.call(c, path_sfc)

  st_sf(
    id = pair_table$id,
    geometry = path_sfc
  )
}

mean_abs_raster_change <- function(r1, r2) {
  d <- abs(r1 - r2)
  cellStats(d, stat = "mean", na.rm = TRUE)
}
```

# 4. Run iterative path extraction and model updating

```{r iterative-loop, eval=FALSE}
# store outputs from each iteration
iter_summary <- data.frame(
  iteration = integer(),
  oob_mse = numeric(),
  percent_var_explained = numeric(),
  mean_raster_change = numeric(),
  stringsAsFactors = FALSE
)

pred_prev <- NULL

for (iter in seq_len(n_iter)) {
  cat("\n-----------------------------\n")
  cat("Iteration", iter, "of", n_iter, "\n")
  cat("-----------------------------\n")

  # 1. extract summaries along current paths
  path_table <- extract_env_along_paths(
    lines_sf = current_lines_sf,
    env = env,
    geo_dist = geo_dist,
    pair_table = G.table,
    n_cores = n_cores
  )

  write.csv(
    path_table,
    file.path(results_dir, paste0("Gff_cse_envCostPaths_iter", iter, ".csv")),
    row.names = FALSE
  )

  # 2. fit RF model using mean predictors only
  rf_model <- fit_rf_mean_model(
    path_table = path_table,
    ntree = 500,
    seed = 1000 + iter
  )

  saveRDS(
    rf_model,
    file.path(results_dir, paste0("rf_mean_full_tuned_iter", iter, ".rds"))
  )

  # 3. predict new surface
  pred_file <- file.path(results_dir, paste0("fullRF_CSE_iter", iter, ".tif"))
  pred_raster <- predict_rf_surface(
    envstack = envstack,
    rf_model = rf_model,
    out_file = pred_file
  )

  # 4. compare with previous iteration
  raster_change <- NA_real_
  if (!is.null(pred_prev)) {
    raster_change <- mean_abs_raster_change(pred_raster, pred_prev)
    cat("Mean absolute raster change from previous iteration:", raster_change, "\n")
  }

  # 5. record model summary
  iter_summary <- rbind(
    iter_summary,
    data.frame(
      iteration = iter,
      oob_mse = rf_model$mse[rf_model$ntree],
      percent_var_explained = rf_model$rsq[rf_model$ntree] * 100,
      mean_raster_change = raster_change
    )
  )

  write.csv(
    iter_summary,
    file.path(results_dir, "iterative_rf_summary.csv"),
    row.names = FALSE
  )

  # 6. update paths for next iteration
  if (iter < n_iter) {
    current_lines_sf <- make_lcp_paths_from_surface(
      pair_table = G.table,
      cost_raster = pred_raster,
      crs_string = crs_geo,
      n_cores = n_cores
    )

    st_write(
      current_lines_sf,
      file.path(data_dir, "processed", paste0("LC_paths_iter", iter + 1, ".shp")),
      delete_layer = TRUE,
      quiet = TRUE
    )
  }

  pred_prev <- pred_raster
}
```

# 5. Review iteration summaries

```{r review-summary, eval=FALSE}
iter_summary

plot(
  iter_summary$iteration,
  iter_summary$percent_var_explained,
  type = "b",
  xlab = "Iteration",
  ylab = "% Variance Explained"
)

plot(
  iter_summary$iteration,
  iter_summary$oob_mse,
  type = "b",
  xlab = "Iteration",
  ylab = "OOB MSE"
)

plot(
  iter_summary$iteration,
  iter_summary$mean_raster_change,
  type = "b",
  xlab = "Iteration",
  ylab = "Mean absolute raster change"
)
```

# 6. Notes

1. Iteration 1 uses the existing lake-avoiding least-cost paths as the starting point.  
2. Each subsequent iteration uses the RF-predicted surface from the previous iteration to generate new least-cost paths.  
3. Environmental summaries are re-extracted along the updated paths each round.  
4. The RF is refit each round using the mean environmental summaries plus path-based geographic distance.  
5. The table `iterative_rf_summary.csv` can be used to judge whether the procedure stabilizes before 10 iterations.  

# 7. Likely places to adjust

1. Whether `samp_20km` should be neutralized before projection.  
2. Whether you want to keep `pix_dist` as summed path length along the current path each round.  
3. Whether you want to parallelize path building or keep it serial for easier debugging.  
4. Whether you want an early stopping rule when raster change becomes very small.  
5. Whether you want to save only the final iteration or all intermediate products.  
