---
title: "Spatial Evaluation - Full RF model of raw CSE, alternative km projections"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../scripts_knitted_md", 
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/knitted_mds/" 
)
```
  
# Setup  

RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 16 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-np (allows multiple simultaneous jobs automatically now)  
- **Memory per job:** 400G (cluster limit: 1000G total; avoid exceeding half)  

## Load libraries
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(randomForest)
```
## Inputs
  - `../input/Gff_cse_envCostPaths.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2), same input used for building model.
  - `../results_dir/fullRF_CSE.tif` - Final full model projected estimates of CSE at 1 pixel geodist.
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
  
## Outputs   
Major outputs from this script are saved in two locations:  
**Small text outputs saved to results_dir within git repo:**
 - `spatial_eval_rawCSE_LCPpred.csv` Pairwise least-cost path summaries for each site pair: LCP_mean, LCP_sum.
 - `spatial_eval_rawCSE_cber.csv` Pairwise circuit-based effective resistance (CBER) values for each site pair.
 - `spatial_eval_rawCSE_predictions.csv` Combined pairwise evaluation table containing observed CSE and spatial predictions for each site pair: CSE, LCP_mean, LCP_sum, CBER.
 - `spatial_eval_rawCSE_eval_metrics.csv` Summary table of evaluation metrics for each spatial summary method (LCP_mean, LCP_sum, CBER): RMSE, R², MSE, MAE, Correlation.

**Large or reusable outputs saved to results_dir_big:**
 - `least_cost_paths/LC_paths_fullRF_rawCSE.shp` Saved least-cost-path spatial lines for all evaluated site pairs. This is a reusable spatial output and does not need to be recomputed unless the resistance surface changes.
 - `spatial_eval_rawCSE_LCPpred.csv` Backup copy of least-cost path summaries.
 - `spatial_eval_rawCSE_cber.csv` Backup copy of circuit-based effective resistance (CBER) results.
 - `spatial_eval_rawCSE_predictions.rds` R object version of the combined evaluation table (eval_results) for quick reloading in later scripts.
 - `spatial_eval_rawCSE_predictions.csv` Backup copy of the combined evaluation table saved in the large-results directory.
 - `spatial_eval_rawCSE_predictions_calibrated.rds` R object version of the evaluation table with calibrated predictions (eval_df), including LCP_mean_cal, LCP_sum_cal, and CBER_cal.
 - `spatial_eval_rawCSE_eval_metrics.rds` R object version of the evaluation metrics table (metrics_df) for quick reloading in later scripts.
 - `spatial_eval_rawCSE_eval_metrics.csv` Backup copy of the evaluation metrics table saved in the large-results directory.

# Code outline

Estimate evaluation metrics for multiple projection   
Sum across least-cost path (sum LCP)  

Metrics to calculate for this screening step: 

Primary metrics:  
1. correlation (Correlation here answers: does the method recover relative ordering?)
2. R² (1 − SSE/SST, after correcting for scale mismatch) 

Secondary evaluation metrics:  
3. RMSE (after correcting for scale mismatch)
4. MSE (after correcting for scale mismatch) 
5. MAE(after correcting for scale mismatch) 

NOTE: RMSE, MSE, MAE, and R2 after scaling answer: after correcting for simple scale mismatch, how well does the method recover observed CSE?). These screening metrics are used only to decide which response formulation and spatial summary to carry forward into the final LOPOCV validation framework and permutation-based significance testing.

# Step 1: Inputs
```{r directories-inputs}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

######################################
# What surface are you using?
surf_file  <- file.path(results_dir_big, "fullRF_CSE.tif")
########################################

lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)

# Load env stack with named layers
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Neutralize sampling layer to average
env$samp_20km <- 1.027064e-11 #neutralize sampling bias

# Load rdf of final model
rf_predicted <- readRDS(file.path(results_dir_big, "rf_mean_full_tuned.rds"))
```

# Step 2. Functions

```{r kdist}
# --------------------------------------------------
# Screening: compare projection distances using LCP_sum only
# --------------------------------------------------

# projection distances to test
proj_km <- c(1, 2, 5, 10, 20, 50)

# distance layer name in model
dist_layer <- "geo_dist"

# keep only within-cluster comparisons
cse_df_eval <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates
sites_coords <- bind_rows(
  cse_df_eval %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df_eval %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# pair table
site_pairs <- cse_df_eval %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# calibration function
calibrate_pred <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  df <- data.frame(obs = obs[keep], pred = pred[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(obs))
  out[keep] <- predict(fit, newdata = data.frame(pred = pred[keep]))
  out
}

# metrics function
# correlation uses raw predictions
# RSQ and error metrics use calibrated predictions
calc_metrics <- function(obs, pred_raw, pred_cal) {
  keep <- complete.cases(obs, pred_raw, pred_cal)

  obs      <- obs[keep]
  pred_raw <- pred_raw[keep]
  pred_cal <- pred_cal[keep]

  sse  <- sum((obs - pred_cal)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred_cal)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred_cal))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred_raw)

  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Correlation = corr
  )
}

# output directory
screen_dir <- file.path(results_dir_big, "projection_screening_LCPsum")
dir.create(screen_dir, showWarnings = FALSE, recursive = TRUE)

all_metrics <- list()
```

# Step 3. Test 1 km versus 5 km
```{r test}
kdist <- 1
env_k1 <- env
env_k1[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k1[["pix_dist"]]))
pred1 <- predict(env_k1, rf_predicted, type = "response")

kdist <- 5
env_k5 <- env
env_k5[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k5[["pix_dist"]]))
pred5 <- predict(env_k5, rf_predicted, type = "response")

cellStats(abs(pred1 - pred5), stat = "max", na.rm = TRUE)
cellStats(abs(pred1 - pred5), stat = "mean", na.rm = TRUE)

kdist <- 150
env_k5 <- env
env_k5[["pix_dist"]] <- setValues(env[[1]], kdist)

summary(values(env_k5[["pix_dist"]]))
pred5 <- predict(env_k5, rf_predicted, type = "response")

cellStats(abs(pred1 - pred5), stat = "max", na.rm = TRUE)
cellStats(abs(pred1 - pred5), stat = "mean", na.rm = TRUE)
```

1 vs 5 are identical. 1 vs 150 km are not. Where is the split?

# Step 4. Find the Split
```{r find-split}
kvals <- c(1, 2, 5, 10, 20, 50, 100, 150)

pred_list <- lapply(kvals, function(k) {
  env_k <- env
  env_k[["pix_dist"]] <- setValues(env[[1]], k)
  predict(env_k, rf_predicted, type = "response")
})

names(pred_list) <- paste0("k", kvals)

expand.grid(i = seq_along(kvals), j = seq_along(kvals)) |>
  dplyr::filter(i < j) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    k1 = kvals[i],
    k2 = kvals[j],
    mean_diff = cellStats(abs(pred_list[[i]] - pred_list[[j]]), stat = "mean", na.rm = TRUE),
    max_diff  = cellStats(abs(pred_list[[i]] - pred_list[[j]]), stat = "max", na.rm = TRUE)
  ) |>
  dplyr::select(k1, k2, mean_diff, max_diff)
```

1, 2, 5 km are identical
mean_diff = 0
max_diff = 0
→ these produce exactly the same raster
10 km is barely different
mean diff ~ 4e-05
→ essentially negligible
20 km = small change
mean diff ~ 3e-04
→ still tiny
50+ km = real change
mean diff jumps to ~ 4e-03 → 1e-02 → 4e-02
→ now you’re actually moving into different RF partitions

# Step 5. Loop through 1-100 km

```{r loop, eval=FALSE}
proj_km <- c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)

n_cores <- 16
cl <- makeCluster(n_cores)
registerDoParallel(cl)

for (kdist in proj_km) {

  cat("\n----------------------------\n")
  cat("Running projection at", kdist, "km\n")
  cat("----------------------------\n")

  # -----------------------------------
  # Step 1: project surface at kdist
  # -----------------------------------
  env_k <- env

  # assign constant raster for projection distance
  env_k[["pix_dist"]] <- setValues(env[[1]], kdist)

  cse_surface <- predict(env_k, rf_predicted, type = "response")

  # align lake mask if needed
  if (!compareRaster(cse_surface, lake_mask,
                     extent = TRUE, rowcol = TRUE,
                     crs = TRUE, res = TRUE,
                     stopiffalse = FALSE)) {
    lake_mask_use <- projectRaster(lake_mask, cse_surface, method = "ngb")
  } else {
    lake_mask_use <- lake_mask
  }

  # assign lakes high cost
  cse_surface[lake_mask_use[] == 1] <- max(values(cse_surface), na.rm = TRUE)

  # save projected surface
  writeRaster(
    cse_surface,
    filename = file.path(screen_dir, paste0("fullRF_CSE_", kdist, "km.tif")),
    format = "GTiff",
    overwrite = TRUE
  )

  # -----------------------------------
  # Step 2: prepare sites in raster CRS
  # -----------------------------------
  sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
    st_transform(crs(cse_surface))

  sites_sp <- as(sites_df, "Spatial")
  site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

  # -----------------------------------
  # Step 3: build transition object
  # -----------------------------------
  resistance_rast <- cse_surface
  min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
  resistance_rast[resistance_rast <= 0] <- min_pos

  conductance_rast <- 1 / resistance_rast
  tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)
  tr_lcp <- geoCorrection(tr, type = "c")

  # -----------------------------------
  # Step 4: recompute least-cost paths
  # -----------------------------------

  paths_list <- foreach(ii = seq_len(nrow(site_pairs)),
                        .packages = c("gdistance", "sp", "sf")) %dopar% {
    i <- site_index[site_pairs$Var1[ii]]
    j <- site_index[site_pairs$Var2[ii]]

    path <- tryCatch({
      shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
    }, error = function(e) NULL)

    if (!is.null(path)) {
      path_sf <- st_as_sf(path)
      path_sf$id <- site_pairs$id[ii]
      return(path_sf)
    } else {
      return(NULL)
    }
  }

  paths_list <- paths_list[!sapply(paths_list, is.null)]
  paths_sf <- do.call(rbind, paths_list)
  st_crs(paths_sf) <- st_crs(cse_surface)

  # extract raster values along each path
  path_vals <- raster::extract(cse_surface, as(paths_sf, "Spatial"))

  lcp_summary <- data.frame(
    id = paths_sf$id,
    LCP_sum = sapply(path_vals, function(x) {
      if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
    })
  )

  # -----------------------------------
  # Step 5: combine and calibrate
  # -----------------------------------
  eval_results <- site_pairs %>%
    left_join(lcp_summary, by = "id")

  eval_results$LCP_sum_cal <- calibrate_pred(eval_results$CSE, eval_results$LCP_sum)

  # -----------------------------------
  # Step 6: calculate metrics
  # -----------------------------------
  metrics_k <- calc_metrics(
    obs = eval_results$CSE,
    pred_raw = eval_results$LCP_sum,
    pred_cal = eval_results$LCP_sum_cal
  ) %>%
    mutate(
      projection_km = kdist,
      method = "LCP_sum"
    ) %>%
    dplyr::select(projection_km, method, everything())

  # save outputs
  write.csv(
    eval_results,
    file.path(screen_dir, paste0("eval_results_LCPsum_", kdist, "km.csv")),
    row.names = FALSE
  )

  write.csv(
    metrics_k,
    file.path(screen_dir, paste0("eval_metrics_LCPsum_", kdist, "km.csv")),
    row.names = FALSE
  )

  all_metrics[[paste0("proj_", kdist, "km")]] <- metrics_k
}

stopCluster(cl)

# bind all metrics
metrics_all <- bind_rows(all_metrics)

write.csv(
  metrics_all,
  file.path(screen_dir, "projection_distance_comparison_metrics_LCPsum.csv"),
  row.names = FALSE
)

metrics_all
```

```{r review-km-projections}

km_summary<- read.csv("/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/projection_screening_LCPsum/projection_distance_comparison_metrics_LCPsum.csv")

plot(
  km_summary$projection_km,
  km_summary$RSQ,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "% Variance Explained"
)

plot(
  km_summary$projection_km,
  km_summary$MSE,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "OOB MSE"
)

plot(
  km_summary$projection_km,
  km_summary$Correlation,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "Mean absolute raster change"
)

# Open PDF device
pdf("/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/projections_km_fig.pdf", width = 7, height = 3)

# Set up 1 row, 3 columns
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))  # adjust margins if needed

# (a) % Variance Explained
plot(
  km_summary$projection_km,
  km_summary$RSQ,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "% Variance Explained"
)
mtext("(a)", side = 3, adj = 0, line = 0.5, font = 2)

# (b) OOB MSE
plot(
  km_summary$projection_km,
  km_summary$MSE,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "OOB MSE"
)
mtext("(b)", side = 3, adj = 0, line = 0.5, font = 2)

# (c) Correlation
plot(
  km_summary$projection_km,
  km_summary$Correlation,
  type = "b",
  xlab = "Projection Scale in Km",
  ylab = "Correlation"
)
mtext("(c)", side = 3, adj = 0, line = 0.5, font = 2)

# Close PDF device
dev.off()

# Reset plotting layout (optional)
par(mfrow = c(1, 1))
```