---
title: "Spatial Evaluation - Full RF model of raw CSE"
author: "Norah Saarman"
date: "2025-06-23"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 8 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-np (allows multiple simultaneous jobs automatically now)  
- **Memory per job:** 200G (cluster limit: 1000G total; avoid exceeding half)    


# Plan for this script:
This is a screening step (full model only). Use the fitted full model to compare candidate response formulations and spatial summaries.  

Response formulations to compare:  
- raw CSE (this script, 7a)  
- residuals from simple IBD model (script 7b)  
- CSE/km (script 7c)

**For this script (7a):**  
Run spatial evaluation across the raw CSE resistance surface (projected estimates of raw CSE at 1 pixel geodist) to estimate evaluation metrics by 3 methods:   
1. Mean across least-cost path (mean LCP)   
2. Sum across least-cost path (sum LCP)  
3. Circuit-based effective resistance (CBER)  

Metrics to calculate for this screening step: 

Primary metrics:  
1. RMSE  
2. R² (1 − SSE/SST)  

NOTE: RMSE is included alongside R² because it measures the absolute magnitude of prediction error in the original response scale, while R² measures variance explained relative to the mean model. Together they distinguish between models that explain variation and models that produce accuratepredictions.

Secondary evaluation metrics:  
3. MSE  
4. MAE  
5. correlation  

NOTE: These screening metrics are used only to decide which response formulation and spatial summary to carry forward into the final LOPOCV validation framework and permutation-based significance testing.

Idea: Can these estimates later be used as starting points for next iteration of the model?

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

# Inputs
  - `../input/Gff_cse_envCostPaths.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2), same input used for building model.
  - `../results_dir/fullRF_CSE.tif` - Final full model projected estimates of CSE at 1 pixel geodist.
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
  
# Outputs   
Major outputs from this script are saved in two locations:  
## Small text outputs saved to results_dir within git repo:
 - `spatial_eval_rawCSE_predictions.csv` Pairwise evaluation table containing observed CSE and spatial predictions for each site pair: CSE, LCP_mean, LCP_sum, CBER
 - `spatial_eval_rawCSE_metrics.csv` Summary table of evaluation metrics for each spatial summary method (LCP_mean, LCP_sum, CBER): RMSE, R², MSE, MAE, Correlation.  
## Large or reusable outputs saved to results_dir_big:
 - `least_cost_paths/LC_paths_fullRF_rawCSE.shp` Saved least-cost-path spatial lines for all evaluated site pairs. This is a reusable spatial output and does not need to be recomputed unless the resistance surface changes.  
 - `spatial_eval_rawCSE_predictions.rds` R object version of the full pairwise evaluation table (eval_df) for quick reloading in later scripts.  
 - `spatial_eval_rawCSE_predictions.csv`  Backup copy of the pairwise evaluation table saved in the large-results directory.  
 - `spatial_eval_rawCSE_metrics.rds` R object version of the evaluation metrics table (metrics_df) for quick reloading in later scripts.  
 - `spatial_eval_rawCSE_metrics.csv` Backup copy of the evaluation metrics table saved in the large-results directory.

# Step 1: Inputs
```{r directories-inputs}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")
surf_file  <- file.path(results_dir_big, "fullRF_CSE.tif")
lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")


# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)
```
  
# Step 2: Check raster alignment and assign high cost to water

```{r check-alignment}
if (!compareRaster(cse_surface, lake_mask,
                   extent = TRUE, rowcol = TRUE,
                   crs = TRUE, res = TRUE,
                   stopiffalse = FALSE)) {
  lake_mask <- projectRaster(lake_mask, cse_surface, method = "ngb")
}

# Inspect projected raw CSE values before altering water cells
summary(values(cse_surface))

# Keep water traversable but very costly
cse_surface[lake_mask[] == 1] <- 2

# Check again after assigning lakes
summary(values(cse_surface))

# Quick plot
plot(cse_surface, main = "Projected raw CSE surface with water = high cost")

# Check column names of cse_df
names(cse_df)
```

# Step 3: Prepare pair table, site objects, and transition object  

Diagonal movement is allowed in the 16-way transition matrix, and conductance is corrected for geographic distance, with geoCorrection(type = "c"), so diagonal steps are appropriately penalized relative to orthogonal steps.

```{r prep-transition}
# keep only within-cluster comparisons, matching your earlier workflow
cse_df <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates from both columns
sites_coords <- bind_rows(
  cse_df %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# convert to sf and then to SpatialPoints in raster CRS
sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(cse_surface))

sites_sp <- as(sites_df, "Spatial")

# lookup index for site names
site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

# pair table for evaluation
site_pairs <- cse_df %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# treat projected raw CSE as local resistance
resistance_rast <- cse_surface

# protect against zero or negative values before inversion
min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
resistance_rast[resistance_rast <= 0] <- min_pos

# convert resistance to conductance for gdistance
conductance_rast <- 1 / resistance_rast

# transition object:
# directions = 16 allows diagonal/extended neighborhood movement
# geoCorrection(type = "c") corrects conductance for inter-cell geographic distance,
# including longer diagonal connections
tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)
tr_corr <- geoCorrection(tr, type = "c")
```

# Step 4: Sanity checks
```{r sanity}
nrow(cse_df)
nrow(site_pairs)
nrow(sites_coords)

head(site_pairs)

plot(cse_surface, main = "Projected raw CSE surface")
plot(sites_sp, add = TRUE, pch = 16, cex = 0.5)
```  

# Step 5: Test one least-cost path before parallel run
```{r test-lcp}
# choose first pair in site_pairs
i <- site_index[site_pairs$Var1[100]]
j <- site_index[site_pairs$Var2[100]]

test_path <- shortestPath(tr_corr, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")

plot(cse_surface, main = "Test least-cost path")
lines(test_path, col = "red", lwd = 2)
points(sites_sp[c(i, j), ], pch = 16)
```  

# Step 6: Compute least-cost paths in parallel (once) and save output
```{r save-lcp, eval = FALSE}
output_dir <- file.path(results_dir_big, "least_cost_paths")
dir.create(output_dir, showWarnings = FALSE)

# Set number of cores and register cluster
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

paths_list <- foreach(k = 1:nrow(site_pairs),
                      .packages = c("gdistance", "sp", "sf")) %dopar% {

  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]

  path <- tryCatch({
    shortestPath(tr_corr, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

stopCluster(cl)

# drop NULLs and combine
paths_list <- paths_list[!sapply(paths_list, is.null)]
paths_sf <- do.call(rbind, paths_list)

# assign CRS and transform back to geographic coordinates for saving
st_crs(paths_sf) <- st_crs(cse_surface)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# join observed CSE values
paths_sf <- left_join(paths_sf, site_pairs, by = "id")

# save paths
st_write(paths_sf,
         file.path(output_dir, "LC_paths_fullRF_rawCSE.shp"),
         delete_layer = TRUE)
```

# Step 7: Extract raw CSE values along LCPs and calculate mean and sum
```{r extract-mean-sum, eval = FALSE}
# if needed later, reload:
# paths_sf <- st_read(file.path(results_dir_big, "least_cost_paths", "LC_paths_fullRF_rawCSE.shp"), quiet = TRUE)

# transform paths to raster CRS for extraction
paths_extract <- st_transform(paths_sf, crs = crs(cse_surface))

# extract raster values along each path
path_vals <- raster::extract(cse_surface, as(paths_extract, "Spatial"))

# summarize values along each least-cost path
lcp_summary <- data.frame(
  id = paths_extract$id,
  LCP_mean = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
  }),
  LCP_sum = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
  })
)

# join back to pair table
eval_df <- site_pairs %>%
  left_join(lcp_summary, by = "id")

head(eval_df)
summary(eval_df$LCP_mean)
summary(eval_df$LCP_sum)


# Save outputs
saveRDS(
  eval_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_predictions.rds")
)

write.csv(
  eval_df,
  file.path(results_dir, "spatial_eval_rawCSE_predictions.csv"),
  row.names = FALSE
)

write.csv(
  eval_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_predictions.csv"),
  row.names = FALSE
)
```  

# Step 8: Compute circuit-based effective resistance (CBER)  
Also save output from steps 6-8 to file again...
```{r cber, eval = FALSE}

n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

cber_list <- foreach(k = 1:nrow(site_pairs),
                     .packages = c("gdistance", "sp")) %dopar% {

  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]

  val <- tryCatch({
    dmat <- commuteDistance(tr_corr, sites_sp[c(i, j), ])
    as.numeric(dmat[1, 2])
  }, error = function(e) NA_real_)

  data.frame(
    id = site_pairs$id[k],
    CBER = val
  )
}

stopCluster(cl)

cber_df <- bind_rows(cber_list)

eval_df <- eval_df %>%
  left_join(cber_df, by = "id")

# Print evaluation table
print(head(eval_df))
summary(eval_df$CBER)

# Save outputs
saveRDS(
  eval_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_predictions.rds")
)

write.csv(
  eval_df,
  file.path(results_dir, "spatial_eval_rawCSE_predictions.csv"),
  row.names = FALSE
)

write.csv(
  eval_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_predictions.csv"),
  row.names = FALSE
)
```

# Step 9: Calculate evaluation metrics from eval_df (load from file) 
```{r}
# load eval_df from file
eval_df <- read.csv(file.path(results_dir, "spatial_eval_rawCSE_predictions.csv"))
  
calc_metrics <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  obs  <- obs[keep]
  pred <- pred[keep]
  
  sse  <- sum((obs - pred)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred)
  
  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Correlation = corr
  )
}

metrics_df <- bind_rows(
  calc_metrics(eval_df$CSE, eval_df$LCP_mean) %>% mutate(method = "LCP_mean"),
  calc_metrics(eval_df$CSE, eval_df$LCP_sum)  %>% mutate(method = "LCP_sum"),
  calc_metrics(eval_df$CSE, eval_df$CBER)     %>% mutate(method = "CBER")
) %>%
  dplyr::select(method, everything())

# Print metrics
print(metrics_df)

# Save outputs
saveRDS(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_metrics.rds")
)

write.csv(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_rawCSE_metrics.csv"),
  row.names = FALSE
)

write.csv(
  metrics_df,
  file.path(results_dir, "spatial_eval_rawCSE_metrics.csv"),
  row.names = FALSE
)

```

# This script should be followed by:

First, 7b-7f. Choose one response formulation and spatial summary.

Then, after selecting the best response formulation and spatial summary:  

Run the final validation framework, full LOPOCV evaluation, including:  
- non-spatial evaluation  
- spatial evaluation (mean LCP, sum LCP, or CBER depending on screening results)  
- Calculate final metrics on held-out data (RMSE, R², MSE, MAE, correlation)

Permutation test for inference  
- Run ~100 permutations where the response is randomized.

For each permutation:  
- rerun the full LOPOCV workflow  
- recompute spatial predictions  
- calculate the same evaluation metrics  
Compare observed metrics (especially RMSE and R²) to the permutation distribution to test whether predictive performance is better than expected by chance.
