---
title: "Least-cost-paths across full RF model resistance surface"
author: "Norah Saarman"
date: "2025-06-23"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 8 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
- **Memory per job:** 200G (cluster limit: 1000G total; avoid exceeding half)    

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

# Define Paths to directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
output_dir <- paste0(results_dir,"lopocv_spatial")
dir.create(output_dir, showWarnings = FALSE)

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84
```

# Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../results_dir/fullRF_CSE_resistance.tif` - Final full model projected resistance surface
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
# Outputs  
  - `../results_dir/LC_paths_fullRF.shp"` - 
  
# Step 1: Prepare least-cost paths from full resistance surface
**NOTE:** eval = FALSE so that it skips on knit
```{r paths, eval=FALSE}
# Set number of cores and register cluster
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Load full model prediction raster as resistance surface
resistance_rast <- raster(paste0(results_dir, "fullRF_CSE_resistance.tif"))

# Load lake mask and assign high CSE (e.g., 2)
lake_mask <- raster(file.path(data_dir, "processed", "lake_binary.tif"))
resistance_rast[lake_mask[,] == 1] <- 2

# Convert resistance raster to transition object
tr <- transition(1/resistance_rast, transitionFunction = mean, directions = 16)
tr_corr <- geoCorrection(tr, type = "c")

# Load V.table (after filtering)
V.table_full <- read.csv("../input/Gff_cse_envCostPaths.csv")
V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB") %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(id = paste(Var1, Var2, sep = "_"))

# Get unique site coordinates
sites_coords <- V.table %>%
  dplyr::select(Site = Var1, lon = long1, lat = lat1) %>%
  distinct()

sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(resistance_rast))

# Create named SpatialPoints
sites_sp <- as(sites_df, "Spatial")
site_index <- setNames(seq_len(nrow(sites_df)), sites_df$Site)

# Get list of site pairs to extract
site_pairs <- V.table %>%
  dplyr::select(Var1, Var2, id) %>%
  distinct()

# Parallel least-cost path computation with metadata
paths_list <- foreach(k = 1:nrow(site_pairs), .packages = c("gdistance", "sp", "sf")) %dopar% {
  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]
  
  path <- tryCatch({
    shortestPath(tr_corr, sites_sp[i,], sites_sp[j,], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

# Stop cluster
stopCluster(cl)

# Combine list into single sf object
paths_sf <- do.call(rbind, paths_list)
st_crs(paths_sf) <- st_crs(resistance_rast)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# Join CSEdistance from V.table into lcp_sf
paths_sf <- left_join(paths_sf, V.table[, c("id", "CSEdistance")], by = "id")
paths_sf <- paths_sf %>%
  rename(CSE = CSEdistance) # note rename!!!

# Save to shapefile
st_write(paths_sf, file.path(results_dir, "LC_paths_fullRF.shp"), delete_layer = TRUE)

```
# Visualize Least Cost Paths
```{r map-paths}
# Load shapefile of least-cost paths
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp")) 
lcp_sf <- st_transform(lcp_sf, crs = st_crs(crs_geo))

# Load Uganda country boundary and lakes
uganda <- ne_countries(scale = "medium", country = "Uganda", returnclass = "sf") %>%
  st_transform(crs = st_crs(crs_geo))
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>%
  st_transform(st_crs(crs_geo))

# (Optional) Load resistance raster for background
resistance_rast <- raster(file.path(results_dir, "fullRF_CSE_resistance.tif"))

xlim <- c(28.6, 35.4)
ylim <- c(-1.500000 , 4.733333)

# Basic plot with paths and country
ggplot() +
  geom_raster(data = as.data.frame(rasterToPoints(resistance_rast)) %>%
                setNames(c("x", "y", "resist")),
              aes(x = x, y = y, fill = resist)) +
  scale_fill_viridis_c(option = "C", na.value = "transparent") +
  geom_sf(data = lakes, fill = "gray80", color = NA) +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = lcp_sf, color = "dark blue", size = 0.4, alpha = 0.8) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Least-Cost Paths on Resistance Surface",
       x = "Longitude", y = "Latitude", fill = "Resistance") 

```


# Compare 3 spatial predicitons: predicted pairwise value for each site pair:

## 1. mean along LCP

## 2. sum along LCP

## 3. circuit-based effective resistance

R package 'gdistance' commuteDistance() function is mathematically identical to the effective resistance used in Circuitscape. If a reviewer asks whether we used Circuitscape, we can say that the same circuit-theory metric was computed directly in R using gdistance.

Once each method gives a pairwise prediction, you can compare those predictions to observed pairwise CSE using the same evaluation framework.

Evaluative metrics:

1. RMSE
2. MSE
3. MAE
4. R-squared from observed vs predicted (1 - SSE/SST; from observed vs predicted on the held-out data.)*
5. correlation between observed and predicted
6. permutation-based significance of those metrics
7. leave-one-population-out framework for inference of error across space

*Better than relying on whatever internal random forest R-squared was reported during training, because now you are comparing projected path-level summaries across methods.

Notes of caution:
1. Pairwise non-independence. All pairwise distances are not independent because each site appears in many pairs. So the absolute p-values from ordinary regression are not very trustworthy. Use RMSE and R-squared:
 - mainly as predictive comparison metrics
 - use permutation or your leave-one-population-out framework for inference rather than naive regression p-values
2. Circuit resistance assumes the raster represents local resistance per step. Be sure raster values can be interpreted as resistance before you run gedistance

```{r circuit-resistance}
library(raster)
library(gdistance)

# ------------------------------------------------
# 1. Load projected connectivity surface from your RF model
#    (assuming larger values = higher connectivity)
# ------------------------------------------------

rf_surface <- raster("rf_projection.tif")

# ------------------------------------------------
# 2. Convert connectivity surface to resistance surface
#    (high connectivity = low resistance)
# ------------------------------------------------

resistance <- max(values(rf_surface), na.rm = TRUE) - rf_surface

# ------------------------------------------------
# 3. Convert resistance to conductance (gdistance expects conductance)
# ------------------------------------------------

conductance <- 1 / resistance

# ------------------------------------------------
# 4. Build transition object (8 directions)
# ------------------------------------------------

tr <- transition(conductance, transitionFunction = mean, directions = 8)

# correct for geographic distance
tr <- geoCorrection(tr, type = "c")

# transition(..., directions = 8) allows diagonal pathways and produces more realistic connectivity surfaces.

# geoCorrection(type="c") adjusts transition values so that movement probabilities or conductance values are scaled appropriately for cell geometry. In other words, it accounts for the longer distance of diagonal connections and prevents the grid structure from biasing results.

# ------------------------------------------------
# 5. Site coordinates
# ------------------------------------------------

coords <- as.matrix(site_df[, c("x","y")])

# ------------------------------------------------
# 6. Compute circuit-based effective resistance
# ------------------------------------------------

res_matrix <- commuteDistance(tr, coords)

# ------------------------------------------------
# 7. Convert matrix to pairwise vector
# ------------------------------------------------
res_vector <- res_matrix[upper.tri(res_matrix)]

# observed genetic distances
obs_vector <- CSE_matrix[upper.tri(CSE_matrix)]

# -----------------------------------------------
# 8. Simple evaluation
# -----------------------------------------------

# R-squared
rsq <- cor(obs_vector, res_vector)^2

# RMSE
rmse <- sqrt(mean((obs_vector - res_vector)^2))

# MSE
mse <- mean((obs_vector - res_vector)^2)

print(rsq)
print(rmse)
print(mse)

# ------------------------------------------------
# 9. Quick diagnostic plot
# ------------------------------------------------

plot(res_vector, obs_vector,
     xlab = "Circuit-based resistance distance",
     ylab = "Observed CSE",
     pch = 16)

abline(lm(obs_vector ~ res_vector), col = "red")

```