---
title: "Spatial Evaluation – residuals (IBD by using LC lakes paths)"
author: "Norah Saarman"
date: "2026-04-01"
output:
  github_document:
    toc: true
---

RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 16 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-np (allows multiple simultaneous jobs automatically now)  
- **Memory per job:** 400G (cluster limit: 1000G total; avoid exceeding half)    

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## Load libraries
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```
## Inputs
  - `../input/Gff_cse_envCostPaths.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2), same input used for building model.
  - `../results_dir/"fullRF_residuals.tif"` - Final full model projected estimates of CSE at 1 pixel geodist.
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
  
## Outputs   
Major outputs from this script are saved in two locations:  
**Small text outputs saved to results_dir within git repo:**
 - `spatial_eval_residuals_LCPpred.csv` Pairwise least-cost path summaries for each site pair: LCP_mean, LCP_sum.
 - `spatial_eval_residuals_cber.csv` Pairwise circuit-based effective resistance (CBER) values for each site pair.
 - `spatial_eval_residuals_predictions.csv` Combined pairwise evaluation table containing observed CSE and spatial predictions for each site pair: CSE, LCP_mean, LCP_sum, CBER.
 - `spatial_eval_residuals_eval_metrics.csv` Summary table of evaluation metrics for each spatial summary method (LCP_mean, LCP_sum, CBER): RMSE, R², MSE, MAE, Correlation.

**Large or reusable outputs saved to results_dir_big:**
 - `least_cost_paths/LC_paths_fullRF_residuals.shp` Saved least-cost-path spatial lines for all evaluated site pairs. This is a reusable spatial output and does not need to be recomputed unless the resistance surface changes.
 - `spatial_eval_residuals_LCPpred.csv` Backup copy of least-cost path summaries.
 - `spatial_eval_residuals_cber.csv` Backup copy of circuit-based effective resistance (CBER) results.
 - `spatial_eval_residuals_predictions.rds` R object version of the combined evaluation table (eval_results) for quick reloading in later scripts.
 - `spatial_eval_residuals_predictions.csv` Backup copy of the combined evaluation table saved in the large-results directory.
 - `spatial_eval_residuals_predictions_calibrated.rds` R object version of the evaluation table with calibrated predictions (eval_df), including LCP_mean_cal, LCP_sum_cal, and CBER_cal.
 - `spatial_eval_residuals_eval_metrics.rds` R object version of the evaluation metrics table (metrics_df) for quick reloading in later scripts.
 - `spatial_eval_residuals_eval_metrics.csv` Backup copy of the evaluation metrics table saved in the large-results directory.
 
# Step 0: Script Outline
This is a screening step (full model only). Use the fitted full model to compare candidate response formulations and spatial summaries.  

Response formulations to compare:  
- raw CSE (this script, 7a)  
- residuals from simple IBD model (script 7b)  
- CSE/km (script 7c)

**For this script (7b):**  
Run spatial evaluation across the residuals resistance surface (projected estimates of residuals) to estimate evaluation metrics by 3 methods:   
1. Mean across least-cost path (mean LCP)   
2. Sum across least-cost path (sum LCP)  
3. Circuit-based effective resistance (CBER)  

Metrics to calculate for this screening step: 

Primary metrics:  
1. correlation (Correlation here answers: does the method recover relative ordering?)
2. R² (1 − SSE/SST, after correcting for scale mismatch) 

Secondary evaluation metrics:  
3. RMSE (after correcting for scale mismatch)
4. MSE (after correcting for scale mismatch) 
5. MAE(after correcting for scale mismatch) 

NOTE: RMSE, MSE, MAE, and R2 after scaling answer: after correcting for simple scale mismatch, how well does the method recover observed CSE?). These screening metrics are used only to decide which response formulation and spatial summary to carry forward into the final LOPOCV validation framework and permutation-based significance testing.

IDEA: LCP and CBER might serve as starting points for a next iteration of resistance optimization. Especially CBER. Current density (circuit theory) – longer-term idea:  
- Would show where movement concentrates across the landscape, and identify broad corridors, alternative routes, and pinch points.   
- More biologically realistic than a single LCP because it integrates multiple possible paths... This would extend the current LCP-based spatial evaluation toward a more biologically realistic representation of multi-route dispersal and provide a natural next step beyond the Bishop et al. and Pless et al. random-forest LCTA framework.  
- Not implemented here because gdistance::commuteDistance() only returns pairwise effective resistance, not current maps, and true current density would require a Circuitscape-style current-flow analysis using the RF-derived resistance surface.
- We would then need to figure out a way to weight the original predictive surfaces in a meaningful way based on the current-flow density for the next iteration of the model.  
- So, for now, I'll use CBER as a screening summary alongside mean LCP and sum LCP. Keep current density as a future extension if the circuit-based approach looks promising.

# Step 1: Inputs
```{r directories-inputs}

# Define Paths to directories
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir_big <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
results_dir <- "../results/"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

######################################
# What surface are you using?
surf_file  <- file.path(results_dir_big, "fullRF_residuals.tif")
########################################

lake_file  <- file.path(data_dir, "processed", "lake_binary.tif")

cse_file   <- file.path("..", "input", "Gff_cse_envCostPaths.csv")

# Read pairwise CSE table
cse_df <- read.csv(cse_file)
# This was added only after completing LOPOCV...
# Filter out western outlier "50-KB" 
cse_df <- cse_df %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")


# Read projected raw CSE surface
cse_surface <- raster(surf_file)

# Read lake mask
lake_mask <- raster(lake_file)
```
  
# Step 2: Check raster alignment and assign high cost to water

```{r check-alignment}
if (!compareRaster(cse_surface, lake_mask,
                   extent = TRUE, rowcol = TRUE,
                   crs = TRUE, res = TRUE,
                   stopiffalse = FALSE)) {
  lake_mask <- projectRaster(lake_mask, cse_surface, method = "ngb")
}

# Inspect projected raw CSE values before altering water cells
summary(values(cse_surface))

# Keep water traversible but very costly (max value from model)
cse_surface[lake_mask[] == 1] <- max(values(cse_surface))

# Check again after assigning lakes
summary(values(cse_surface))

# Quick plot
plot(cse_surface, main = "Projected raw CSE surface with water = high cost")

# Check column names of cse_df
names(cse_df)
```

# Step 3: Prepare pair table, site objects, and transition object  

Diagonal movement is allowed in the 16-way transition matrix, and conductance is corrected for geographic distance, with geoCorrection(type = "c"), so diagonal steps are appropriately penalized relative to orthogonal steps.

```{r prep-transition}
# keep only within-cluster comparisons, matching your earlier workflow
cse_df <- cse_df %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(
    id  = paste(Var1, Var2, sep = "_"),
    CSE = CSEdistance
  )

# unique site coordinates from both columns
sites_coords <- bind_rows(
  cse_df %>% dplyr::select(Site = Var1, lon = long1, lat = lat1),
  cse_df %>% dplyr::select(Site = Var2, lon = long2, lat = lat2)
) %>%
  distinct()

# convert to sf and then to SpatialPoints in raster CRS
sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(cse_surface))

sites_sp <- as(sites_df, "Spatial")

# lookup index for site names
site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

# pair table for evaluation
site_pairs <- cse_df %>%
  dplyr::select(Var1, Var2, id, CSE) %>%
  distinct()

# treat projected raw CSE as local resistance
resistance_rast <- cse_surface

# protect against zero or negative values before inversion
min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
resistance_rast[resistance_rast <= 0] <- min_pos

# convert resistance to conductance for gdistance
conductance_rast <- 1 / resistance_rast

# transition object:
# directions = 16 allows diagonal/extended neighborhood movement
tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)

# correction for least-cost paths
# geoCorrection(type = "c") corrects with longer diagonal
tr_lcp <- geoCorrection(tr, type = "c")

# correction for circuit / random-walk distances
# geoCorrection(type = "r") corrects for random walk
tr_cber <- geoCorrection(tr, type = "r")
```

# Step 4: Sanity checks
```{r sanity}
nrow(cse_df)
nrow(site_pairs)
nrow(sites_coords)

head(site_pairs)

plot(cse_surface, main = "Projected raw CSE surface")
plot(sites_sp, add = TRUE, pch = 16, cex = 0.5)
```  

# Step 5: Compute least-cost paths in parallel (once) and save output  
Test one path first:
```{r test-lcp}
# Test one least-cost path before parallel run
i <- site_index[site_pairs$Var1[100]]
j <- site_index[site_pairs$Var2[100]]

test_path <- shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")

plot(cse_surface, main = "Test least-cost path")
lines(test_path, col = "red", lwd = 2)
points(sites_sp[c(i, j), ], pch = 16)
```  
Run full loop:
```{r save-lcp}
output_dir <- file.path(results_dir_big, "least_cost_paths")
dir.create(output_dir, showWarnings = FALSE)

# Set number of cores and register cluster
n_cores <- 16
cl <- makeCluster(n_cores)
registerDoParallel(cl)

paths_list <- foreach(k = 1:nrow(site_pairs),
                      .packages = c("gdistance", "sp", "sf")) %dopar% {

  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]

  path <- tryCatch({
    shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

stopCluster(cl)

# drop NULLs and combine
paths_list <- paths_list[!sapply(paths_list, is.null)]
paths_sf <- do.call(rbind, paths_list)

# assign CRS and transform back to geographic coordinates for saving
st_crs(paths_sf) <- st_crs(cse_surface)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# join observed CSE values
paths_sf <- left_join(paths_sf, site_pairs, by = "id")

# save paths
st_write(paths_sf,
         file.path(output_dir, "LC_paths_fullRF_residuals.shp"),
         delete_layer = TRUE)
```

# Step 7: Extract predicted CSE values along LCPs and calculate mean and sum
```{r extract-mean-sum}
# if needed later, reload:
paths_sf <- st_read(file.path(results_dir_big, "least_cost_paths", "LC_paths_fullRF_residuals.shp"), quiet = TRUE)

# transform paths to raster CRS for extraction
paths_extract <- st_transform(paths_sf, crs = crs(cse_surface))

# extract raster values along each path
path_vals <- raster::extract(cse_surface, as(paths_extract, "Spatial"))

# summarize values along each least-cost path
lcp_summary <- data.frame(
  id = paths_extract$id,
  LCP_mean = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
  }),
  LCP_sum = sapply(path_vals, function(x) {
    if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
  })
)

# join back to pair table
LCP_results <- site_pairs %>%
  left_join(lcp_summary, by = "id")

head(LCP_results)
summary(LCP_results$LCP_mean)
summary(LCP_results$LCP_sum)

# Save LCP output as text file in results directories
write.csv(LCP_results,file.path(results_dir, "spatial_eval_residuals_LCPpred.csv"), row.names = FALSE)

write.csv(LCP_results,file.path(results_dir_big, "spatial_eval_residuals_LCPpred.csv"), row.names = FALSE)

```  

# Step 8: Compute circuit-based effective resistance (CBER)  
Test one CBER calculation first:
```{r cber-test, eval = TRUE}
# test one CBER calculation first
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])

as.numeric(dmat_test)[1] # returns the single number for pair
```
Run as a loop sequentially, also save output from steps 6-8 to file again...

```{r cber_calculations}
library(doParallel)
library(foreach)
library(gdistance)
library(sp)

# number of cores
n_cores <- 16

# on Linux/CHPC, FORK is usually more efficient for large objects like tr_cber
cl <- parallel::makeForkCluster(n_cores)
doParallel::registerDoParallel(cl)

# optional: quick test outside parallel
i <- site_index[site_pairs$Var1[1]]
j <- site_index[site_pairs$Var2[1]]

dmat_test <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
as.numeric(dmat_test)[1]

# parallel loop across all 1091 pairs
cber_results <- foreach(
  k = seq_len(nrow(site_pairs)),
  .combine = rbind,
  .packages = c("gdistance", "sp")
) %dopar% {
  
  site1 <- site_pairs$Var1[k]
  site2 <- site_pairs$Var2[k]
  pair_id <- site_pairs$id[k]
  
  i <- site_index[site1]
  j <- site_index[site2]
  
  # compute commute distance for the pair
  dmat <- commuteDistance(tr_cber, sites_sp[c(i, j), ])
  
  # extract the single numeric value from the dist object
  cber_val <- as.numeric(dmat)[1]
  
  data.frame(
    Var1 = site1,
    Var2 = site2,
    id   = pair_id,
    CBER = cber_val
  )
}

# stop cluster
stopCluster(cl)

# inspect
head(cber_results)
str(cber_results)

# save output in results directories
write.csv(cber_results,file.path(results_dir, "spatial_eval_residuals_cber.csv"), row.names = FALSE)

write.csv(cber_results,file.path(results_dir_big, "spatial_eval_residuals_cber.csv"), row.names = FALSE)

```

# Step 9: Combine LCP and CBER results
```{r combine}
# load and bind your existing evaluation data
library(dplyr)

# load the LCP sum and mean results you just computed
LCP_results <- read.csv(file.path(results_dir, "spatial_eval_residuals_LCPpred.csv"))

# bind CBER and LCP results
eval_results <- LCP_results %>%
  left_join(
    cber_results %>% select(id, CBER),
    by = "id"
  )

# Save results output as RDS for later ease of loading
saveRDS(
  eval_results,
  file.path(results_dir_big, "spatial_eval_residuals_predictions.rds")
)
```

# Step 10: Calculate evaluation metrics from eval_results
```{r calibrate-and-metrics}
library(dplyr)
library(ggplot2)

# Start eval_df from eval_results
eval_df <- eval_results

# Fit linear calibration and return calibrated predictions
calibrate_pred <- function(obs, pred) {
  keep <- complete.cases(obs, pred)
  df <- data.frame(obs = obs[keep], pred = pred[keep])
  
  fit <- lm(obs ~ pred, data = df)
  
  out <- rep(NA_real_, length(obs))
  out[keep] <- predict(fit, newdata = data.frame(pred = pred[keep]))
  out
}

# Create calibrated predictions
eval_df$LCP_sum_cal  <- calibrate_pred(eval_df$CSE, eval_df$LCP_sum)
eval_df$LCP_mean_cal <- calibrate_pred(eval_df$CSE, eval_df$LCP_mean)
eval_df$CBER_cal     <- calibrate_pred(eval_df$CSE, eval_df$CBER)

# Plots: raw predicted vs observed, with fitted line
plot1 <- ggplot(eval_df, aes(x = LCP_sum, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP sum, raw scale)",
    y = "Observed CSE"
  )
plot1

plot2 <- ggplot(eval_df, aes(x = LCP_mean, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (LCP mean, raw scale)",
    y = "Observed CSE"
  )
plot2

plot3 <- ggplot(eval_df, aes(x = CBER, y = CSE)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Predicted CSE (CBER, raw scale)",
    y = "Observed CSE"
  )
plot3

# Calculate formal metrics
# Correlation is calculated on raw predictions because it is scale-invariant
# Error metrics and R2 are calculated on calibrated predictions
calc_metrics <- function(obs, pred_raw, pred_cal) {
  keep <- complete.cases(obs, pred_raw, pred_cal)
  obs      <- obs[keep]
  pred_raw <- pred_raw[keep]
  pred_cal <- pred_cal[keep]
  
  sse  <- sum((obs - pred_cal)^2)
  sst  <- sum((obs - mean(obs))^2)
  mse  <- mean((obs - pred_cal)^2)
  rmse <- sqrt(mse)
  mae  <- mean(abs(obs - pred_cal))
  rsq  <- 1 - (sse / sst)
  corr <- cor(obs, pred_raw)
  
  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    RSQ = rsq,
    Correlation = corr
  )
}

metrics_df <- bind_rows(
  calc_metrics(eval_df$CSE, eval_df$LCP_mean, eval_df$LCP_mean_cal) %>%
    mutate(method = "LCP_mean"),
  
  calc_metrics(eval_df$CSE, eval_df$LCP_sum, eval_df$LCP_sum_cal) %>%
    mutate(method = "LCP_sum"),
  
  calc_metrics(eval_df$CSE, eval_df$CBER, eval_df$CBER_cal) %>%
    mutate(method = "CBER")
) %>%
  dplyr::select(method, everything())

# Print metrics
print(metrics_df)

# Save outputs
saveRDS(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_residuals_eval_metrics.rds")
)

write.csv(
  metrics_df,
  file.path(results_dir_big, "spatial_eval_residuals_eval_metrics.csv"),
  row.names = FALSE
)

write.csv(
  metrics_df,
  file.path(results_dir, "spatial_eval_residuals_eval_metrics.csv"),
  row.names = FALSE
)
```

# This script should be followed by:

First, 7b-7f. Choose one response formulation and spatial summary.

Then, after selecting the best response formulation and spatial summary:  

Run the final validation framework, full LOPOCV evaluation, including:  
- non-spatial evaluation  
- spatial evaluation (mean LCP, sum LCP, or CBER depending on screening results)  
- Calculate final metrics on held-out data (RMSE, R², MSE, MAE, correlation)

Permutation test for inference  
- Run ~100 permutations where the response is randomized.

For each permutation:  
- rerun the full LOPOCV workflow  
- recompute spatial predictions  
- calculate the same evaluation metrics  
Compare observed metrics (especially RMSE and R²) to the permutation distribution to test whether predictive performance is better than expected by chance.