---
title: "LOPOCV: Random Forest (Internal) and Spatial Cross-Validation"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../scripts_knitted_md", 
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/knitted_mds/" 
)
```
  
# Setup  

RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 16 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-np (allows multiple simultaneous jobs automatically now)  
- **Memory per job:** 400G (cluster limit: 1000G total; avoid exceeding half)  

## Overview of script
Although model configuration and projection parameters were selected based on full-model performance (RSQ), during LOPOCV, some held-out folds exhibited limited variance in observed CSE, making fold-specific RSQ unstable. We therefore report fold-level performance using Spearman and Pearson correlation alongside RMSE and MAE, which provide more robust summaries under these conditions. 

Emphasis should be on Spearman's Correlation because it weighs correct ranking more heavily than Pearsons, as it is calculated based on the rank order of data points rather than their raw numerical values.

### Notes from Jun 20, 2025

My notes from my analysis on Model Design and Diagnostic Attempts Jun 20, 2025, when using RSQ per fold:

I tried removing sampling effort as a predictor:\
- Result: Worsened performance in west, no benefit in north/south.

I tried subdividing by subclusters (K=3):\
- Result: Lower R² overall due to smaller training sets.

I tried filtering by geographic distance:\
- Result: Worsened performance due to fewer training pairs.

I tested exclusion of site 50-KB:\
- Result: **Improved** performance in west.

I also tested removing site 54-MS:\
- Result: No benefit, and it actually reduced accuracy in neighboring KAF.

Final decision:\
- Stick with K=2 major clusters (north / south).\
- Keep all intra-cluster pairs (no geodist pruning).\
- Just exclude 50-KB as a clear spatial outlier.\
- Use subclusters only for interpretation, not model restriction.


(In contrast, model configuration and projection parameters were selected based on full-model performance (RSQ) becaues in these cases, variance in CSE was sufficiently large for stable estimation.)

## Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../results_dir/fullRF_CSE_resistance.tif` - Final full model projected resistance surface
  - `../results_dir/LC_paths_fullRF.shp"` - 
  - `../data_dir/processed/env_stack.grd`  - Final prediction env stack with named layers (18 variables)
env <- stack(file.path(
  - `../results_dir/lopocv/rf_model_01.rds` - 67 LOPOCV rf models leaving one point out
  
## Outputs  
  -   `../results_dir/lopocv/` - rf_model_01.rds to rf_model_67.rds for each fold of LOPOCV  
  - `../results_dir/lopocv_spatial_LCPsum_k/` - Spatial lopocv predictions and metrics for each fold 1-67
  - `../results_dir/spatial_LOPOCV_LCPsum_k1_summary.csv` - Spatial lopocv evaluation metrics  
  
  
## Libraries
```{r libraries, warning=FALSE, results=FALSE, message=FALSE, eval = TRUE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(randomForest)
library(readr)
library(ggplot2)
library(sp)
library(tidyverse)
library(rnaturalearth)
library(rnaturalearthdata)
library(reshape2)
```

## Directory Paths

### LOPOCV directories
```{r general-dir-set-up, eval = TRUE}
input_dir <- "../input"
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

# define ggplot2 extent
xlim <- c(28.6, 35.4)
ylim <- c(-1.500000 , 4.733333)

# Input: shape file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo

# Build list of unique sites from Var1 and Var2
sites <- sort(unique(c(lcp_sf$Var1, lcp_sf$Var2)))

# Main LOPOCV output directory
output_dir <- file.path(results_dir, "lopocv")
#dir.create(output_dir, showWarnings = FALSE)

# Spatial output directory
spatial_output_dir <- file.path(results_dir, "lopocv_spatial_LCPsum_k1")

if (!dir.exists(spatial_output_dir)) {
  dir.create(spatial_output_dir, recursive = TRUE)
}
```
# LOPOCV

## Chunk 1: Set up folds of the cross validation
```{r 1-fold-files, eval=TRUE}
# Load data
V.table_full <- read.csv(file.path(input_dir, "Gff_cse_envCostPaths.csv"))

# estimate mean sampling density
mean(V.table_full$samp_20km_mean, na.rm = TRUE)

# Filter out western outlier "50-KB" 
V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# Filter for within-cluster pairs AND geographic distance ≤ 100 km
#V.table <- V.table_full %>%
#  filter(Pop1_cluster == Pop2_cluster) %>%
#  filter(pix_dist <= 100)

# Create unique ID after filtering
V.table$id <- paste(V.table$Var1, V.table$Var2, sep = "_")

# Define site list
sites <- sort(unique(c(V.table$Var1, V.table$Var2)))

# How many rows of data for each?
table(V.table$Pop1_cluster)

# How many unique sites?
length(sites)

# Choose predictors for RF model (adjust names if necessary)
predictor_vars <- c("BIO1_mean","BIO2_mean","BIO3_mean","BIO4_mean", "BIO5_mean","BIO6_mean","BIO7_mean", "BIO8S_mean", "BIO9S_mean","BIO10S_mean", "BIO11S_mean","BIO12_mean", "BIO13_mean","BIO14_mean","BIO15_mean","BIO16S_mean","BIO17S_mean", "BIO18S_mean","BIO19S_mean","slope_mean","alt_mean", "lakes_mean","riv_3km_mean", "samp_20km_mean","pix_dist")

# Filter to modeling-relevant columns only
V.model <- V.table[, c("CSEdistance", predictor_vars)]
```
  
## Chunk 2: Test LOPOCV with first fold

```{r 2-lopocv-test}
# Pick a random site for the test fold
set.seed(1298373)
i <- sample(length(sites), 1)
site <- sites[i]
cat(sprintf("Testing fold %02d with site = %s\n", i, site))

# Split into test and train rows
test_rows  <- V.table %>% filter(Var1 == site | Var2 == site)
train_rows <- V.table %>% filter(!(Var1 == site | Var2 == site))

# Add row indices for matching with V.model
test_idx  <- which(V.table$id %in% test_rows$id)
train_idx <- which(V.table$id %in% train_rows$id)

# Create modeling input frames
train_df <- V.model[train_idx, ]
test_df  <- V.model[test_idx, ]

# Train RF model (no tuning)
set.seed(42)
rf_model <- randomForest(
  CSEdistance ~ .,
  data = train_df,
  ntree = 500,
  importance = TRUE
)

# Predict and calculate metrics
pred_train <- predict(rf_model, newdata = train_df)
pred_test  <- predict(rf_model, newdata = test_df)

rsq   <- tail(rf_model$rsq, 1)
rmse  <- sqrt(mean((pred_test - test_df$CSEdistance)^2))
mae   <- mean(abs(pred_test - test_df$CSEdistance))
cor1  <- cor(pred_train, train_df$CSEdistance)
cor2  <- cor(pred_test, test_df$CSEdistance)

varImpPlot(rf_model)

cat(sprintf("Fold %02d completed.\n", i))
```
  
## Chunk 3: RF models for each fold of LOPOCV

**NOTE:** eval = FALSE so that it skips on knit
```{r 3-lopocv, eval = FALSE}
# Parallel setup
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Run LOPOCV with foreach
foreach(
  i = seq_along(sites),
  .packages = c("dplyr", "randomForest")
) %dopar% {

  # Identify current test site
  site <- sites[i]

  # Identify rows where site appears as Var1 or Var2 (test set)
  test_rows <- V.table %>% filter(Var1 == site | Var2 == site)

  # All other rows go into the training set
  train_rows <- V.table %>% filter(!(Var1 == site | Var2 == site))

  # Use precomputed pair IDs to match rows in V.model
  test_idx  <- which(V.table$id %in% test_rows$id)
  train_idx <- which(V.table$id %in% train_rows$id)

  # Subset predictor data to match training/testing rows
  train_df <- V.model[train_idx, ]
  test_df  <- V.model[test_idx, ]

  # Train Random Forest model (fixed ntree and default mtry)
  set.seed(42)
  rf_model <- randomForest(
    CSEdistance ~ .,
    data = train_df,
    ntree = 500,
    importance = TRUE
  )

  # Save Random Forest model for this fold
  saveRDS(rf_model, file.path(output_dir, sprintf("rf_model_%02d.rds", i)))

  NULL
}

stopCluster(cl)
```

## Chunk 4: Calculate metrics
This chunk:  
 - Helpers for LOPOCV Evaluative metrics
 - computes all metrics
 - Spearman replaces RSQ
 - computes pooled R² and other metrics (new improvement)
 - Plots and summarizes metrics

Helpers:
```{r 4-non-spatial-helpers, eval = TRUE}
calibrate_pred <- function(obs_train, pred_train, pred_test) {
  keep <- complete.cases(obs_train, pred_train)
  df <- data.frame(obs = obs_train[keep], pred = pred_train[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(pred_test))
  keep_test <- complete.cases(pred_test)
  out[keep_test] <- predict(fit, newdata = data.frame(pred = pred_test[keep_test]))
  out
}

calc_metrics <- function(obs, pred) {
  keep <- complete.cases(obs, pred)

  obs  <- obs[keep]
  pred <- pred[keep]

  mse      <- mean((obs - pred)^2)
  rmse     <- sqrt(mse)
  mae      <- mean(abs(obs - pred))
  pearson  <- cor(obs, pred, method = "pearson")
  spearman <- cor(obs, pred, method = "spearman")

  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    Pearson = pearson,
    Spearman = spearman
  )
}
```
Run in loop
```{r 4-lopocv-metrics-non-spatial, eval = FALSE}
# -----------------------------------
# Step 1: generate fold-level predictions from saved .rds models
# -----------------------------------
for (i in seq_along(sites)) {

  site <- sites[i]

  fold_eval_file  <- file.path(results_dir, sprintf("lopocv_eval_fold_%02d.csv", i))
  fold_model_file <- file.path(output_dir, sprintf("rf_model_%02d.rds", i))

  if (!file.exists(fold_model_file)) {
    warning(sprintf("Missing model file: %s", fold_model_file))
    next
  }

  test_rows  <- V.table %>% filter(Var1 == site | Var2 == site)
  train_rows <- V.table %>% filter(!(Var1 == site | Var2 == site))

  test_idx  <- which(V.table$id %in% test_rows$id)
  train_idx <- which(V.table$id %in% train_rows$id)

  train_df <- V.model[train_idx, ]
  test_df  <- V.model[test_idx, ]

  rf_model <- readRDS(fold_model_file)

  train_obs <- train_df$CSEdistance
  test_obs  <- test_df$CSEdistance

  train_df <- train_df %>% dplyr::select(-CSEdistance)
  test_df  <- test_df %>% dplyr::select(-CSEdistance)

  names(train_df) <- paste0(names(train_df), "_mean")
  names(test_df)  <- paste0(names(test_df), "_mean")

  names(train_df)[names(train_df) == "pix_dist_mean"] <- "pix_dist"
  names(test_df)[names(test_df) == "pix_dist_mean"] <- "pix_dist"

  stopifnot(identical(names(train_df), attr(rf_model$terms, "term.labels")))
  stopifnot(identical(names(test_df), attr(rf_model$terms, "term.labels")))

  pred_train <- predict(rf_model, newdata = train_df)
  pred_test  <- predict(rf_model, newdata = test_df)

  pred_test_cal <- calibrate_pred(
    obs_train = train_obs,
    pred_train = pred_train,
    pred_test = pred_test
  )

  eval_df <- data.frame(
    Var1 = test_rows$Var1,
    Var2 = test_rows$Var2,
    id = test_rows$id,
    CSE = test_obs,
    pred_raw = pred_test,
    pred_cal = pred_test_cal
  )

  write.csv(eval_df, fold_eval_file, row.names = FALSE)
}

# -----------------------------------
# Step 2: calculate fold-level metrics
# -----------------------------------
for (i in seq_along(sites)) {

  site <- sites[i]

  fold_eval_file    <- file.path(results_dir, sprintf("lopocv_eval_fold_%02d.csv", i))
  fold_metrics_file <- file.path(results_dir, sprintf("metrics_fold_%02d.csv", i))
  fold_model_file   <- file.path(output_dir, sprintf("rf_model_%02d.rds", i))

  if (!file.exists(fold_eval_file)) {
    warning(sprintf("Missing eval file: %s", fold_eval_file))
    next
  }

  if (!file.exists(fold_model_file)) {
    warning(sprintf("Missing model file: %s", fold_model_file))
    next
  }

  df <- read.csv(fold_eval_file)
  rf_model <- readRDS(fold_model_file)

  metrics_k <- calc_metrics(
    obs = df$CSE,
    pred = df$pred_cal
  ) %>%
    mutate(site = site) %>%
    dplyr::select(site, everything())

  imp <- as.data.frame(importance(rf_model))
  imp_mse <- setNames(as.list(imp[, "%IncMSE"]), paste0("IncMSE_", rownames(imp)))
  imp_purity <- setNames(as.list(imp[, "IncNodePurity"]), paste0("NodePurity_", rownames(imp)))

  metrics_k <- bind_cols(
    metrics_k,
    as.data.frame(imp_mse),
    as.data.frame(imp_purity)
  )

  write.csv(metrics_k, fold_metrics_file, row.names = FALSE)
}

# -----------------------------------
# Step 3: combine fold metrics
# -----------------------------------
metric_files <- list.files(
  results_dir,
  pattern = "^metrics_fold_[0-9]{2}\\.csv$",
  full.names = TRUE
)

metrics_all <- bind_rows(lapply(metric_files, read.csv)) %>%
  mutate(site = factor(site, levels = sites)) %>%
  arrange(site)

write.csv(metrics_all, file.path(results_dir, "LOPOCV_summary.csv"), row.names = FALSE)

print(metrics_all)

# -----------------------------------
# Step 4: pooled metrics across all folds
# -----------------------------------
eval_files <- list.files(
  results_dir,
  pattern = "^lopocv_eval_fold_[0-9]{2}\\.csv$",
  full.names = TRUE
)

eval_all <- bind_rows(lapply(eval_files, read.csv))

keep_all <- complete.cases(eval_all$CSE, eval_all$pred_cal)

obs_all  <- eval_all$CSE[keep_all]
pred_all <- eval_all$pred_cal[keep_all]

R2_cv <- 1 - sum((obs_all - pred_all)^2) / sum((obs_all - mean(obs_all))^2)

pooled_summary <- data.frame(
  n = length(obs_all),
  pooled_R2 = R2_cv,
  pooled_Pearson = cor(obs_all, pred_all, method = "pearson"),
  pooled_Spearman = cor(obs_all, pred_all, method = "spearman"),
  pooled_RMSE = sqrt(mean((obs_all - pred_all)^2)),
  pooled_MAE = mean(abs(obs_all - pred_all))
)

write.csv(pooled_summary, file.path(results_dir, "LOPOCV_pooled_summary.csv"), row.names = FALSE)

print(pooled_summary)
```
## Chunk 5: Visualize Pooled and Fold-level Metrics (Spearman instead of RSQ)
```{r 5-viz-spearman, warning = FALSE, eval = TRUE}
pooled_summary <- read.csv(file.path(results_dir, "LOPOCV_pooled_summary.csv"))

print(pooled_summary)

metrics_all <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))
max(metrics_all$Spearman)
min(metrics_all$Spearman)

altitude <- raster::raster(file.path(
  "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data/processed",
  "altitude_1KMmedian_MERIT_UgandaClip.tif"
))
crs(altitude) <- 4326

indinfo <- read.delim("../input/Gff_11loci_allsites_indinfo.txt")
site_clusters <- indinfo %>%
  dplyr::select(Site = SiteCode, Subcluster = SiteMajCluster) %>%
  distinct()

site_metadata <- V.table %>%
  dplyr::select(Site = Var1, Latitude = lat1, Longitude = long1) %>%
  distinct() %>%
  left_join(site_clusters, by = "Site") %>%
  left_join(metrics_all, by = c("Site" = "site")) %>%
  mutate(Symbol = ifelse(Spearman < 0.3, "low", "circle")) %>%
  arrange(desc(Spearman))

r_ext <- extent(altitude)
xlim <- c(r_ext@xmin, r_ext@xmax)
ylim <- c(r_ext@ymin, r_ext@ymax)

uganda <- ne_countries(scale = "medium", continent = "Africa", returnclass = "sf") %>% st_transform(4326)
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>% st_transform(4326)

ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.1) +
  geom_sf(data = lakes, fill = "black", color = NA) +
  geom_point(
    data = filter(site_metadata, Symbol == "circle"),
    aes(x = Longitude, y = Latitude, size = Spearman, fill = Subcluster),
    shape = 21, color = "black", stroke = 0.3
  ) +
  geom_point(
    data = filter(site_metadata, Symbol == "low"),
    aes(x = Longitude, y = Latitude, fill = Subcluster),
    shape = 21, color = "black", stroke = 0.3, size = 1
  ) +
  scale_fill_manual(
    name = "Subcluster",
    values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")
  ) +
  scale_size_continuous(
  name = "LOPOCV \nSpearman's r",
  limits = c(0.4, 1.0),
  breaks = c(0.4, 0.7, 1.0),
  range = c(2,7)
) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "LOPOCV Spearman's r by Sampling Site", x = "Longitude", y = "Latitude")

ggplot(site_metadata, aes(x = Spearman, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(
    title = "Distribution of Spearman's r by Subcluster",
    x = "Spearman's r (LOPOCV)",
    y = "Density"
  )

ggplot(site_metadata, aes(x = Spearman, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = after_stat(count))) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(
    title = "Spearman's r by Subcluster (Scaled by Count)",
    x = "Spearman's r (LOPOCV)",
    y = "Count"
  )

site_metadata$Cluster <- site_metadata$Subcluster
site_metadata$Cluster[site_metadata$Subcluster == "west"] <- "south"

ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.1) +
  geom_sf(data = lakes, fill = "black", color = NA) +
  geom_point(
    data = filter(site_metadata, Symbol == "circle"),
    aes(x = Longitude, y = Latitude, size = Spearman, fill = Cluster),
    shape = 21, color = "black", stroke = 0.3, alpha = 0.4
  ) +
  geom_point(
    data = filter(site_metadata, Symbol == "low"),
    aes(x = Longitude, y = Latitude, fill = Cluster),
    shape = 21, color = "black", size = 1, stroke = 0.3, alpha = 0.3
  ) +
  scale_fill_manual(
    name = "Group",
    values = c("north" = "#1f78b4", "south" = "#e66101")
  ) +
  scale_size_continuous(
  name = "LOPOCV \nSpearman's r",
  limits = c(0.4, 1.0),
  breaks = c(0.4, 0.7, 1.0),
  range = c(2,7)
) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "LOPOCV Spearman's r by Site", x = "Longitude", y = "Latitude")

ggplot(site_metadata, aes(x = Spearman, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(
    title = "Distribution of Spearman's r by Cluster",
    x = "Spearman's r (LOPOCV)",
    y = "Density"
  )

ggplot(site_metadata, aes(x = Spearman, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = after_stat(count))) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(
    title = "Spearman's r by Cluster (Scaled by Count)",
    x = "Spearman's r (LOPOCV)",
    y = "Count"
  )

# write.csv(site_metadata, file.path(results_dir, "site_metadata.csv"), row.names = FALSE)
```


## Chunk 6: Plot Full Model: Predicted vs Observed
```{r 6-plot-R2, eval = TRUE}
# Load full RF model trained on V.model (non-spatial)
rf_model_raw <- readRDS(file.path(results_dir, "rf_mean_full_tuned.rds"))
rownames(rf_model_raw$importance) # predictor variables used in final model

# Filter to modeling-relevant columns only
predictor_vars <- c("BIO1_mean","BIO2_mean","BIO3_mean","BIO4_mean", "BIO5_mean","BIO6_mean","BIO7_mean", "BIO8S_mean", "BIO9S_mean","BIO10S_mean", "BIO11S_mean","BIO12_mean", "BIO13_mean","BIO14_mean","BIO15_mean","BIO16S_mean","BIO17S_mean", "BIO18S_mean","BIO19S_mean","slope_mean","alt_mean", "lakes_mean","riv_3km_mean", "samp_20km_mean","pix_dist")

V.model <- V.table[, c("CSEdistance", predictor_vars)]

# Rename predictors by removing "_mean" for later projections
names(V.model) <- gsub("_mean$", "", names(V.model))

# Predict on full data
pred_all <- predict(rf_model_raw, newdata = V.model[,-1])
obs_all <- V.model[,1]
# Plot predicted vs observed
ggplot(data.frame(obs = obs_all, pred = pred_all), aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  labs(x = "Observed CSE", y = "(No Model Projection) \n Predicted CSE")
```
  
## Chunk 7: Variable importance with  pca-pruned env. variable importance

### %IncMSE
```{r 7-variable-imp-mse-plus-pca-pruned, eval = TRUE}
# Load full model
full_model <- readRDS(file.path(results_dir, "rf_mean_full_tuned.rds"))

print(full_model)
importance(full_model)
varImpPlot(full_model)

full_imp <- importance(full_model, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(IncMSE = `%IncMSE`) %>%
  mutate(model = "full")

# Load LOPOCV models
lopocv_imp <- map_df(1:67, function(fold_idx) {
  rf_model <- readRDS(sprintf("%s/lopocv/rf_model_%02d.rds", results_dir, fold_idx))
  imp <- importance(rf_model, type = 1)
  data.frame(
    variable = rownames(imp),
    IncMSE = imp[,1],
    model = paste0("fold_", fold_idx)
  )
}) %>%
  mutate(variable = str_replace(variable, "_mean$", ""))


# Load PCA-pruned model
pruned_model <- readRDS(file.path(results_dir, "rf_mean_pcapruned_tuned.rds"))

print(pruned_model)
importance(pruned_model)
varImpPlot(pruned_model)

pruned_imp <- importance(pruned_model, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(IncMSE = `%IncMSE`) %>%
  mutate(
    variable = str_replace(variable, "_mean$", ""),
    model = "pruned"
  )

# Combine
all_imp <- bind_rows(full_imp, lopocv_imp, pruned_imp)

# Define custom labels
label_map <- c(
  BIO1   = "Annual Mean Temperature (BIO1)",
  BIO2   = "Mean Diurnal Temp Range (BIO2)",
  BIO3   = "Isothermality (BIO3)",
  BIO4   = "Temperature Seasonality (BIO4)",
  BIO5   = "Max Temp of Warmest Month (BIO5)",
  BIO6   = "Min Temp of Coldest Month (BIO6)",
  BIO7   = "Temperature Annual Range (BIO7)",
  BIO8S  = "Mean Temp of Wettest Season (BIO8S)",
  BIO9S  = "Mean Temp of Driest Season (BIO9S)",
  BIO10S = "Mean Temp of Warmest Season (BIO10S)",
  BIO11S = "Mean Temp of Coldest Season (BIO11S)",
  BIO12  = "Annual Precipitation (BIO12)",
  BIO13  = "Precipitation of Wettest Month (BIO13)",
  BIO14  = "Precipitation of Driest Month (BIO14)",
  BIO15  = "Precipitation Seasonality (BIO15)",
  BIO16S = "Precipitation of Wettest Season (BIO16S)",
  BIO17S = "Precipitation of Driest Season (BIO17S)",
  BIO18S = "Precipitation of Warmest Season (BIO18S)",
  BIO19S = "Precipitation of Coldest Season (BIO19S)",
  slope  = "Slope",
  alt    = "Altitude",
  lakes  = "Lake Presence/Absence",
  riv_3km= "River Kernel Density (3 km bandwidth)",
  samp_20km = "Sampling Density (20 km bandwidth)",
  pix_dist    = "Geographic Distance"
)

# Define abreviated custom labels (NOT USED IN CURRENT SCRIPT)
label_map_abrev <- c(
  BIO1   = "BIO1",
  BIO2   = "BIO2",
  BIO3   = "BIO3",
  BIO4   = "BIO4",
  BIO5   = "BIO5",
  BIO6   = "BIO6",
  BIO7   = "BIO7",
  BIO8S  = "BIO8S",
  BIO9S  = "BIO9S",
  BIO10S = "BIO10S",
  BIO11S = "BIO11S",
  BIO12  = "BIO12",
  BIO13  = "BIO13",
  BIO14  = "BIO14",
  BIO15  = "BIO15",
  BIO16S = "BIO16S",
  BIO17S = "BIO17S",
  BIO18S = "BIO18S",
  BIO19S = "BIO19S",
  slope  = "Slope",
  alt    = "Altitude",
  lakes  = "Lakes",
  riv_3km= "Rivers",
  samp_20km = "Sampling",
  pix_dist    = "Geo. Dist."
)
# Order by full model's %IncMSE (top to bottom)
full_order <- all_imp %>%
  filter(model == "full") %>%
  arrange(desc(IncMSE)) %>%
  pull(variable)

all_imp$variable <- factor(all_imp$variable, levels = rev(full_order))

# Plot
#pdf("../figures/VarImpPlot.pdf",width =6, height=6)
ggplot(all_imp, aes(x = variable, y = IncMSE)) +
  geom_point(data = filter(all_imp, model != "full" & model != "pruned"),
             aes(group = model),
             color = "grey70", alpha = 0.4, size = 2) +
  geom_point(data = filter(all_imp, model == "full"),
             color = "black", size = 3) +
  geom_point(data = filter(all_imp, model == "pruned"),
             shape = 17, size = 3.5, color = "grey70") +
  coord_flip() +
  scale_y_continuous(name = "%IncMSE") +
  scale_x_discrete(labels = label_map) +
  labs(x = NULL,
       title = "Variable Importance",
       subtitle = "Triangles show model with PCA-pruned env. vars. only") +
  theme_minimal()
#dev.off()
```
### NodePurity
```{r 7-variable-imp-nodepurity-plus-pca-pruned, eval = TRUE}
# Load full model
full_model <- readRDS(file.path(results_dir, "rf_mean_full_tuned.rds"))
full_imp <- importance(full_model, type = 2) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(NodePurity = IncNodePurity) %>%
  mutate(model = "full")

# Load LOPOCV models
lopocv_imp <- map_df(1:67, function(fold_idx) {
  rf_model <- readRDS(sprintf("%s/lopocv/rf_model_%02d.rds", results_dir, fold_idx))
  imp <- importance(rf_model, type = 2)
  data.frame(
    variable = rownames(imp),
    NodePurity = imp[,1],
    model = paste0("fold_", fold_idx)
  )
}) %>%
  mutate(variable = str_replace(variable, "_mean$", ""))


# Load PCA-pruned model
pruned_model <- readRDS(file.path(results_dir, "rf_mean_pcapruned_tuned.rds"))
pruned_imp <- importance(pruned_model, type = 2) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(NodePurity = IncNodePurity) %>%
  mutate(
    variable = str_replace(variable, "_mean$", ""),
    model = "pruned"
  )

# Combine
all_imp <- bind_rows(full_imp, lopocv_imp, pruned_imp)

# Define custom labels
label_map <- c(
  BIO1   = "Annual Mean Temperature (BIO1)",
  BIO2   = "Mean Diurnal Temp Range (BIO2)",
  BIO3   = "Isothermality (BIO3)",
  BIO4   = "Temperature Seasonality (BIO4)",
  BIO5   = "Max Temp of Warmest Month (BIO5)",
  BIO6   = "Min Temp of Coldest Month (BIO6)",
  BIO7   = "Temperature Annual Range (BIO7)",
  BIO8S  = "Mean Temp of Wettest Season (BIO8S)",
  BIO9S  = "Mean Temp of Driest Season (BIO9S)",
  BIO10S = "Mean Temp of Warmest Season (BIO10S)",
  BIO11S = "Mean Temp of Coldest Season (BIO11S)",
  BIO12  = "Annual Precipitation (BIO12)",
  BIO13  = "Precipitation of Wettest Month (BIO13)",
  BIO14  = "Precipitation of Driest Month (BIO14)",
  BIO15  = "Precipitation Seasonality (BIO15)",
  BIO16S = "Precipitation of Wettest Season (BIO16S)",
  BIO17S = "Precipitation of Driest Season (BIO17S)",
  BIO18S = "Precipitation of Warmest Season (BIO18S)",
  BIO19S = "Precipitation of Coldest Season (BIO19S)",
  slope  = "Slope",
  alt    = "Altitude",
  lakes  = "Lake Presence/Absence",
  riv_3km= "River Kernel Density (3 km bandwidth)",
  samp_20km = "Sampling Density (20 km bandwidth)",
  pix_dist    = "Geographic Distance"
)

# Order by full model's node purity (top to bottom)
full_order <- all_imp %>%
  filter(model == "full") %>%
  arrange(desc(NodePurity)) %>%
  pull(variable)

all_imp$variable <- factor(all_imp$variable, levels = rev(full_order))

# Plot
ggplot(all_imp, aes(x = variable, y = NodePurity)) +
  geom_point(data = filter(all_imp, model != "full" & model != "pruned"),
             aes(group = model),
             color = "grey70", alpha = 0.4, size = 2) +
  geom_point(data = filter(all_imp, model == "full"),
             color = "black", size = 3) +
  geom_point(data = filter(all_imp, model == "pruned"),
             shape = 17, size = 3.5, color = "grey70") +
  coord_flip() +
  scale_y_continuous(name = "Node Purity") +
  scale_x_discrete(labels = label_map) +
  labs(x = NULL,
       title = "Variable Importance",
       subtitle = "Triangles show model with PCA-pruned env. vars. only") +
  theme_minimal()
```

# Spatial LOPOCV: 

## Chunk 8: Spatial eval. on existing LOPOCV

Here, we keep internal LOPOCV as already completed
reuse saved results/lopocv/rf_model_01.rds to rf_model_67.rds but change spatial evaluation to:  
 - LCP_sum
 - lakes high cost (maximum predicted CSE for the model)
 - calibration on training pairs only
 - kdist = 1km , this was the best choice for the full model
 
What this chunk does:
- loads saved `rf_model_XX.rds` from `results/lopocv`
- projects a resistance surface with `pix_dist = 1`
- forces lakes to high cost
- recomputes LCPs
- summarizes `LCP_sum`
- calibrates on training pairs only
- evaluates on the held-out site pairs only
- saves per-fold outputs

Helpers
```{r 8a-spatial-helpers, eval = TRUE}
calibrate_pred <- function(obs_train, pred_train, pred_test) {
  keep <- complete.cases(obs_train, pred_train)
  df <- data.frame(obs = obs_train[keep], pred = pred_train[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(pred_test))
  keep_test <- complete.cases(pred_test)
  out[keep_test] <- predict(fit, newdata = data.frame(pred = pred_test[keep_test]))
  out
}

calc_metrics <- function(obs, pred) {
  keep <- complete.cases(obs, pred)

  obs  <- obs[keep]
  pred <- pred[keep]

  mse      <- mean((obs - pred)^2)
  rmse     <- sqrt(mse)
  mae      <- mean(abs(obs - pred))
  pearson  <- cor(obs, pred, method = "pearson")
  spearman <- cor(obs, pred, method = "spearman")

  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    Pearson = pearson,
    Spearman = spearman
  )
}
```

Least-cost-paths: This recomputes paths from the projected resistance surface each fold.
```{r 8b-spatial-inputs, eval = TRUE}
# Load environmental raster stack
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))
names(env) <- c(
  "BIO1_mean", "BIO2_mean", "BIO3_mean", "BIO4_mean", "BIO5_mean", "BIO6_mean",
  "BIO7_mean", "BIO8S_mean", "BIO9S_mean", "BIO10S_mean", "BIO11S_mean",
  "BIO12_mean", "BIO13_mean", "BIO14_mean", "BIO15_mean", "BIO16S_mean",
  "BIO17S_mean", "BIO18S_mean", "BIO19S_mean", "alt_mean", "slope_mean",
  "riv_3km_mean", "samp_20km_mean", "lakes_mean", "pix_dist"
)
crs(env) <- crs_geo

# Build lake mask from lakes layer
lake_mask <- raster(env[["lakes_mean"]])
lake_mask[lake_mask > 0] <- 1

# Load pair table
V.table_full <- read.csv(file.path(input_dir, "Gff_cse_envCostPaths.csv"))

V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB") %>%
  mutate(id = paste(Var1, Var2, sep = "_"))

# Pair table used in spatial evaluation
site_pairs <- V.table %>%
  dplyr::select(id, Var1, Var2, CSE = CSEdistance) %>%
  distinct()

# FORCE same ordering as LOPOCV models
sites <- sort(unique(c(lcp_sf$Var1, lcp_sf$Var2)))

# Site coordinates
sites_coords_1 <- V.table %>%
  dplyr::select(Site = Var1, lon = long1, lat = lat1) %>%
  distinct()

sites_coords_2 <- V.table %>%
  dplyr::select(Site = Var2, lon = long2, lat = lat2) %>%
  distinct()

sites_coords <- bind_rows(sites_coords_1, sites_coords_2) %>%
  distinct(Site, .keep_all = TRUE) %>%
  arrange(Site)

#check and stop if sites not represented in order
stopifnot(all(sites %in% sites_coords$Site))

# fixed projection distance
kdist <- 1
```

Run in loop for sum of predicted CSE along LCP, and calibration
```{r 8c-spatial-lopocv-LCPsum-k1, eval = FALSE}
# Parallel setup
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

for (fold_idx in seq_along(sites)) {

  site <- sites[fold_idx]

  # existing saved LOPOCV model
  fold_model_file <- file.path(output_dir, sprintf("rf_model_%02d.rds", fold_idx))

  # new spatial outputs
  fold_eval_file <- file.path(spatial_output_dir, sprintf("spatial_eval_fold_%02d.csv", fold_idx))

  if (!file.exists(fold_model_file)) {
    warning(sprintf("Missing saved LOPOCV model: %s", fold_model_file))
    next
  }

  # skip completed folds
  if (file.exists(fold_eval_file) && file.exists(fold_metrics_file)) {
    next
  }

  rf_model <- readRDS(fold_model_file)

  # -----------------------------------
  # Step 1: project surface at kdist = 1
  # -----------------------------------
  env_k <- env
  env_k[["pix_dist"]] <- setValues(env[[1]], kdist)
  
  # force raster names to match the saved RF model   exactly
  rf_vars <- attr(rf_model$terms, "term.labels")
  names(env_k) <- rf_vars
  
  # sanity check
  stopifnot(identical(names(env_k), rf_vars))
  
  cse_surface <- predict(env_k, rf_model, type = "response")

  # align lake mask if needed
  if (!compareRaster(cse_surface, lake_mask,
                     extent = TRUE, rowcol = TRUE,
                     crs = TRUE, res = TRUE,
                     stopiffalse = FALSE)) {
    lake_mask_use <- projectRaster(lake_mask, cse_surface, method = "ngb")
  } else {
    lake_mask_use <- lake_mask
  }

  # assign lakes high cost
  cse_surface[lake_mask_use[] == 1] <- max(values(cse_surface), na.rm = TRUE)

  # -----------------------------------
  # Step 2: prepare sites in raster CRS
  # -----------------------------------
  sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
    st_transform(crs(cse_surface))

  sites_sp <- as(sites_df, "Spatial")
  site_index <- setNames(seq_len(nrow(sites_df)), sites_coords$Site)

  # -----------------------------------
  # Step 3: build transition object
  # -----------------------------------
  resistance_rast <- cse_surface
  min_pos <- min(values(resistance_rast)[values(resistance_rast) > 0], na.rm = TRUE)
  resistance_rast[resistance_rast <= 0] <- min_pos

  conductance_rast <- 1 / resistance_rast
  tr <- transition(conductance_rast, transitionFunction = mean, directions = 16)
  tr_lcp <- geoCorrection(tr, type = "c")

  # -----------------------------------
  # Step 4: recompute least-cost paths
  # -----------------------------------
  paths_list <- foreach(
    ii = seq_len(nrow(site_pairs)),
    .packages = c("gdistance", "sp", "sf")
  ) %dopar% {
    i <- site_index[site_pairs$Var1[ii]]
    j <- site_index[site_pairs$Var2[ii]]
  
    path <- tryCatch({
      shortestPath(tr_lcp, sites_sp[i, ], sites_sp[j, ], output =   "SpatialLines")
    }, error = function(e) NULL)
  
    if (!is.null(path)) {
      path_sf <- st_as_sf(path)
      path_sf$id <- site_pairs$id[ii]
      path_sf
    } else {
      NULL
    }
  }

  paths_list <- paths_list[!sapply(paths_list, is.null)]

  if (length(paths_list) == 0) {
    warning(sprintf("No paths recovered for fold %02d", fold_idx))
    next
  }
  
  paths_sf <- do.call(rbind, paths_list)
  st_crs(paths_sf) <- st_crs(cse_surface)

  # extract raster values along each path
  path_vals <- raster::extract(cse_surface, as(paths_sf, "Spatial"))

  lcp_summary <- data.frame(
    id = paths_sf$id,
    LCP_sum = sapply(path_vals, function(x) {
      if (is.null(x) || all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
    })
  )

  # -----------------------------------
  # Step 5: combine and calibrate
  # -----------------------------------
  eval_results <- site_pairs %>%
    left_join(lcp_summary, by = "id")

  test_ids <- eval_results$id[eval_results$Var1 == site | eval_results$Var2 == site]
  train_df <- eval_results[!eval_results$id %in% test_ids, ]
  test_df  <- eval_results[eval_results$id %in% test_ids, ]

  test_df$LCP_sum_cal <- calibrate_pred(
    obs_train = train_df$CSE,
    pred_train = train_df$LCP_sum,
    pred_test = test_df$LCP_sum
  )
  
  # save outputs
  write.csv(test_df, fold_eval_file, row.names = FALSE)
}

# stop cluster
stopCluster(cl)
```

## Chunk 9: Evaluative metrics

After you run the chunk above once, this is the chunk you use for downstream plots. It loads the predictions, estimates evaluative metrics, and combines all fold metrics, then saves it to file in the results_dir.

Calculate metrics: Why Spearman instead of RSQ?

Fold-specific RSQ values were unstable because several held-out sites exhibited limited variance in observed CSE, making the denominator of RSQ (total sum of squares) small and sensitive to minor prediction error. In contrast, Spearman’s rank correlation provides a scale-independent measure of association and more robustly captures whether the model preserves relative differences among site pairs under these conditions. This is particularly relevant for resistance-based distance models, where relative ordering of pairwise connectivity is often more important than absolute prediction accuracy.

```{r 9-fold-metrics, eval = FALSE}
kdist = 1
for (fold_idx in seq_along(sites)) {

  site <- sites[fold_idx]

  fold_eval_file <- file.path(
    spatial_output_dir,
    sprintf("spatial_eval_fold_%02d.csv", fold_idx)
  )

  fold_metrics_file <- file.path(
    spatial_output_dir,
    sprintf("spatial_metrics_fold_%02d.csv", fold_idx)
  )

  if (!file.exists(fold_eval_file)) {
    warning(sprintf("Missing fold evaluation file: %s", fold_eval_file))
    next
  }

  test_df <- read.csv(fold_eval_file)

  metrics_k <- calc_metrics(
    obs = test_df$CSE,
    pred = test_df$LCP_sum_cal
  ) %>%
    mutate(
      site = site,
      projection_km = kdist,
      method = "LCP_sum"
    ) %>%
    dplyr::select(site, projection_km, method, everything())

  write.csv(metrics_k, fold_metrics_file, row.names = FALSE)
}

# combine all fold metrics
metric_files <- list.files(
  spatial_output_dir,
  pattern = "^spatial_metrics_fold_[0-9]{2}\\.csv$",
  full.names = TRUE
)

metrics_all_spatial <- bind_rows(lapply(metric_files, read.csv)) %>%
  mutate(site = factor(site, levels = sites)) %>%
  arrange(site)

write.csv(
  metrics_all_spatial,
  file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_summary.csv"),
  row.names = FALSE
)

print(metrics_all_spatial)

# pooled out-of-fold predictive R2 across all folds
eval_files <- list.files(
  spatial_output_dir,
  pattern = "^spatial_eval_fold_[0-9]{2}\\.csv$",
  full.names = TRUE
)

eval_all_spatial <- bind_rows(lapply(eval_files, read.csv))

keep_all <- complete.cases(eval_all_spatial$CSE, eval_all_spatial$LCP_sum_cal)

obs_all  <- eval_all_spatial$CSE[keep_all]
pred_all <- eval_all_spatial$LCP_sum_cal[keep_all]

R2_cv <- 1 - sum((obs_all - pred_all)^2) / sum((obs_all - mean(obs_all))^2)

pooled_summary_spatial <- data.frame(
  n = length(obs_all),
  pooled_R2 = R2_cv,
  pooled_Pearson = cor(obs_all, pred_all, method = "pearson"),
  pooled_Spearman = cor(obs_all, pred_all, method = "spearman"),
  pooled_RMSE = sqrt(mean((obs_all - pred_all)^2)),
  pooled_MAE = mean(abs(obs_all - pred_all))
)

write.csv(
  pooled_summary_spatial,
  file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_pooled_summary.csv"),
  row.names = FALSE
)

print(pooled_summary_spatial)
```

## Chunk 10: Plot spatial eval of LOPOCV

```{r 10-visualize-spatial-LCPsum, eval = TRUE}
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"

# Load Spatial LOPOCV summary
metrics_all <- read.csv(file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_summary.csv"))
print(metrics_all)
summary(metrics_all$Spearman)

# Load Spatial Pooled Stats
pooled_summary_spatial <- read.csv(file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_pooled_summary.csv"))
print(pooled_summary_spatial)

# Load raster for extent
altitude <- raster::raster(file.path(
  "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data/processed",
  "altitude_1KMmedian_MERIT_UgandaClip.tif"
))
crs(altitude) <- 4326

# Load site metadata including subcluster
indinfo <- read.delim("../input/Gff_11loci_allsites_indinfo.txt")
site_clusters <- indinfo %>%
  dplyr::select(Site = SiteCode, Subcluster = SiteMajCluster) %>%
  distinct()

# Build site metadata
site_metadata <- V.table %>%
  dplyr::select(Site = Var1, Latitude = lat1, Longitude = long1) %>%
  distinct() %>%
  left_join(site_clusters, by = "Site") %>%
  left_join(metrics_all, by = c("Site" = "site")) %>%
  mutate(Symbol = ifelse(Spearman < 0.3, "low", "circle")) %>%
  arrange(desc(Spearman))

# Extract map extent
r_ext <- extent(altitude)
xlim <- c(r_ext@xmin, r_ext@xmax)
ylim <- c(r_ext@ymin, r_ext@ymax)

# Natural Earth background
uganda <- ne_countries(scale = "medium", continent = "Africa", returnclass = "sf") %>% st_transform(4326)
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>% st_transform(4326)

# Plot Spatial LOPOCV Spearman by site
ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.1) +
  geom_sf(data = lakes, fill = "black", color = NA) +
  geom_point(
    data = filter(site_metadata, Symbol == "circle"),
    aes(x = Longitude, y = Latitude, size = Spearman, fill = Subcluster),
    shape = 21, color = "black", stroke = 0.3
  ) +
  geom_point(
    data = filter(site_metadata, Symbol == "low"),
    aes(x = Longitude, y = Latitude, fill = Subcluster),
    shape = 21, color = "black", size = 1, stroke = 0.3
  ) +
  scale_fill_manual(
    name = "Subcluster",
    values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")
  ) +
  scale_size_continuous(
  name = "LOPOCV \nSpearman's r",
  limits = c(0.4, 1.0),
  breaks = c(0.4, 0.7, 1.0),
  range = c(2,7)
) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Spatial LOPOCV Spearman's r by Site", x = "Longitude", y = "Latitude")

# Overlapping density plot
ggplot(site_metadata, aes(x = Spearman, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(
    title = "Distribution of Spearman's r by Subcluster",
    x = "Spearman's r (Spatial LOPOCV)",
    y = "Density"
  )

# Overlapping count plot
ggplot(site_metadata, aes(x = Spearman, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = after_stat(count))) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(
    title = "Spearman's r by Subcluster (Scaled by Count)",
    x = "Spearman's r (Spatial LOPOCV)",
    y = "Count"
  )

# Reassign Cluster
site_metadata$Cluster <- site_metadata$Subcluster
site_metadata$Cluster[site_metadata$Subcluster == "west"] <- "south"

# Plot Spatial LOPOCV Spearman by site just N/S
ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.1) +
  geom_sf(data = lakes, fill = "black", color = NA) +
  geom_point(
    data = filter(site_metadata, Symbol == "circle"),
    aes(x = Longitude, y = Latitude, size = Spearman, fill = Cluster),
    shape = 21, color = "black", stroke = 0.3
  ) +
  geom_point(
    data = filter(site_metadata, Symbol == "low"),
    aes(x = Longitude, y = Latitude, fill = Cluster),
    shape = 21, color = "black", size = 1, stroke = 0.3
  ) +
  scale_fill_manual(
    name = "Group",
    values = c("north" = "#1f78b4", "south" = "#e66101")
  ) +
  scale_size_continuous(
  name = "LOPOCV \nSpearman's r",
  limits = c(0.4, 1.0),
  breaks = c(0.4, 0.7, 1.0),
  range = c(2,7)
) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Spatial LOPOCV Spearman's r by Site", x = "Longitude", y = "Latitude")


# Plot Spatial LOPOCV Spearman by site just N/S with shaded colors
ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.1) +
  geom_sf(data = lakes, fill = "black", color = NA) +
  geom_point(
    data = filter(site_metadata, Symbol == "circle"),
    aes(x = Longitude, y = Latitude, size = Spearman, fill = Cluster),
    shape = 21, color = "black", stroke = 0.3, alpha = 0.4
  ) +
  geom_point(
    data = filter(site_metadata, Symbol == "low"),
    aes(x = Longitude, y = Latitude, fill = Cluster),
    shape = 21, color = "black", size = 1, stroke = 0.3, alpha = 0.4
  ) +
  scale_fill_manual(
    name = "Group",
    #values = c("north" = "#2c3e50", "south" = "#4e342e")
    values = c("north" = "#2A4F6E", "south" = "#7A4A2A")
    #values = c("north" = "#2E8BC6", "south" = "#FFA64D")
  ) +
  scale_size_continuous(
  name = "LOPOCV \nSpearman's r",
  limits = c(0.4, 1.0),
  breaks = c(0.4, 0.7, 1.0),
  range = c(2,7)
) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Spatial LOPOCV Spearman's r by Site", x = "Longitude", y = "Latitude")


# Overlapping density plot just S/N
ggplot(site_metadata, aes(x = Spearman, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(
    title = "Distribution of Spearman's r by Cluster",
    x = "Spearman's r (Spatial LOPOCV)",
    y = "Density"
  )

# Overlapping count plot just S/N
ggplot(site_metadata, aes(x = Spearman, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = after_stat(count))) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(
    title = "Spearman's r by Cluster (Scaled by Count)",
    x = "Spearman's r (Spatial LOPOCV)",
    y = "Count"
  )

# write.csv(site_metadata, file.path(results_dir, "site_metadata_spatial.csv"), row.names = FALSE)
```
