---
title: "LOPOCV: Spatial Cross-Validation"
author: "Norah Saarman"
date: "2025-06-18"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 8 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
- **Memory per job:** 200G (cluster limit: 1000G total; avoid exceeding half)    

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(randomForest)

# Define Paths to directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
output_dir <- paste0(results_dir,"/lopocv_spatial")
#dir.create(output_dir, showWarnings = FALSE)
scratch_dir <- "/scratch/general/vast/u6036559"

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

# define ggplot2 extent
xlim <- c(28.6, 35.4)
ylim <- c(-1.500000 , 4.733333)
```

# Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../results_dir/fullRF_CSE_resistance.tif` - Final full model projected resistance surface
  - `../results_dir/LC_paths_fullRF.shp"` - 
  - `../data_dir/processed/env_stack.grd`  - Final prediction env stack with named layers (18 variables)
env <- stack(file.path(
  - `../results_dir/lopocv/rf_model_01.rds` - 67 LOPOCV rf models leaving one point out
  
# Outputs  
  - `../results_dir/spatial_predictions.csv` - Spatial lopocv predicted geodist, CSE_per_unit, predicted CSE
  - `../results_dir/spatial_eval.csv` - Spatial lopocv evaluation metrics

# 1. Precompute: extract environmental data along each path once
**NOTE:** eval = FALSE so that it skips on knit
```{r env-paths, eval = FALSE}

# Input: raster of env parameters for prediction
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))
names(env) <-  c("BIO1_mean", "BIO2_mean", "BIO3_mean", "BIO4_mean", "BIO5_mean", "BIO6_mean", "BIO7_mean", "BIO8S_mean","BIO9S_mean", "BIO10S_mean", "BIO11S_mean", "BIO12_mean", "BIO13_mean", "BIO14_mean", "BIO15_mean", "BIO16S_mean","BIO17S_mean", "BIO18S_mean", "BIO19S_mean", "alt_mean","slope_mean", "riv_3km_mean","samp_20km_mean","lakes_mean","pix_dist")
crs(env) <- crs_geo
pix_dist <- env$pix_dist
crs(pix_dist) <- crs_geo

# Input: shape file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo

# Build list of unique sites from Var1 and Var2
sites <- sort(unique(c(lcp_sf$Var1, lcp_sf$Var2)))

# Set up parallel backend
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Parallel extraction of environmental values for each path
path_env_list <- foreach(i = 1:nrow(lcp_sf), .packages = c("raster", "sf")) %dopar% {
  path_geom <- st_geometry(lcp_sf[i, ]) |> as("Spatial")
  env_vals <- extract(env, path_geom)[[1]]
  if (is.null(env_vals)) return(NULL)
  as.data.frame(env_vals)
}
names(path_env_list) <- lcp_sf$id

# Stop parallel backend
stopCluster(cl)

# Save result
saveRDS(path_env_list, file.path(output_dir, "path_env_list.rds"))
```
# 2. Spatial R² on full model

## 2a. Spatial R² on full model based on raw CSE
```{r full-spatial-R2}
# Load full model
rf_model_full <- readRDS(file.path(results_dir, "rf_mean18_tuned.rds"))

# Load pre-extracted path-level env data
path_env_list <- readRDS(file.path(output_dir, "path_env_list.rds"))
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Load prepared file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo# Initialize output
fullmodel_preds <- data.frame()

# Define region based on Var1 
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", header=TRUE, sep = "\t")
north_sites <- Gff$SiteCode[Gff$SiteMajCluster == "north"]

# Initialize output again
fullmodel_preds <- data.frame()

# Loop over all path pairs
for (row in seq_len(nrow(lcp_sf))) {
  pair_id <- lcp_sf$id[row]
  env_path <- path_env_list[[pair_id]]
  names(env_path) <-  c("BIO1", "BIO2", "BIO3", "BIO4", "BIO5", "BIO6", "BIO7", "BIO8S","BIO9S", "BIO10S", "BIO11S", "BIO12", "BIO13", "BIO14", "BIO15", "BIO16S","BIO17S", "BIO18S", "BIO19S", "alt","slope", "riv_3km","samp_20km","lakes","pix_dist")
  #env_path$samp_20km <- 1.027064e-11 # neutralize sampling bias = commented out
  if (is.null(env_path) || nrow(env_path) == 0 || all(!complete.cases(env_path))) next

  pred_vals <- predict(rf_model_full, newdata = env_path)
  geo_dist <- sum(env_path$pix_dist, na.rm = TRUE)
  pred_mean <- sum(pred_vals, na.rm = TRUE) / geo_dist

  # Assign region (example: based on Var1)
  site1 <- lcp_sf$Var1[row]
  region <- if (site1 %in% north_sites) "North" else "South"

  # Use region-specific intercept and slope
  if (region == "North") {
    intercept <- 0.2747140
    slope <- 0.0006429
  } else {
    intercept <- 0.2931063
    slope <- 0.0008329
  }

  pred_CSE_IBD <- intercept + slope * geo_dist + pred_mean

  fullmodel_preds <- rbind(fullmodel_preds, data.frame(
    id = pair_id,
    Var1 = site1,
    Var2 = lcp_sf$Var2[row],
    Region = region,
    true_CSE = lcp_sf$CSE[row],
    pred_CSE = pred_mean,
    pred_IBD = pred_CSE_IBD
  ))
}

eval_metrics <- function(true, predicted) {
  resids <- true - predicted
  ss_res <- sum(resids^2, na.rm = TRUE)
  ss_tot <- sum((true - mean(true, na.rm = TRUE))^2, na.rm = TRUE)
  rsq <- 1 - ss_res / ss_tot
  rmse <- sqrt(mean(resids^2, na.rm = TRUE))
  mae <- mean(abs(resids), na.rm = TRUE)
  cor_val <- cor(true, predicted, use = "complete.obs")
  data.frame(R2 = rsq, RMSE = rmse, MAE = mae, Correlation = cor_val)
}

metrics_pred_only <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE)
metrics_pred_only

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (path-mean)")

metrics_IBD_adjusted <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_IBD)
metrics_IBD_adjusted

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_IBD)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (IBD-adjusted)")

# Post-hoc linear calibration
cal_model <- lm(true_CSE ~ pred_IBD, data = fullmodel_preds)
summary(cal_model)
fullmodel_preds$pred_CSE_calibrated <- predict(cal_model)

metrics_IBD_calibrated <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE_calibrated)
metrics_IBD_calibrated 

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE_calibrated)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (IBD-adjusted-calibrated)")
```

## 2b. Spatial R² on full model based on Scaled CSE (SCSE)
```{r scaled}
rf_model_full <- readRDS(file.path(results_dir, "rf_scaled_tuned.rds"))

# Load pre-extracted path-level env data
path_env_list <- readRDS(file.path(output_dir, "path_env_list.rds"))
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Load prepared file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo
names(lcp_sf)
max_CSE <- max(lcp_sf$CSE, na.rm = TRUE)
lcp_sf$CSE_scaled <- lcp_sf$CSE / max_CSE

# Initialize output
fullmodel_preds <- data.frame()

# replace CSE with CSE_scaled

# Define region based on Var1 
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", header=TRUE, sep = "\t")
north_sites <- Gff$SiteCode[Gff$SiteMajCluster == "north"]

# Initialize output again
fullmodel_preds <- data.frame()

# Loop over all path pairs
for (row in seq_len(nrow(lcp_sf))) {
  pair_id <- lcp_sf$id[row]
  env_path <- path_env_list[[pair_id]]
  names(env_path) <-  c("BIO1", "BIO2", "BIO3", "BIO4", "BIO5", "BIO6", "BIO7", "BIO8S","BIO9S", "BIO10S", "BIO11S", "BIO12", "BIO13", "BIO14", "BIO15", "BIO16S","BIO17S", "BIO18S", "BIO19S", "alt","slope", "riv_3km","samp_20km","lakes","pix_dist")
  #env_path$samp_20km <- 1.027064e-11 # neutralize sampling bias = commented out
  if (is.null(env_path) || nrow(env_path) == 0 || all(!complete.cases(env_path))) next

  pred_vals <- predict(rf_model_full, newdata = env_path)
  geo_dist <- sum(env_path$pix_dist, na.rm = TRUE)
  pred_mean <- sum(pred_vals, na.rm = TRUE) / geo_dist

  # Assign region (example: based on Var1)
  site1 <- lcp_sf$Var1[row]
  region <- if (site1 %in% north_sites) "North" else "South"

             
  # Use region-specific intercept and slope
  if (region == "North") {
    intercept <- 0.402927
    slope <- 0.000943
  } else {
    intercept <- 0.429903
    slope <- 0.001222
  }

  pred_CSE_IBD <- intercept + slope * geo_dist + pred_mean

  fullmodel_preds <- rbind(fullmodel_preds, data.frame(
    id = pair_id,
    Var1 = site1,
    Var2 = lcp_sf$Var2[row],
    Region = region,
    true_CSE = lcp_sf$CSE_scaled[row],
    pred_CSE = pred_mean,
    pred_IBD = pred_CSE_IBD
  ))
}



eval_metrics <- function(true, predicted) {
  resids <- true - predicted
  ss_res <- sum(resids^2, na.rm = TRUE)
  ss_tot <- sum((true - mean(true, na.rm = TRUE))^2, na.rm = TRUE)
  rsq <- 1 - ss_res / ss_tot
  rmse <- sqrt(mean(resids^2, na.rm = TRUE))
  mae <- mean(abs(resids), na.rm = TRUE)
  cor_val <- cor(true, predicted, use = "complete.obs")
  data.frame(R2 = rsq, RMSE = rmse, MAE = mae, Correlation = cor_val)
}

metrics_pred_only <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE)
metrics_pred_only

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.7) +
  ylim(0, 1.7) +
  theme_minimal() +
  labs(title = "Predicted vs Observed Scaled CSE (SCSE)",
       x = "Observed SCSE", y = "Predicted SCSE (path-mean)")

metrics_IBD_adjusted <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_IBD)
metrics_IBD_adjusted

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_IBD)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.7) +
  ylim(0, 1.7) +
  theme_minimal() +
  labs(title = "Predicted vs Observed Scaled CSE (SCSE)",
       x = "Observed SCSE", y = "Predicted SCSE (IBD-adjusted)")

# Post-hoc linear calibration
cal_model <- lm(true_CSE ~ pred_IBD, data = fullmodel_preds)
summary(cal_model)
fullmodel_preds$pred_CSE_calibrated <- predict(cal_model)

metrics_IBD_calibrated <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE_calibrated)
metrics_IBD_calibrated 

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE_calibrated)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed Scaled CSE (SCSE)",
       x = "Observed SCSE", y = "Predicted SCSE (IBD-adjusted-calibrated)")

```

The scaled CSE performs only very slighly better... so I will not scale CSE for the main study. 

Specifically:  
- Raw CSE already performs almost as well, and is easier to explain and justify.  
- Scaling CSE provides only a very small gain (ΔR² ≈ 0.0061), with slight trade-offs in RMSE and MAE (lower error in models built with raw CSE). The correlation is already strong (≥0.84) in both cases.  
- Scaling CSE obscures interpretation.

## 2c. Spatial R² raw CSE with sampling effort neutralized 
Sampling effort was retained during spatial evaluation due to its predictive value for CSE. Below we show that replacing env_path$samp_20km values with mean sampling density does not improve the model's predictive performance.

```{r neutralizing-samp-effort, eval = FALSE}
# Load full model
rf_model_full <- readRDS(file.path(results_dir, "rf_mean18_tuned.rds"))

# Load pre-extracted path-level env data
path_env_list <- readRDS(file.path(output_dir, "path_env_list.rds"))
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# Load prepared file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo# Initialize output
fullmodel_preds <- data.frame()

# Define region based on Var1 
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", header=TRUE, sep = "\t")
north_sites <- Gff$SiteCode[Gff$SiteMajCluster == "north"]

# Initialize output again
fullmodel_preds <- data.frame()

# Loop over all path pairs
for (row in seq_len(nrow(lcp_sf))) {
  pair_id <- lcp_sf$id[row]
  env_path <- path_env_list[[pair_id]]
  names(env_path) <-  c("BIO1", "BIO2", "BIO3", "BIO4", "BIO5", "BIO6", "BIO7", "BIO8S","BIO9S", "BIO10S", "BIO11S", "BIO12", "BIO13", "BIO14", "BIO15", "BIO16S","BIO17S", "BIO18S", "BIO19S", "alt","slope", "riv_3km","samp_20km","lakes","pix_dist")
  env_path$samp_20km <- 1.027064e-11 # neutralize sampling bias 
  if (is.null(env_path) || nrow(env_path) == 0 || all(!complete.cases(env_path))) next

  pred_vals <- predict(rf_model_full, newdata = env_path)
  geo_dist <- sum(env_path$pix_dist, na.rm = TRUE)
  pred_mean <- sum(pred_vals, na.rm = TRUE) / geo_dist

  # Assign region (example: based on Var1)
  site1 <- lcp_sf$Var1[row]
  region <- if (site1 %in% north_sites) "North" else "South"

  # Use region-specific intercept and slope
  if (region == "North") {
    intercept <- 0.2747140
    slope <- 0.0006429
  } else {
    intercept <- 0.2931063
    slope <- 0.0008329
  }

  pred_CSE_IBD <- intercept + slope * geo_dist + pred_mean

  fullmodel_preds <- rbind(fullmodel_preds, data.frame(
    id = pair_id,
    Var1 = site1,
    Var2 = lcp_sf$Var2[row],
    Region = region,
    true_CSE = lcp_sf$CSE[row],
    pred_CSE = pred_mean,
    pred_IBD = pred_CSE_IBD
  ))
}

eval_metrics <- function(true, predicted) {
  resids <- true - predicted
  ss_res <- sum(resids^2, na.rm = TRUE)
  ss_tot <- sum((true - mean(true, na.rm = TRUE))^2, na.rm = TRUE)
  rsq <- 1 - ss_res / ss_tot
  rmse <- sqrt(mean(resids^2, na.rm = TRUE))
  mae <- mean(abs(resids), na.rm = TRUE)
  cor_val <- cor(true, predicted, use = "complete.obs")
  data.frame(R2 = rsq, RMSE = rmse, MAE = mae, Correlation = cor_val)
}

metrics_pred_only <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE)
metrics_pred_only

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (path-mean neutralized")

metrics_IBD_adjusted <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_IBD)
metrics_IBD_adjusted

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_IBD)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (IBD-adjusted-neutralized)")

# Post-hoc linear calibration
cal_model <- lm(true_CSE ~ pred_IBD, data = fullmodel_preds)
summary(cal_model)
fullmodel_preds$pred_CSE_calibrated <- predict(cal_model)

metrics_IBD_calibrated <- eval_metrics(fullmodel_preds$true_CSE, fullmodel_preds$pred_CSE_calibrated)
metrics_IBD_calibrated 

ggplot(fullmodel_preds, aes(x = true_CSE, y = pred_CSE_calibrated)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  xlim(0, 1.2) +
  ylim(0, 1.2) +
  theme_minimal() +
  labs(title = "Predicted vs Observed CSE (CSE)",
       x = "Observed CSE", y = "Predicted CSE (IBD-adjusted-calibrated-neutralized)")
```

# 3. Spatial LOPOCV with IBD adjustment

Conceptual Integration of IBD Adjustment into **Spatial LOPOCV**
1. LOPOCV pipeline now... For each LOPOCV fold:  
 - Train a model leaving out one site.
 - Predict CSE across paths involving that site.
 - Sum/Mean predictions across each path → predicted_CSE.
 - Compare to true_CSE.

2. Why IBD adjustment is needed  
 - During **spatial evaluation**, the full model shows consistent under- or over-estimation because the effect of geographic distance is removed in the prediction raster (due to nullified pix_dist, geographic distance is always ZERO).
 - This leads to systematic bias, which you correct post hoc using linear IBD models.  
 
3. How to Apply IBD Adjustment to **Spatial LOPOCV**... For each LOPOCV fold: 
 - Train a model leaving out one site, already done.
 - Compute least-cost-paths from full model, already done.
 - Predict CSE across paths involving that site.
 - Mean pixel-level predictions across each path → pred_mean
 - Note: Mean is the same as per-unit predicted CSE, raw_sum of prediction / geo_dist  
 - Compute total geographic distance → geo_dist = sum of pix_dist along path  
 - Adjust: Use geo_dist to apply IBD adjustment where pred_mean is acting as the residuals for the IBD linear model:
     - Determine region (e.g., north or south) based on site location
    - Use region-specific IBD model (intercept & slope) to get: pred_IBD = intercept + slope * geo_dist + pred_mean
 - Calibrate: Fit the calibration model using the training predictions only:  
 - lm(true_CSE ~ pred_IBD, data = training_df)
 - Apply the calibration model to the testing predictions only.
 - Evaluate R²/RMSE/MAE using the calibrated predictions (calibration was estimated with training paths only, and applied to testing paths only).
 
**NOTE:** I wonder if training the model on the residuals from the IBD analysis as the response variable instead of raw CSE may be a more straight forward way to deal with geographic distance, although you would loose the ability to understand impact of geographic distance in the overall RF model, and so may get lower performance in the non-spatial evaluations. Something to think about!!! 

**NOTE:** eval = FALSE so that it skips on knit
```{r spatial-lopocv-ibd, eval = FALSE}
# Spatial LOPOCV with IBD adjustment and per-fold calibration
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Load inputs
path_env_list <- readRDS(file.path(output_dir, "path_env_list.rds"))
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo

# Define northern sites
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", sep = "\t")
north_sites <- Gff$SiteCode[Gff$SiteMajCluster == "north"]

# Output storage
metrics_calibrated <- foreach(fold_idx = seq_along(sites), .combine = rbind,
  .packages = c("dplyr", "randomForest")) %dopar% {

  rf_model <- readRDS(sprintf("%s/lopocv/rf_model_%02d.rds", results_dir, fold_idx))
  test_site <- sites[fold_idx]
  fold_df <- data.frame()

  for (row in seq_len(nrow(lcp_sf))) {
    pair_id <- lcp_sf$id[row]
    env_path <- path_env_list[[pair_id]]
    if (is.null(env_path) || nrow(env_path) == 0 || all(!complete.cases(env_path))) next

    pred_vals <- predict(rf_model, newdata = env_path)
    geo_dist <- sum(env_path$pix_dist, na.rm = TRUE)
    pred_mean <- sum(pred_vals, na.rm = TRUE) / geo_dist

    site1 <- lcp_sf$Var1[row]
    
    # Use region-specific intercept and slope for raw CSE fit to geodist in km
    if (region == "North") {
      intercept <- 0.2747140
      slope <- 0.0006429
    } else {
      intercept <- 0.2931063
      slope <- 0.0008329
    }

    pred_IBD <- intercept + slope * geo_dist + pred_mean

    fold_df <- rbind(fold_df, data.frame(
      Fold = fold_idx,
      Var1 = site1,
      Var2 = lcp_sf$Var2[row],
      id = pair_id,
      geo_dist = geo_dist,
      region = region,
      true_CSE = lcp_sf$CSE[row],
      pred_CSE = pred_mean,
      pred_IBD = pred_IBD
    ))
  }

  test_ids <- fold_df$id[fold_df$Var1 == test_site | fold_df$Var2 == test_site]
  train_df <- fold_df[!fold_df$id %in% test_ids, ]
  test_df  <- fold_df[fold_df$id %in% test_ids, ]

  cal_model <- lm(true_CSE ~ pred_IBD, data = train_df)
  test_df$pred_CSE_calibrated <- predict(cal_model, newdata = test_df)

  ss_res <- sum((test_df$true_CSE - test_df$pred_CSE_calibrated)^2, na.rm = TRUE)
  ss_tot <- sum((test_df$true_CSE - mean(test_df$true_CSE, na.rm = TRUE))^2)
  rsq <- 1 - ss_res / ss_tot
  rmse <- sqrt(mean((test_df$true_CSE - test_df$pred_CSE_calibrated)^2, na.rm = TRUE))
  mae <- mean(abs(test_df$true_CSE - test_df$pred_CSE_calibrated), na.rm = TRUE)
  cor_val <- cor(test_df$true_CSE, test_df$pred_CSE_calibrated, use = "complete.obs")

  data.frame(
    site = test_site,
    rsq_test = rsq,
    rmse = rmse,
    mae = mae,
    cor_test = cor_val
  )
}

stopCluster(cl)

# Save
write.csv(metrics_calibrated, file.path(results_dir, "spatial_LOPOCV_calibrated_summary.csv"), row.names = FALSE)
```

## Visualize Spatial LOPOCV results
```{r visualize}
# Load LOPOCV summary if not already in memory
metrics_all <- read.csv(file.path(results_dir, "spatial_LOPOCV_calibrated_summary.csv"))

# Load raster for extent
altitude <- raster::raster(file.path(
  "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data/processed",
  "altitude_1KMmedian_MERIT_UgandaClip.tif"
))
crs(altitude) <- 4326

# Load site metadata including subcluster
indinfo <- read.delim("../input/Gff_11loci_allsites_indinfo.txt")
site_clusters <- indinfo %>%
  dplyr::select(Site = SiteCode, Subcluster = SiteMajCluster) %>%
  distinct()

# Build site metadata from V.table and join with subclusters and metrics
site_metadata <- V.table %>%
  dplyr::select(Site = Var1, Latitude = lat1, Longitude = long1) %>%
  distinct() %>%
  left_join(site_clusters, by = "Site") %>%
  left_join(metrics_all, by = c("Site" = "site")) %>%
  mutate(Symbol = ifelse(rsq_test < 0.5, "low", "circle")) %>%
  arrange(desc(rsq_test))

# Extract map extent
r_ext <- extent(altitude)
xlim <- c(r_ext@xmin, r_ext@xmax)
ylim <- c(r_ext@ymin, r_ext@ymax)

# Natural Earth background
uganda <- ne_countries(scale = "medium", continent = "Africa", returnclass = "sf") %>% st_transform(4326)
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>% st_transform(4326)

# Plot LOPOCV R² by site
ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.5) +
  geom_sf(data = lakes, fill = "gray80", color = NA) +

  geom_point(data = filter(site_metadata, Symbol == "circle"),
             aes(x = Longitude, y = Latitude, size = rsq_test, fill = Subcluster),
             shape = 21, color = "black", stroke = 0.3) +

  geom_point(data = filter(site_metadata, Symbol == "low"),
             aes(x = Longitude, y = Latitude, color = Subcluster),
             shape = 8, size = 3) +

  scale_fill_manual(name = "Subcluster", values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  scale_color_manual(name = "Subcluster", values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  scale_size_continuous(name = "Test R²", range = c(2, 6)) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Spatial LOPOCV Test R² by Site", x = "Longitude", y = "Latitude")

# Overlapping density plot
ggplot(site_metadata, aes(x = rsq_test, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(title = "Distribution of Test R² by Subcluster",
       x = "Test R² (Spatial LOPOCV)", y = "Density")

# Overlapping count plot
ggplot(site_metadata, aes(x = rsq_test, fill = Subcluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = ..count..)) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101", "west" = "#39005A")) +
  theme_minimal() +
  labs(title = "Test R² by Subcluster (Scaled by Count)",
       x = "Test R² (Spatial LOPOCV)",
       y = "Count")

# reassign Cluster
site_metadata$Cluster <- site_metadata$Subcluster
site_metadata$Cluster[site_metadata$Subcluster == "west"]  <- "south"

# Plot LOPOCV R² by site just N/S
ggplot() +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.5) +
  geom_sf(data = lakes, fill = "gray80", color = NA) +

  geom_point(data = filter(site_metadata, Symbol == "circle"),
             aes(x = Longitude, y = Latitude, size = rsq_test, fill = Cluster),
             shape = 21, color = "black", stroke = 0.3) +

  geom_point(data = filter(site_metadata, Symbol == "low"),
             aes(x = Longitude, y = Latitude, color = Cluster),
             shape = 8, size = 3) +

  scale_fill_manual(name = "Cluster", values = c("north" = "#1f78b4", "south" = "#e66101")) +
  scale_color_manual(name = "Cluster", values = c("north" = "#1f78b4", "south" = "#e66101")) +
  scale_size_continuous(name = "Test R²", range = c(2, 6)) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Spatial LOPOCV Test R² by Site", x = "Longitude", y = "Latitude")

# Overlapping density plot just S/N
ggplot(site_metadata, aes(x = rsq_test, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(title = "Distribution of Test R² by Cluster",
       x = "Test R² (Spatial LOPOCV)", y = "Density")

# Overlapping count plot just S/N
ggplot(site_metadata, aes(x = rsq_test, fill = Cluster)) +
  geom_density(alpha = 0.5, color = NA, position = "identity", aes(y = ..count..)) +
  scale_fill_manual(values = c("north" = "#1f78b4", "south" = "#e66101")) +
  theme_minimal() +
  labs(title = "Test R² by Cluster (Scaled by Count)",
       x = "Test R² (Spatial LOPOCV)",
       y = "Count")
```

# 4. Permutation test for spatial LOPOCV R²

GOAL: Create a 67 x 100 matrix where each cell is the rsq_test value for one LOPOCV fold (site) under a permuted response (CSEdistance)

## Spatial LOPOCV on permuted models
**NOTE:** eval = FALSE so that it skips on knit
```{r perm-spatial-lopocv-calibrated, eval = FALSE}
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Load inputs
path_env_list <- readRDS(file.path(output_dir, "path_env_list.rds"))
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo

# Define northern sites
Gff <- read.csv("../input/Gff_11loci_allsites_indinfo.txt", sep = "\t")
north_sites <- Gff$SiteCode[Gff$SiteMajCluster == "north"]

metrics_perm_list <- foreach(p = 1:100, .packages = c("dplyr", "randomForest")) %dopar% {
  metrics_all <- data.frame()

  for (fold_idx in seq_along(sites)) {
    rf_model <- readRDS(sprintf("/scratch/general/vast/u6036559/rf_model_perm_%02d_fold_%02d.rds", p, fold_idx))
    test_site <- sites[fold_idx]
    fold_df <- data.frame()

    for (row in seq_len(nrow(lcp_sf))) {
      pair_id <- lcp_sf$id[row]
      env_path <- path_env_list[[pair_id]]
      if (is.null(env_path) || nrow(env_path) == 0 || all(!complete.cases(env_path))) next

      pred_vals <- predict(rf_model, newdata = env_path)
      geo_dist <- sum(env_path$pix_dist, na.rm = TRUE)
      pred_mean <- sum(pred_vals, na.rm = TRUE) / geo_dist

      site1 <- lcp_sf$Var1[row]
      region <- if (site1 %in% north_sites) "North" else "South"

      if (region == "North") {
        intercept <- 0.2747140
        slope <- 0.0006429
      } else {
        intercept <- 0.2931063
        slope <- 0.0008329
      }

      pred_IBD <- intercept + slope * geo_dist + pred_mean

      fold_df <- rbind(fold_df, data.frame(
        Fold = fold_idx,
        Var1 = site1,
        Var2 = lcp_sf$Var2[row],
        id = pair_id,
        geo_dist = geo_dist,
        region = region,
        true_CSE = lcp_sf$CSE[row],
        pred_CSE = pred_mean,
        pred_IBD = pred_IBD
      ))
    }

    test_ids <- fold_df$id[fold_df$Var1 == test_site | fold_df$Var2 == test_site]
    train_df <- fold_df[!fold_df$id %in% test_ids, ]
    test_df  <- fold_df[fold_df$id %in% test_ids, ]

    cal_model <- lm(true_CSE ~ pred_IBD, data = train_df)
    test_df$pred_CSE_calibrated <- predict(cal_model, newdata = test_df)

    ss_res <- sum((test_df$true_CSE - test_df$pred_CSE_calibrated)^2, na.rm = TRUE)
    ss_tot <- sum((test_df$true_CSE - mean(test_df$true_CSE, na.rm = TRUE))^2)
    rsq_test <- 1 - ss_res / ss_tot
    rmse <- sqrt(mean((test_df$true_CSE - test_df$pred_CSE_calibrated)^2, na.rm = TRUE))
    mae <- mean(abs(test_df$true_CSE - test_df$pred_CSE_calibrated), na.rm = TRUE)
    cor_val <- cor(test_df$true_CSE, test_df$pred_CSE_calibrated, use = "complete.obs")

    metrics_all <- rbind(metrics_all, data.frame(
      permutation = p,
      site = test_site,
      rsq_test = rsq_test,
      rmse = rmse,
      mae = mae,
      cor_test = cor_val
    ))
  }

  write.csv(metrics_all,
            file.path(output_dir, sprintf("spatial_perm_LOPOCV_calibrated_%03d.csv", p)),
            row.names = FALSE)

  return(NULL)
}

stopCluster(cl)
message("Calibrated permutation-based spatial LOPOCV complete.")

# Combine into rsq matrix
n_sites <- length(sites)
n_perms <- 100
rsq_null_spatial <- matrix(NA, nrow = n_sites, ncol = n_perms)
rownames(rsq_null_spatial) <- sites
colnames(rsq_null_spatial) <- paste0("perm_", 1:n_perms)

for (p in 1:n_perms) {
  perm_file <- file.path(output_dir, sprintf("spatial_perm_LOPOCV_calibrated_%03d.csv", p))
  if (!file.exists(perm_file)) {
    warning(sprintf("File missing: %s", perm_file))
    next
  }

  df <- read.csv(perm_file)
  df <- df[match(sites, df$site), ]
  rsq_null_spatial[, p] <- df$rsq_test
}

write.csv(rsq_null_spatial, file.path(output_dir, "rsq_spatial_LOPOCV_calibrated_null_matrix.csv"), row.names = TRUE)
write.csv(rsq_null_spatial, "../results/rsq_spatial_LOPOCV_calibrated_null_matrix.csv", row.names = TRUE)

message("Calibrated 67 x 100 spatial R² null matrix saved.")
```

## Wilcoxon signed rank test for empirical p-values
GOAL: Run wilcox.test() for each column of rsq_null_spatial... This provides a distribution of test statistics (or p-values), from which you compute an empirical p-value.

Plot the distribution of observed/permuted R² values 
```{r plot-r2-spatial}
# Load permuted R² matrix and convert to long format
rsq_null_spatial_mat <- as.matrix(read.csv("../results/rsq_spatial_LOPOCV_calibrated_null_matrix.csv"))
rsq_null_spatial_long <- as.vector(rsq_null_spatial_mat)

# Load LOPOCV summary if not already in memory
metrics_all <- read.csv(file.path(results_dir, "spatial_LOPOCV_calibrated_summary.csv"))
rsq_obs <- metrics_all$rsq_test

# Build data frame for plotting
plot_df_rsq <- data.frame(
  rsq = c(rsq_obs, rsq_null_spatial_long),
  type = c(rep("Observed", length(rsq_obs)),
           rep("Permuted", length(rsq_null_spatial_long)))
)

# Plot density comparison
ggplot(plot_df_rsq, aes(x = rsq, fill = type)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c("Observed" = "#1f78b4", "Permuted" = "gray70")) +
  xlim(-1, 1) +
  theme_minimal() +
  labs(title = "Observed vs. Permuted R² (LOPOCV)",
       x = "Test R²",
       y = "Density")
```

Calculate a Wilcoxon signed-rank statistic for each column (permuted replicate) and a one-sided p-value as the proportion of permuted statistics ≥ observed.
```{r Wilcoxon-spatial}
# Vector of observed R² values (length = number of folds)
rsq_obs <- metrics_all$rsq_test

# Matrix of permuted R² values: rows = folds, columns = permutations
rsq_null_spatial_mat <- as.matrix(read.csv("../results/rsq_spatial_LOPOCV_null_matrix.csv"))

# Function to compute signed rank statistic for a single permuted replicate
signed_rank_stat <- function(null_rsq) {
  # Remove any rows with missing observed values
  non_na_idx <- !is.na(rsq_obs) & !is.na(null_rsq)
  obs <- rsq_obs[non_na_idx]
  null <- null_rsq[non_na_idx]
  
  # Calculate signed ranks
  diffs <- obs - null
  signed_ranks <- rank(abs(diffs)) * sign(diffs)
  sum(signed_ranks)
}

# Observed signed-rank statistic (using observed vs. permuted mean)
obs_stat <- signed_rank_stat(rowMeans(rsq_null_spatial_mat, na.rm = TRUE))

# Null distribution of signed-rank stats across permutations
perm_stats <- apply(rsq_null_spatial_mat, 2, signed_rank_stat)

# Empirical p-value (one-sided: observed > permuted)
p_val <- mean(perm_stats <= obs_stat)

# Output
cat(sprintf("Empirical p-value for spatial eval (Wilcoxon signed-rank test across spatial folds): %.4f\n", p_val))
```