---
title: "LOPOCV with permutated data as Null, then GeoDist as null"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../scripts_knitted_md", 
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/knitted_mds/" 
)
```

RStudio Configuration:  
   - **R version:** R 4.4.0 (Geospatial packages)  
   - **Number of cores:** 4 (up to 32 available)   
   - **Account:** saarman-np  
   - **Partition:** saarman-np (now auto allows multiple simultaneous jobs)  
   - **Memory per job:** 100G (cluster limit: 1000G total; avoid exceeding half)
  
  
  
## Overview of script
Although model configuration and projection parameters were selected based on full-model performance (RSQ), during LOPOCV, some held-out folds exhibited limited variance in observed CSE, making fold-specific RSQ unstable. We therefore report fold-level performance using Spearman and Pearson correlation alongside RMSE and MAE.

(In contrast, model configuration and projection parameters were selected based on full-model performance (RSQ) because in these cases, variance in CSE was sufficiently large for stable estimation.)

## Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../results_dir/fullRF_CSE_resistance.tif` - Final full model projected resistance surface
  - `../results_dir/LC_paths_fullRF.shp"` - 
  - `../data_dir/processed/env_stack.grd`  - Final prediction env stack with named layers (18 variables)
env <- stack(file.path(
  - `../results_dir/lopocv/rf_model_01.rds` - 67 LOPOCV rf models leaving one point out
  
## Libraries
```{r libraries, warning=FALSE, results=FALSE, message=FALSE, eval=TRUE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(randomForest)
library(readr)
library(ggplot2)
library(sp)
library(tidyverse)
library(rnaturalearth)
library(rnaturalearthdata)
library(reshape2)
```

## Define Paths to Directories
```{r general-dir-set-up, eval=TRUE}

# Define paths
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
input_dir <- "../input"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
output_dir <- file.path(results_dir, "lopocv")
perm_model_dir <- file.path(results_dir, "perm_rds_temp")
dir.create(perm_model_dir, recursive = TRUE, showWarnings = FALSE)
stopifnot(dir.exists(perm_model_dir))
stopifnot(file.access(perm_model_dir, 2) == 0)

# High-quality figure output (for manuscript)
fig_dir <- file.path(results_dir, "figures_pub")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84

# define ggplot2 extent
xlim <- c(28.6, 35.4)
ylim <- c(-1.500000 , 4.733333)

# Input: shape file of least-cost paths (already filtered to 67 sites and has CSEdistance)
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp"))
st_crs(lcp_sf) <- crs_geo

# Build list of unique sites from Var1 and Var2
sites <- sort(unique(c(lcp_sf$Var1, lcp_sf$Var2)))

output_dir <- file.path(results_dir, "lopocv")
#dir.create(output_dir, showWarnings = FALSE)
scratch_dir <- "/scratch/local/u6036559"
```


# Permutation test for LOPOCV Spearman's

GOAL: Create and test if different: 67 x 100 matrix where each cell is the Spearmans's correlation for one LOPOCV fold (site) under a permuted response (CSEdistance).

## Chunk 1: LOPOCV on permuted models

helpers:
```{r 1a-helpers-permutation-LOPOCV-models, eval = TRUE}

# Chunk 1: setup + data + helpers 

# Load required packages
library(dplyr)
library(readr)
library(randomForest)
library(doParallel)
library(foreach)

# Load data
V.table_full <- read.csv(file.path(input_dir, "Gff_cse_envCostPaths.csv"))

# Filter out western outlier
V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB")

# Create unique pair ID
V.table$id <- paste(V.table$Var1, V.table$Var2, sep = "_")

# Define site list
sites <- sort(unique(c(V.table$Var1, V.table$Var2)))

# Predictors used in the original LOPOCV models
predictor_vars <- c(
  "BIO1_mean","BIO2_mean","BIO3_mean","BIO4_mean","BIO5_mean","BIO6_mean",
  "BIO7_mean","BIO8S_mean","BIO9S_mean","BIO10S_mean","BIO11S_mean",
  "BIO12_mean","BIO13_mean","BIO14_mean","BIO15_mean","BIO16S_mean",
  "BIO17S_mean","BIO18S_mean","BIO19S_mean","slope_mean","alt_mean",
  "lakes_mean","riv_3km_mean","samp_20km_mean","pix_dist"
)

# Modeling table
V.model <- V.table[, c("CSEdistance", predictor_vars)]
```
1b run the permutations and save .rds files
```{r 1b-permutation-LOPOCV-models, eval = FALSE}
#run permuted LOPOCV models and save .rds files

# Set number of permutations
n_perm <- 100

# Set up parallel backend
cl <- makeCluster(8)
registerDoParallel(cl)

# Run permutations and save models
foreach(
  p = 1:n_perm,
  .packages = c("randomForest", "dplyr")
) %dopar% {
  set.seed(500 + p)

  # Shuffle response column while preserving row order
  V.model_perm <- V.model
  V.model_perm$CSEdistance <- sample(V.model_perm$CSEdistance)

  for (i in seq_along(sites)) {
    site <- sites[i]

    # Identify test/train indices
    test_ids  <- V.table$id[V.table$Var1 == site | V.table$Var2 == site]
    test_idx  <- which(V.table$id %in% test_ids)
    train_idx <- which(!V.table$id %in% test_ids)

    train_df <- V.model_perm[train_idx, ]
    test_df  <- V.model_perm[test_idx, ]

    # Fit Random Forest on permuted response
    rf_model <- randomForest(
      CSEdistance ~ .,
      data = train_df,
      ntree = 500
    )

    # Save RF model for this permutation–fold pair
    saveRDS(
  rf_model,
  file.path(perm_model_dir, sprintf("rf_model_perm_%02d_fold_%02d.rds", p, i))
)
    # Save metrics file for this permutation–fold pair
    #metrics_file <- file.path(results_dir, sprintf("perm_progress_%03d.csv", p))                               # commented this out because it is pretty much a blank file with just the permutation number, some mistake i made...
#write.csv(data.frame(perm = p), metrics_file, row.names = FALSE)
  }

  NULL
}

# Stop cluster
stopCluster(cl)

```
## Chunk 2: Calculate the metrics
```{r permutation-LOPOCV-metrics, eval = FALSE}
# Chunk 2: calculate Spearman null matrix from saved permutation models

# Helper for calibration
calibrate_pred <- function(obs_train, pred_train, pred_test) {
  keep <- complete.cases(obs_train, pred_train)
  df <- data.frame(obs = obs_train[keep], pred = pred_train[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(pred_test))
  keep_test <- complete.cases(pred_test)
  out[keep_test] <- predict(fit, newdata = data.frame(pred = pred_test[keep_test]))
  out
}

# Set number of permutations
n_perm <- 100

# Preallocate matrix for results
spearman_null <- matrix(NA_real_, nrow = length(sites), ncol = n_perm)
rownames(spearman_null) <- sites
colnames(spearman_null) <- paste0("perm_", seq_len(n_perm))

for (p in seq_len(n_perm)) {
  set.seed(500 + p)

  # Recreate permuted response exactly as in Chunk 1
  V.model_perm <- V.model
  V.model_perm$CSEdistance <- sample(V.model_perm$CSEdistance)

  spearman_perm <- numeric(length(sites))

  for (i in seq_along(sites)) {
    site <- sites[i]

    fold_model_file <- file.path(
      perm_model_dir,
      sprintf("rf_model_perm_%02d_fold_%02d.rds", p, i)
    )

    if (!file.exists(fold_model_file)) {
      warning(sprintf("Missing model file: %s", fold_model_file))
      spearman_perm[i] <- NA_real_
      next
    }

    # Identify test/train indices
    test_ids  <- V.table$id[V.table$Var1 == site | V.table$Var2 == site]
    test_idx  <- which(V.table$id %in% test_ids)
    train_idx <- which(!V.table$id %in% test_ids)

    train_df <- V.model_perm[train_idx, ]
    test_df  <- V.model_perm[test_idx, ]

    rf_model <- readRDS(fold_model_file)

    train_obs <- train_df$CSEdistance
    test_obs  <- test_df$CSEdistance

    train_df <- train_df %>% dplyr::select(-CSEdistance)
    test_df  <- test_df %>% dplyr::select(-CSEdistance)

    stopifnot(!"CSEdistance" %in% names(train_df))
    stopifnot(!"CSEdistance" %in% names(test_df))

    stopifnot(identical(names(train_df), predictor_vars))
    stopifnot(identical(names(test_df), predictor_vars))
    stopifnot(identical(attr(rf_model$terms, "term.labels"), predictor_vars))

    pred_train <- predict(rf_model, newdata = train_df)
    pred_test  <- predict(rf_model, newdata = test_df)

    pred_test_cal <- calibrate_pred(
      obs_train = train_obs,
      pred_train = pred_train,
      pred_test = pred_test
    )

    keep <- complete.cases(test_obs, pred_test_cal)

    spearman_perm[i] <- cor(
      test_obs[keep],
      pred_test_cal[keep],
      method = "spearman"
    )
  }

  spearman_null[, p] <- spearman_perm
}

# Save matrix to file for later use
write.csv(
  spearman_null,
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = TRUE
)

print(spearman_null)
message("Permutation Spearman null matrix created and saved.")
```
  
## Chunk 3: Compare observed vs permuted LOPOCV Spearman
```{r plot-spearman-lopocv,eval=TRUE}
# observed LOPOCV summary from the true data
metrics_all <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))

# permutation matrix
spearman_null <- read.csv(
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = 1,
  check.names = FALSE
)

# make sure site order matches the permutation matrix
metrics_all <- metrics_all[match(rownames(spearman_null), metrics_all$site), ]

spearman_obs <- as.numeric(metrics_all$Spearman)

spearman_null_mat <- as.matrix(spearman_null)

# reorder rows to observed site order if needed
spearman_null_mat <- spearman_null_mat[match(metrics_all$site, rownames(spearman_null_mat)), ]

plot_df_spearman <- data.frame(
  spearman = c(spearman_obs, as.numeric(spearman_null_mat)),
  type = c(
    rep("Observed", length(spearman_obs)),
    rep("Permuted", length(as.numeric(spearman_null_mat)))
  )
)

ggplot(plot_df_spearman, aes(x = spearman, fill = type)) +
  geom_density(alpha = 0.5, color = NA) +
  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Observed vs. Permuted Spearman Correlation (LOPOCV)",
    x = "Spearman Correlation",
    y = "Density"
  )
```

## Chunk 4: Empirical p-value

GOAL: Calculate an empirical p-value comparing observed vs permuted Spearman correlation.

```{r empirical-p-lopocv, warning=FALSE, eval=TRUE}

# Define region based on SiteMajCluster
Gff <- read.csv(file.path(input_dir, "Gff_11loci_allsites_indinfo.txt"),
                header = TRUE, sep = "\t")
north_sites <- unique(Gff$SiteCode[Gff$SiteMajCluster == "north"])

# Load observed Spearman
metrics_all <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))
spearman_obs_df <- data.frame(
  site = metrics_all$site,
  spearman = metrics_all$Spearman
)
spearman_obs_df$cluster <- ifelse(spearman_obs_df$site %in% north_sites, "north", "south")
spearman_obs_df$type <- "Observed"

spearman_null_raw <- read.csv(
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = 1,
  check.names = FALSE
)
spearman_null_mat <- as.matrix(spearman_null_raw)

# align rows to observed site order
spearman_null_mat <- spearman_null_mat[match(metrics_all$site, rownames(spearman_null_mat)), ]

obs_mean_spearman <- mean(spearman_obs_df$spearman, na.rm = TRUE)
null_mean_spearman <- apply(spearman_null_mat, 2, mean, na.rm = TRUE)

# one-sided empirical p-value: observed mean Spearman greater than null
p_empirical <- (sum(null_mean_spearman >= obs_mean_spearman, na.rm = TRUE) + 1) /
  (sum(!is.na(null_mean_spearman)) + 1)

cat("Observed mean Spearman:", round(obs_mean_spearman, 4), "\n")
cat("Null mean Spearman range:", round(range(null_mean_spearman, na.rm = TRUE), 4), "\n")
cat("Empirical one-sided p-value:", round(p_empirical, 4), "\n\n")

obs_north <- spearman_obs_df$spearman[spearman_obs_df$cluster == "north"]
obs_south <- spearman_obs_df$spearman[spearman_obs_df$cluster == "south"]

null_north_mat <- spearman_null_mat[rownames(spearman_null_mat) %in% north_sites, , drop = FALSE]
null_south_mat <- spearman_null_mat[!rownames(spearman_null_mat) %in% north_sites, , drop = FALSE]

obs_mean_north <- mean(obs_north, na.rm = TRUE)
obs_mean_south <- mean(obs_south, na.rm = TRUE)

null_mean_north <- apply(null_north_mat, 2, mean, na.rm = TRUE)
null_mean_south <- apply(null_south_mat, 2, mean, na.rm = TRUE)

p_north <- (sum(null_mean_north >= obs_mean_north, na.rm = TRUE) + 1) /
  (sum(!is.na(null_mean_north)) + 1)
p_south <- (sum(null_mean_south >= obs_mean_south, na.rm = TRUE) + 1) /
  (sum(!is.na(null_mean_south)) + 1)

cat("Observed mean Spearman, north:", round(obs_mean_north, 4), "\n")
cat("Empirical p-value, north:", round(p_north, 4), "\n\n")

cat("Observed mean Spearman, south:", round(obs_mean_south, 4), "\n")
cat("Empirical p-value, south:", round(p_south, 4), "\n")
```

## Chunk 5: Visualize permutated vs observed LOPOCV side-by-side

```{r plot-side-by-side, eval=TRUE}

# Define region based on SiteMajCluster
Gff <- read.csv(file.path(input_dir, "Gff_11loci_allsites_indinfo.txt"),
                header = TRUE, sep = "\t")
north_sites <- unique(Gff$SiteCode[Gff$SiteMajCluster == "north"])

# Load observed Spearman
metrics_all <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))
spearman_obs_df <- data.frame(
  site = metrics_all$site,
  spearman = metrics_all$Spearman
)
spearman_obs_df$cluster <- ifelse(spearman_obs_df$site %in% north_sites, "north", "south")
spearman_obs_df$type <- "Observed"

# Load permuted Spearman matrix
spearman_null <- read.csv(
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = 1,
  check.names = FALSE
)

spearman_null_df <- data.frame(
  site = rownames(spearman_null),
  cluster = ifelse(rownames(spearman_null) %in% north_sites, "north", "south"),
  spearman_null,
  check.names = FALSE
)

# Stack all permutation columns
spearman_null_long <- reshape2::melt(
  spearman_null_df,
  id.vars = c("site", "cluster"),
  variable.name = "perm",
  value.name = "spearman"
)
spearman_null_long$type <- "Permuted"

# Combine
spearman_plot_df <- rbind(
  spearman_obs_df[, c("site", "cluster", "spearman", "type")],
  spearman_null_long[, c("site", "cluster", "spearman", "type")]
)

# Plot
ggplot(spearman_plot_df, aes(x = spearman, fill = interaction(type, cluster))) +
  geom_density(
    data = subset(spearman_plot_df, type == "Observed"),
    aes(y = after_stat(count)),
    alpha = 0.6,
    color = NA,
    position = "identity"
  ) +
  geom_density(
    data = subset(spearman_plot_df, type == "Permuted"),
    aes(y = after_stat(count / 100)),
    alpha = 0.3,
    color = NA,
    position = "identity"
  ) +
  scale_fill_manual(
    name = "Group",
    values = c(
      "Observed.north" = "#1f78b4",
      "Observed.south" = "#e66101",
      "Permuted.north" = "#4A545E",
      "Permuted.south" = "#5A4F49"
      #"Permuted.north" = "#2c3e50",
      #"Permuted.south" = "#4e342e"
    ),
    labels = c(
      "Observed.north" = "North",
      "Observed.south" = "South",
      "Permuted.north" = "Permuted (North)",
      "Permuted.south" = "Permuted (South)"
    )
  ) +
  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Observed vs. Permuted Spearman by cluster",
    x = "Spearman's r (LOPOCV)",
    y = "Density"
  )


```

## Chunk 6: Spatial included, all side-by-side

Now with the Spatial Spearman's R included... 
"north" = "#2A4F6E"
"south" = "#7A4A2A"
```{r plot-side-by-side-with-spatial, eval=TRUE}

results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
input_dir <- "../input"
fig_dir <- file.path(results_dir, "figures_pub")

# Run just once to save high quality png for publication
png(file.path(fig_dir, "Fig_LOPOCV_obs_vs_permuted.png"),
    width = 7, height = 2.8, units = "in", res = 600)

# Define region based on SiteMajCluster
Gff <- read.csv(
  file.path(input_dir, "Gff_11loci_allsites_indinfo.txt"),
  header = TRUE, sep = "\t"
)
north_sites <- unique(Gff$SiteCode[Gff$SiteMajCluster == "north"])

# 1. Non-spatial LOPOCV Spearman
metrics_obs <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))

obs_df <- data.frame(
  site = metrics_obs$site,
  cluster = ifelse(metrics_obs$site %in% north_sites, "north", "south"),
  spearman = metrics_obs$Spearman,
  method = "Observed"
)

# 2. Null LOPOCV Spearman from permutations
spearman_null <- read.csv(
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = 1,
  check.names = FALSE
)

perm_df <- data.frame(
  site = rownames(spearman_null),
  cluster = ifelse(rownames(spearman_null) %in% north_sites, "north", "south"),
  spearman_null,
  check.names = FALSE
)

perm_long <- reshape2::melt(
  perm_df,
  id.vars = c("site", "cluster"),
  variable.name = "perm",
  value.name = "spearman"
) %>%
  mutate(method = "Permuted")

# 3. Spatial LOPOCV Spearman
metrics_spatial <- read.csv(
  file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_summary.csv")
)

spatial_df <- data.frame(
  site = metrics_spatial$site,
  cluster = ifelse(metrics_spatial$site %in% north_sites, "north", "south"),
  spearman = metrics_spatial$Spearman,
  method = "Spatial"
)

# Combine all three
plot_df <- bind_rows(
  obs_df[, c("site", "cluster", "spearman", "method")],
  perm_long[, c("site", "cluster", "spearman", "method")],
  spatial_df[, c("site", "cluster", "spearman", "method")]
)

# Set method and cluster order for legend/display
plot_df$method <- factor(
  plot_df$method,
  levels = c("Observed", "Spatial", "Permuted")
)

plot_df$cluster <- factor(
  plot_df$cluster,
  levels = c("north", "south")
)

# Plot
ggplot(plot_df, aes(x = spearman, fill = interaction(method, cluster, sep = "."))) +

  geom_density(
    data = subset(plot_df, method == "Observed"),
    aes(y = after_stat(count)),
    alpha = 0.6,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df, method == "Spatial"),
    aes(y = after_stat(count)),
    alpha = 0.5,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df, method == "Permuted"),
    aes(y = after_stat(count / 100)),
    alpha = 0.3,
    color = NA,
    position = "identity"
  ) +

  scale_fill_manual(
    name = "Prediction type",
    breaks = c(
      "Observed.north", "Observed.south",
      "Spatial.north",  "Spatial.south",
      "Permuted.north", "Permuted.south"
    ),
    values = c(
      "Observed.north" = "#1f78b4",
      "Observed.south" = "#e66101",
      "Spatial.north"  = "#2A4F6E",
      "Spatial.south"  = "#7A4A2A",
      "Permuted.north" = "#4A545E",
      "Permuted.south" = "#5A4F49"
    ),
    labels = c(
      "Observed.north" = "RF-predicted North",
      "Observed.south" = "RF-predicted South",
      "Spatial.north"  = "LCP-predicted North",
      "Spatial.south"  = "LCP-predicted South",
      "Permuted.north" = "Null-predicted North",
      "Permuted.south" = "Null-predicted South"
    )
  ) +
  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Fold-level LOPOCV Spearman by prediction type and cluster",
    x = "Spearman's r",
    y = "Count"
  )


dev.off()
```

# Geodist baseline test for LOPOCV Spearman's

GOAL: Create and test if different from full model: 67 x 100 matrix where each cell is the Spearmans's correlation for one LOPOCV fold (site) under a baseline model (geographic distance + sampling density only, excluding all environmental variables).

## Chunk 7: Geodist-only baseline LOPOCV eval 

This baseline model uses only:
1. pix_dist
2. samp_20km_mean

Sampling density is included during training, but neutralized before prediction, matching the approach used in the spatial eval.

```{r geodist-null-LOPOCV-models-and-metrics, eval = FALSE}

# Chunk 7: run one geodist-only LOPOCV baseline model and calculate metrics

# Output directory for geodist baseline model objects
geodist_model_dir <- file.path(results_dir, "geodist_rds_temp")
dir.create(geodist_model_dir, recursive = TRUE, showWarnings = FALSE)
stopifnot(dir.exists(geodist_model_dir))
stopifnot(file.access(geodist_model_dir, 2) == 0)

# Mean sampling density used for neutralizing sampling effort
geodist_samp_neutral <- mean(V.table$samp_20km_mean, na.rm = TRUE)

# Geodist-only model predictors
geodist_predictor_vars <- c("pix_dist", "samp_20km_mean")

# Modeling table
V.model_geodist <- V.table[, c("CSEdistance", geodist_predictor_vars)]

# Helper for calibration
calibrate_pred <- function(obs_train, pred_train, pred_test, return_fit = FALSE) {
  keep <- complete.cases(obs_train, pred_train)
  df <- data.frame(obs = obs_train[keep], pred = pred_train[keep])
  fit <- lm(obs ~ pred, data = df)

  out <- rep(NA_real_, length(pred_test))
  keep_test <- complete.cases(pred_test)
  out[keep_test] <- predict(fit, newdata = data.frame(pred = pred_test[keep_test]))

  if (return_fit) {
    list(
      pred_cal = out,
      fit = fit
    )
  } else {
    out
  }
}

# Helper for fold-level metrics
calc_metrics <- function(obs, pred) {
  keep <- complete.cases(obs, pred)

  obs  <- obs[keep]
  pred <- pred[keep]

  mse      <- mean((obs - pred)^2)
  rmse     <- sqrt(mse)
  mae      <- mean(abs(obs - pred))
  pearson  <- cor(obs, pred, method = "pearson")
  spearman <- cor(obs, pred, method = "spearman")

  data.frame(
    n = length(obs),
    MSE = mse,
    RMSE = rmse,
    MAE = mae,
    Pearson = pearson,
    Spearman = spearman
  )
}

# Helper for pooled metrics
calc_pooled_metrics <- function(obs, pred) {
  keep <- complete.cases(obs, pred)

  obs  <- obs[keep]
  pred <- pred[keep]

  mse      <- mean((obs - pred)^2)
  rmse     <- sqrt(mse)
  mae      <- mean(abs(obs - pred))
  pearson  <- cor(obs, pred, method = "pearson")
  spearman <- cor(obs, pred, method = "spearman")

  sse <- sum((obs - pred)^2)
  sst <- sum((obs - mean(obs))^2)
  r2  <- 1 - (sse / sst)

  data.frame(
    n = length(obs),
    pooled_MSE = mse,
    pooled_RMSE = rmse,
    pooled_MAE = mae,
    pooled_Pearson = pearson,
    pooled_Spearman = spearman,
    pooled_R2 = r2
  )
}

# Set up parallel backend
cl <- makeCluster(8)
registerDoParallel(cl)

# Run one geodist-only LOPOCV pass in parallel
geodist_results <- foreach(
  i = seq_along(sites),
  .packages = c("dplyr", "randomForest")
) %dopar% {

  site <- sites[i]

  # Identify test/train indices
  test_ids  <- V.table$id[V.table$Var1 == site | V.table$Var2 == site]
  test_idx  <- which(V.table$id %in% test_ids)
  train_idx <- which(!V.table$id %in% test_ids)

  train_df <- V.model_geodist[train_idx, ]
  test_df  <- V.model_geodist[test_idx, ]

  # Fit geodist-only RF model
  set.seed(7000 + i)
  geodist_rf_model <- randomForest(
    CSEdistance ~ .,
    data = train_df,
    ntree = 500
  )

  # Save model immediately
  saveRDS(
    geodist_rf_model,
    file.path(geodist_model_dir, sprintf("rf_model_geodist_fold_%02d.rds", i))
  )

  # Observed values
  train_obs <- train_df$CSEdistance
  test_obs  <- test_df$CSEdistance

  # Predictor data
  train_x <- train_df %>% dplyr::select(-CSEdistance)
  test_x  <- test_df %>% dplyr::select(-CSEdistance)

  # Neutralize sampling density before prediction
  train_x$samp_20km_mean <- geodist_samp_neutral
  test_x$samp_20km_mean  <- geodist_samp_neutral

  # Predict
  pred_train <- predict(geodist_rf_model, newdata = train_x)
  pred_test  <- predict(geodist_rf_model, newdata = test_x)

  # Calibrate test predictions using training predictions
  calibration <- calibrate_pred(
    obs_train = train_obs,
    pred_train = pred_train,
    pred_test = pred_test,
    return_fit = TRUE
  )
  
  pred_test_cal <- calibration$pred_cal
  calibration_fit <- calibration$fit

  saveRDS(
    calibration_fit,
    file.path(geodist_model_dir, sprintf("geodist_calibration_fit_fold_%02d.rds", i))
  )

  # Calculate fold-level metrics
  fold_metrics <- calc_metrics(test_obs, pred_test_cal) %>%
    mutate(
      fold = i,
      site = site
    ) %>%
    dplyr::select(fold, site, everything())

  # Save predictions for this fold
  fold_predictions <- data.frame(
    fold = i,
    site = site,
    id = V.table$id[test_idx],
    obs = test_obs,
    pred = pred_test_cal
  )

  # Save intermediate fold files immediately
  utils::write.csv(
    fold_metrics,
    file.path(geodist_model_dir, sprintf("geodist_metrics_fold_%02d.csv", i)),
    row.names = FALSE
  )

  utils::write.csv(
    fold_predictions,
    file.path(geodist_model_dir, sprintf("geodist_predictions_fold_%02d.csv", i)),
    row.names = FALSE
  )

  saveRDS(
    fold_metrics,
    file.path(geodist_model_dir, sprintf("geodist_metrics_fold_%02d.rds", i))
  )

  saveRDS(
    fold_predictions,
    file.path(geodist_model_dir, sprintf("geodist_predictions_fold_%02d.rds", i))
  )

  # Return file paths only
  list(
    metrics_file = file.path(
      geodist_model_dir,
      sprintf("geodist_metrics_fold_%02d.rds", i)
    ),
    predictions_file = file.path(
      geodist_model_dir,
      sprintf("geodist_predictions_fold_%02d.rds", i)
    )
  )
}

# Stop cluster
stopCluster(cl)

# Reload intermediate files
geodist_metric_files <- file.path(
  geodist_model_dir,
  sprintf("geodist_metrics_fold_%02d.rds", seq_along(sites))
)

geodist_prediction_files <- file.path(
  geodist_model_dir,
  sprintf("geodist_predictions_fold_%02d.rds", seq_along(sites))
)

# Check that all expected files exist
missing_metric_files <- geodist_metric_files[!file.exists(geodist_metric_files)]
missing_prediction_files <- geodist_prediction_files[!file.exists(geodist_prediction_files)]

if (length(missing_metric_files) > 0) {
  stop(
    "Missing geodist metric files:\n",
    paste(missing_metric_files, collapse = "\n")
  )
}

if (length(missing_prediction_files) > 0) {
  stop(
    "Missing geodist prediction files:\n",
    paste(missing_prediction_files, collapse = "\n")
  )
}

# Combine fold-level outputs
geodist_fold_metrics <- bind_rows(lapply(geodist_metric_files, readRDS))
geodist_pooled_predictions <- bind_rows(lapply(geodist_prediction_files, readRDS))

# Calculate pooled metrics
# Calculate pooled metrics
geodist_pooled_metrics <- calc_pooled_metrics(
  obs = geodist_pooled_predictions$obs,
  pred = geodist_pooled_predictions$pred
)

# Save final combined outputs
utils::write.csv(
  geodist_fold_metrics,
  file.path(results_dir, "geodist_LOPOCV_summary.csv"),
  row.names = FALSE
)

utils::write.csv(
  geodist_pooled_predictions,
  file.path(results_dir, "geodist_LOPOCV_pooled_predictions.csv"),
  row.names = FALSE
)

utils::write.csv(
  geodist_pooled_metrics,
  file.path(results_dir, "geodist_LOPOCV_pooled_summary.csv"),
  row.names = FALSE
)

saveRDS(
  geodist_fold_metrics,
  file.path(results_dir, "geodist_LOPOCV_summary.rds")
)

saveRDS(
  geodist_pooled_predictions,
  file.path(results_dir, "geodist_LOPOCV_pooled_predictions.rds")
)

saveRDS(
  geodist_pooled_metrics,
  file.path(results_dir, "geodist_LOPOCV_pooled_summary.rds")
)
```

## Chunk 8: Visualize observed, spatial, permuted, and geodist-only LOPOCV results
```{r visualize-geodist-baseline-LOPOCV, eval = TRUE}

# Chunk 8: load geodist baseline results and visualize with previous LOPOCV metrics

# Directories
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
input_dir <- "../input"
fig_dir <- file.path(results_dir, "figures_pub")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

# Define region based on SiteMajCluster
Gff <- read.csv(
  file.path(input_dir, "Gff_11loci_allsites_indinfo.txt"),
  header = TRUE,
  sep = "\t"
)

north_sites <- unique(Gff$SiteCode[Gff$SiteMajCluster == "north"])

# 1. Full RF LOPOCV Spearman
metrics_obs <- read.csv(file.path(results_dir, "LOPOCV_summary.csv"))

obs_df <- data.frame(
  site = metrics_obs$site,
  cluster = ifelse(metrics_obs$site %in% north_sites, "north", "south"),
  spearman = metrics_obs$Spearman,
  method = "Observed"
)

# 2. Permuted Spearman null
spearman_null <- read.csv(
  file.path(results_dir, "spearman_LOPOCV_null_matrix.csv"),
  row.names = 1,
  check.names = FALSE
)

perm_df <- data.frame(
  site = rownames(spearman_null),
  cluster = ifelse(rownames(spearman_null) %in% north_sites, "north", "south"),
  spearman_null,
  check.names = FALSE
)

perm_long <- reshape2::melt(
  perm_df,
  id.vars = c("site", "cluster"),
  variable.name = "perm",
  value.name = "spearman"
) %>%
  mutate(method = "Permuted")

# 3. Spatial LOPOCV Spearman
metrics_spatial <- read.csv(
  file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_summary.csv")
)

spatial_df <- data.frame(
  site = metrics_spatial$site,
  cluster = ifelse(metrics_spatial$site %in% north_sites, "north", "south"),
  spearman = metrics_spatial$Spearman,
  method = "Spatial"
)

# 4. Geodist-only baseline LOPOCV Spearman
metrics_geodist <- read.csv(
  file.path(results_dir, "geodist_LOPOCV_summary.csv")
)

geodist_df <- data.frame(
  site = metrics_geodist$site,
  cluster = ifelse(metrics_geodist$site %in% north_sites, "north", "south"),
  spearman = metrics_geodist$Spearman,
  method = "Geodist"
)

# Combine all prediction types
plot_df <- bind_rows(
  obs_df[, c("site", "cluster", "spearman", "method")],
  spatial_df[, c("site", "cluster", "spearman", "method")],
  geodist_df[, c("site", "cluster", "spearman", "method")],
  perm_long[, c("site", "cluster", "spearman", "method")]
)

# Set method and cluster order
plot_df$method <- factor(
  plot_df$method,
  levels = c("Observed", "Spatial", "Geodist", "Permuted")
)

plot_df$cluster <- factor(
  plot_df$cluster,
  levels = c("north", "south")
)

# Plot 1: fold-level Spearman distributions with all four prediction types
p_lopocv_all4 <- ggplot(
  plot_df,
  aes(x = spearman, fill = interaction(method, cluster, sep = "."))
) +

  geom_density(
    data = subset(plot_df, method == "Observed"),
    aes(y = after_stat(count)),
    alpha = 0.6,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df, method == "Spatial"),
    aes(y = after_stat(count)),
    alpha = 0.5,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df, method == "Geodist"),
    aes(y = after_stat(count)),
    alpha = 0.45,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df, method == "Permuted"),
    aes(y = after_stat(count / 100)),
    alpha = 0.3,
    color = NA,
    position = "identity"
  ) +

  scale_fill_manual(
    name = "Prediction type",
    breaks = c(
      "Observed.north", "Observed.south",
      "Spatial.north",  "Spatial.south",
      "Geodist.north",  "Geodist.south",
      "Permuted.north", "Permuted.south"
    ),
    values = c(
      "Observed.north" = "#1f78b4",
      "Observed.south" = "#e66101",
      "Spatial.north"  = "#2A4F6E",
      "Spatial.south"  = "#7A4A2A",
      "Geodist.north"  = "#7BA7C7",
      "Geodist.south"  = "#C79A72",
      "Permuted.north" = "#4A545E",
      "Permuted.south" = "#5A4F49"
    ),
    labels = c(
      "Observed.north" = "RF-predicted North",
      "Observed.south" = "RF-predicted South",
      "Spatial.north"  = "LCP-predicted North",
      "Spatial.south"  = "LCP-predicted South",
      "Geodist.north"  = "Geodist-predicted North",
      "Geodist.south"  = "Geodist-predicted South",
      "Permuted.north" = "Null-predicted North",
      "Permuted.south" = "Null-predicted South"
    )
  ) +

  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  theme(
    #panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title = "Fold-level LOPOCV Spearman by prediction type and cluster",
    x = "Spearman's r",
    y = "Count"
  )

p_lopocv_all4

ggsave(
  filename = file.path(fig_dir, "Fig_LOPOCV_obs_spatial_geodist_permuted.png"),
  plot = p_lopocv_all4,
  width = 7,
  height = 2.8,
  units = "in",
  dpi = 600,
  bg = "white"
)


# Plot 2: fold-level Spearman distributions with geodist-only baseline
plot_df_geodist_baseline <- plot_df %>%
  filter(method %in% c("Observed", "Spatial", "Geodist")) %>%
  droplevels()

p_lopocv_geodist_baseline <- ggplot(
  plot_df_geodist_baseline,
  aes(x = spearman, fill = interaction(method, cluster, sep = "."))
) +

  geom_density(
    data = subset(plot_df_geodist_baseline, method == "Observed"),
    aes(y = after_stat(count)),
    alpha = 0.6,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df_geodist_baseline, method == "Spatial"),
    aes(y = after_stat(count)),
    alpha = 0.5,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df_geodist_baseline, method == "Geodist"),
    aes(y = after_stat(count)),
    alpha = 0.45,
    color = NA,
    position = "identity"
  ) +

  scale_fill_manual(
    name = "Prediction type",
    breaks = c(
      "Observed.north", "Observed.south",
      "Spatial.north",  "Spatial.south",
      "Geodist.north",  "Geodist.south"
    ),
    values = c(
      "Observed.north" = "#1f78b4",
      "Observed.south" = "#e66101",
      "Spatial.north"  = "#2A4F6E",
      "Spatial.south"  = "#7A4A2A",
      "Geodist.north"  = "#7BA7C7",
      "Geodist.south"  = "#C79A72"
    ),
    labels = c(
      "Observed.north" = "RF-predicted North",
      "Observed.south" = "RF-predicted South",
      "Spatial.north"  = "LCP-predicted North",
      "Spatial.south"  = "LCP-predicted South",
      "Geodist.north"  = "Geodist-predicted North",
      "Geodist.south"  = "Geodist-predicted South"
    )
  ) +

  coord_cartesian(xlim = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Fold-level LOPOCV Spearman with geodist-only baseline",
    x = "Spearman's r",
    y = "Count"
  )

p_lopocv_geodist_baseline

ggsave(
  filename = file.path(fig_dir, "Fig_LOPOCV_obs_spatial_geodist_only.png"),
  plot = p_lopocv_geodist_baseline,
  width = 7,
  height = 2.8,
  units = "in",
  dpi = 600
)

# Save high-quality figure
ggsave(
  filename = file.path(fig_dir, "Fig_LOPOCV_obs_spatial_geodist_permuted.png"),
  plot = p_lopocv_all4,
  width = 7,
  height = 2.8,
  units = "in",
  dpi = 600
)

# Plot 3: fold-level Spearman distributions for internal LOPOCV vs geodist-only baseline
plot_df_lopocv_geodist <- plot_df %>%
  filter(method %in% c("Observed", "Geodist")) %>%
  droplevels()

# Set x-axis limits based only on the plotted subset
spearman_range_lopocv_geodist <- range(
  plot_df_lopocv_geodist$spearman,
  na.rm = TRUE
)

spearman_pad_lopocv_geodist <- diff(spearman_range_lopocv_geodist) * 0.05

if (spearman_pad_lopocv_geodist == 0) {
  spearman_pad_lopocv_geodist <- 0.05
}

spearman_xlim_lopocv_geodist <- c(
  spearman_range_lopocv_geodist[1] - spearman_pad_lopocv_geodist,
  spearman_range_lopocv_geodist[2] + spearman_pad_lopocv_geodist
)

p_lopocv_vs_geodist <- ggplot(
  plot_df_lopocv_geodist,
  aes(x = spearman, fill = interaction(method, cluster, sep = "."))
) +

  geom_density(
    data = subset(plot_df_lopocv_geodist, method == "Observed"),
    aes(y = after_stat(count)),
    alpha = 0.6,
    color = NA,
    position = "identity"
  ) +

  geom_density(
    data = subset(plot_df_lopocv_geodist, method == "Geodist"),
    aes(y = after_stat(count)),
    alpha = 0.45,
    color = NA,
    position = "identity"
  ) +

  scale_fill_manual(
    name = "Prediction type",
    breaks = c(
      "Observed.north", "Observed.south",
      "Geodist.north",  "Geodist.south"
    ),
    values = c(
      "Observed.north" = "#1f78b4",
      "Observed.south" = "#e66101",
      "Geodist.north"  = "#4A545E",
      "Geodist.south"  = "#5A4F49"
    ),
    
    labels = c(
      "Observed.north" = "Full model North",
      "Observed.south" = "Full model South",
      "Geodist.north"  = "GeoDist baseline North",
      "Geodist.south"  = "Geodist baseline South"
    )
  ) +

  coord_cartesian(xlim = spearman_xlim_lopocv_geodist) +
  theme_minimal() +
  theme(
    #panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title = "Fold-level LOPOCV Spearman: full RF vs geodist-only baseline",
    x = "Spearman's r",
    y = "Count"
  )

p_lopocv_vs_geodist

ggsave(
  filename = file.path(fig_dir, "Fig_LOPOCV_fullRF_vs_geodist_Spearman.png"),
  plot = p_lopocv_vs_geodist,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600,
  bg = "white"
)

# Load pooled summaries
pooled_obs <- read.csv(file.path(results_dir, "LOPOCV_pooled_summary.csv"))

pooled_spatial <- read.csv(
  file.path(results_dir, "spatial_LOPOCV_LCPsum_k1_pooled_summary.csv")
)

pooled_geodist <- read.csv(
  file.path(results_dir, "geodist_LOPOCV_pooled_summary.csv")
)

# Pooled comparison table
pooled_compare_geodist <- data.frame(
  method = c("Observed", "Spatial", "Geodist"),
  pooled_R2 = c(
    pooled_obs$pooled_R2[1],
    pooled_spatial$pooled_R2[1],
    pooled_geodist$pooled_R2[1]
  ),
  pooled_RMSE = c(
    pooled_obs$pooled_RMSE[1],
    pooled_spatial$pooled_RMSE[1],
    pooled_geodist$pooled_RMSE[1]
  ),
  pooled_MAE = c(
    pooled_obs$pooled_MAE[1],
    pooled_spatial$pooled_MAE[1],
    pooled_geodist$pooled_MAE[1]
  ),
  pooled_Pearson = c(
    pooled_obs$pooled_Pearson[1],
    pooled_spatial$pooled_Pearson[1],
    pooled_geodist$pooled_Pearson[1]
  ),
  pooled_Spearman = c(
    pooled_obs$pooled_Spearman[1],
    pooled_spatial$pooled_Spearman[1],
    pooled_geodist$pooled_Spearman[1]
  )
)

write.csv(
  pooled_compare_geodist,
  file.path(results_dir, "LOPOCV_pooled_observed_spatial_geodist_comparison.csv"),
  row.names = FALSE
)

pooled_compare_geodist

# Fold-level summary by method and cluster
fold_summary_geodist <- plot_df %>%
  filter(method != "Permuted") %>%
  group_by(method, cluster) %>%
  summarise(
    n_sites = n(),
    mean_Spearman = mean(spearman, na.rm = TRUE),
    median_Spearman = median(spearman, na.rm = TRUE),
    sd_Spearman = sd(spearman, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(
  fold_summary_geodist,
  file.path(results_dir, "LOPOCV_fold_summary_observed_spatial_geodist.csv"),
  row.names = FALSE
)

# Load final metrics to view results all in one place
geodist_fold_metrics <- read.csv(file.path(results_dir, "geodist_LOPOCV_summary.csv"))
geodist_fold_metrics
summary(geodist_fold_metrics)
geodist_pooled_metrics <- read.csv(file.path(results_dir, "geodist_LOPOCV_pooled_summary.csv"))
geodist_pooled_metrics
fold_summary_geodist
```
## Chunk 9: Paired empirical sign-permutation test across the 67 sites

This tests whether the real LOPOCV Spearman values are higher than the geodist-only Spearman values, paired by held-out site.
```{r paired-geodist-test-spearman, eval = TRUE}
# Empirical paired permutation test:
# Does the full RF LOPOCV have higher fold-level Spearman than geodist-only?

spearman_compare_geodist <- metrics_obs %>%
  dplyr::select(site, Spearman_obs = Spearman) %>%
  inner_join(
    metrics_geodist %>%
      dplyr::select(site, Spearman_geodist = Spearman),
    by = "site"
  ) %>%
  mutate(
    diff_obs_minus_geodist = Spearman_obs - Spearman_geodist
  )

# Observed mean paired difference
obs_mean_diff_spearman <- mean(
  spearman_compare_geodist$diff_obs_minus_geodist,
  na.rm = TRUE
)

# Empirical sign-permutation test
# Randomly flips the sign of each paired site-level difference
set.seed(12345)
n_perm_geodist <- 10000

perm_mean_diff_spearman <- replicate(
  n_perm_geodist,
  {
    signs <- sample(c(-1, 1), size = nrow(spearman_compare_geodist), replace = TRUE)
    mean(spearman_compare_geodist$diff_obs_minus_geodist * signs, na.rm = TRUE)
  }
)

# One-sided p-value: observed full model improvement greater than geodist baseline
p_empirical_geodist_spearman <- (
  sum(perm_mean_diff_spearman >= obs_mean_diff_spearman, na.rm = TRUE) + 1
) / (n_perm_geodist + 1)

cat("Observed mean Spearman, full RF:",
    round(mean(spearman_compare_geodist$Spearman_obs, na.rm = TRUE), 4), "\n")

cat("Observed mean Spearman, geodist-only:",
    round(mean(spearman_compare_geodist$Spearman_geodist, na.rm = TRUE), 4), "\n")

cat("Mean paired Spearman difference, full RF - geodist:",
    round(obs_mean_diff_spearman, 4), "\n")

cat("Empirical one-sided p-value:",
    round(p_empirical_geodist_spearman, 4), "\n")

# Save test results
geodist_spearman_empirical_test <- data.frame(
  metric = "Spearman",
  comparison = "Full RF LOPOCV vs geodist-only LOPOCV",
  n_sites = nrow(spearman_compare_geodist),
  mean_full_RF = mean(spearman_compare_geodist$Spearman_obs, na.rm = TRUE),
  mean_geodist = mean(spearman_compare_geodist$Spearman_geodist, na.rm = TRUE),
  mean_difference = obs_mean_diff_spearman,
  empirical_p_one_sided = p_empirical_geodist_spearman,
  n_permutations = n_perm_geodist
)

write.csv(
  geodist_spearman_empirical_test,
  file.path(results_dir, "LOPOCV_full_vs_geodist_Spearman_empirical_pvalue.csv"),
  row.names = FALSE
)

geodist_spearman_empirical_test
```

```{r del, include = FALSE, eval = FALSE}
## TODO: Spatial eval of geodist with least-cost-paths framework

# Paired empirical sign-permutation test across the 67 sites?
# 
# This would test whether the LOPOCV least-cost-path-based estimates Spearman 
# values are higher than the geodist-only least-cost-path-based estimates 
# Spearman values, paired by held-out site. 

## Delete RDS of permutation test when completely done

unlink(perm_model_dir, recursive = TRUE)
```


