---
title: "Environmental correlations with final model (opening the RF black box)"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../scripts_knitted_md", 
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # Saves PNGs to the root /figures folder
  fig.path = "../figures/knitted_mds/" 
)
```
  
# Setup  

RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 16 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-np (allows multiple simultaneous jobs automatically now)  
- **Memory per job:** 400G (cluster limit: 1000G total; avoid exceeding half)  

# Outline:

Pearson's correlation in a sliding window, of the current variable versus full model, add this to figure with direction of change

Top less-correlated predictors:
BIO3, BIO4, BIO7, BIO11S, BIO13

Figure X. Local Pearson correlations between predicted CSE and selected environmental predictor rasters, calculated in a sliding window with neighborhood size = 21. Positive values indicate local areas where higher predictor values are associated with higher predicted CSE; negative values indicate areas where higher predictor values are associated with lower predicted CSE.

# Color Pallet
```{r colors}
# Connectivity (low = yellow, high = dark purple/red)
connectivity_colors <- colorRampPalette(
  c("#FFFF00", "#FFA500", "#FF4500", "#700E40", "#2E003E")
)(100)

# Suitability (low = light, high = dark blue/purple)
suitability_colors <- colorRampPalette(
  c("white", "#8db4c9", "#3c94c8", "#55306a")
)(100)

# Diverging (for correlations or deltas)
diverging_colors <- colorRampPalette(
  c("blue4", "deepskyblue3", "white", "gold", "darkorange", "#924532")
)(100)
```

# Chunk 1

Reproduce the projected raster from the final tuned mean-based RF model using the same logic as script 07: load the model, load env_stack.grd, neutralize samp_20km, set pix_dist to 1, and predict the raster.

```{r predict-final-cse-raster, warning=FALSE, message=FALSE}
library(raster)
library(randomForest)

# paths
data_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results"

# load final tuned RF model
rf_mean_full_tuned <- readRDS(file.path(results_dir, "rf_mean_full_tuned.rds"))

# load environmental stack
env <- stack(file.path(data_dir, "processed", "env_stack.grd"))

# neutralize sampling bias layer
env$samp_20km <- 1.027064e-11

# set projection geographic distance to 1 km
env$pix_dist <- setValues(env[[1]], 1)

# predict current CSE raster
predicted_cse <- predict(env, rf_mean_full_tuned, type = "response")

# optional save
# writeRaster(predicted_cse,
#             filename = file.path(results_dir, "fullRF_CSE_resistance.tif"),
#             format = "GTiff",
#             overwrite = TRUE)

plot(predicted_cse, main = "Predicted CSE")

```
# Chunk 2
```{r local-correlation-top-predictors, warning=FALSE, message=FALSE}
library(raster)
library(sf)

# select top focal predictors from the same current env stack
pred_names <- c("BIO3", "BIO4", "BIO7", "BIO11S", "BIO13")
pred_stack <- env[[pred_names]]

# local Pearson correlations with sliding window
# Bishop used corLocal() for this kind of post hoc map
# ngb = 21 matches the neighborhood size described in the Bishop figure caption
cor_stack <- stack(lapply(pred_names, function(v) {
  corLocal(predicted_cse, pred_stack[[v]], ngb = 21, method = "pearson")
}))

names(cor_stack) <- paste0("cor_", pred_names)

# optional summary table for quick interpretation
cor_summary <- data.frame(
  variable = pred_names,
  mean_r = sapply(1:nlayers(cor_stack), function(i) {
    cellStats(cor_stack[[i]], stat = "mean", na.rm = TRUE)
  }),
  median_r = sapply(1:nlayers(cor_stack), function(i) {
    cellStats(cor_stack[[i]], stat = "median", na.rm = TRUE)
  })
)

print(cor_summary)

# diverging palette centered on zero
cor_cols <- colorRampPalette(c("blue4", "deepskyblue3", "white", "gold", "darkorange", "#924532"))(100)

# plot
par(mfrow = c(2, 3), mar = c(3, 3, 2, 4))

for (i in 1:nlayers(cor_stack)) {
  plot(cor_stack[[i]],
       col = cor_cols,
       zlim = c(-1, 1),
       main = gsub("cor_", "", names(cor_stack)[i]),
       axes = FALSE,
       box = FALSE,
       legend.args = list(text = "Pearson's r", side = 4, line = 2, cex = 0.8))

  # add these only if they already exist in your environment
  if (exists("lakes")) {
    plot(st_geometry(lakes), col = "black", border = NA, add = TRUE)
  }
  if (exists("uganda_cropped")) {
    plot(st_geometry(uganda_cropped), border = "black", lwd = 0.5, add = TRUE)
  } else if (exists("uganda")) {
    plot(st_geometry(uganda), border = "black", lwd = 0.5, add = TRUE)
  }
}

plot.new()
par(mfrow = c(1, 1))
```
