---
title: "Least-cost-paths across full RF model resistance surface"
author: "Norah Saarman"
date: "2025-06-23"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
RStudio Configuration:  
- **R version:** R 4.4.0 (Geospatial packages)  
- **Number of cores:** 8 (up to 32 available)   
- **Account:** saarman-np  
- **Partition:** saarman-shared-np (allows multiple simultaneous jobs)  
- **Memory per job:** 200G (cluster limit: 1000G total; avoid exceeding half)    

# Setup
```{r libraries, warning=FALSE, results=FALSE, message=FALSE}
# load only required packages
library(doParallel)
library(foreach)
library(raster)
library(gdistance)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

# Define Paths to directories
data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/"
output_dir <- paste0(results_dir,"lopocv_spatial")
dir.create(output_dir, showWarnings = FALSE)

# define coordinate reference system
crs_geo <- 4326     # EPSG code for WGS84
```

# Inputs
  - `../input/Gff_11loci_68sites_cse.csv` - Combined CSE table with coordinates (long1, lat1, long2, lat2)
  - `../results_dir/fullRF_CSE_resistance.tif` - Final full model projected resistance surface
  - `../data_dir/processed/lake_binary.tif` # binary lake mask (1 = water, 0 = land)   
# Outputs  
  - `../results_dir/LC_paths_fullRF.shp"` - 
  
# Step 1: Prepare least-cost paths from full resistance surface
**NOTE:** eval = FALSE so that it skips on knit
```{r paths, eval=FALSE}
# Set number of cores and register cluster
n_cores <- 8
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Load full model prediction raster as resistance surface
resistance_rast <- raster(paste0(results_dir, "fullRF_CSE_resistance.tif"))

# Load lake mask and assign high CSE (e.g., 2)
lake_mask <- raster(file.path(data_dir, "processed", "lake_binary.tif"))
resistance_rast[lake_mask[,] == 1] <- 2

# Convert resistance raster to transition object
tr <- transition(1/resistance_rast, transitionFunction = mean, directions = 16)
tr_corr <- geoCorrection(tr, type = "c")

# Load V.table (after filtering)
V.table_full <- read.csv("../input/Gff_cse_envCostPaths.csv")
V.table <- V.table_full %>%
  filter(Var1 != "50-KB", Var2 != "50-KB") %>%
  filter(Pop1_cluster == Pop2_cluster) %>%
  mutate(id = paste(Var1, Var2, sep = "_"))

# Get unique site coordinates
sites_coords <- V.table %>%
  dplyr::select(Site = Var1, lon = long1, lat = lat1) %>%
  distinct()

sites_df <- st_as_sf(sites_coords, coords = c("lon", "lat"), crs = crs_geo) %>%
  st_transform(crs(resistance_rast))

# Create named SpatialPoints
sites_sp <- as(sites_df, "Spatial")
site_index <- setNames(seq_len(nrow(sites_df)), sites_df$Site)

# Get list of site pairs to extract
site_pairs <- V.table %>%
  dplyr::select(Var1, Var2, id) %>%
  distinct()

# Parallel least-cost path computation with metadata
paths_list <- foreach(k = 1:nrow(site_pairs), .packages = c("gdistance", "sp", "sf")) %dopar% {
  i <- site_index[site_pairs$Var1[k]]
  j <- site_index[site_pairs$Var2[k]]
  
  path <- tryCatch({
    shortestPath(tr_corr, sites_sp[i,], sites_sp[j,], output = "SpatialLines")
  }, error = function(e) NULL)

  if (!is.null(path)) {
    path_sf <- st_as_sf(path)
    path_sf$Var1 <- site_pairs$Var1[k]
    path_sf$Var2 <- site_pairs$Var2[k]
    path_sf$id   <- site_pairs$id[k]
    return(path_sf)
  } else {
    return(NULL)
  }
}

# Stop cluster
stopCluster(cl)

# Combine list into single sf object
paths_sf <- do.call(rbind, paths_list)
st_crs(paths_sf) <- st_crs(resistance_rast)
paths_sf <- st_transform(paths_sf, crs = st_crs(crs_geo))

# Join CSEdistance from V.table into lcp_sf
paths_sf <- left_join(paths_sf, V.table[, c("id", "CSEdistance")], by = "id")
paths_sf <- paths_sf %>%
  rename(CSE = CSEdistance) # note rename!!!

# Save to shapefile
st_write(paths_sf, file.path(results_dir, "LC_paths_fullRF.shp"), delete_layer = TRUE)

```
# Visualize Least Cost Paths
```{r map-paths}
# Load shapefile of least-cost paths
lcp_sf <- st_read(file.path(results_dir, "LC_paths_fullRF.shp")) 
lcp_sf <- st_transform(lcp_sf, crs = st_crs(crs_geo))

# Load Uganda country boundary and lakes
uganda <- ne_countries(scale = "medium", country = "Uganda", returnclass = "sf") %>%
  st_transform(crs = st_crs(crs_geo))
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>%
  st_transform(st_crs(crs_geo))

# (Optional) Load resistance raster for background
resistance_rast <- raster(file.path(results_dir, "fullRF_CSE_resistance.tif"))

xlim <- c(28.6, 35.4)
ylim <- c(-1.500000 , 4.733333)

# Basic plot with paths and country
ggplot() +
  geom_raster(data = as.data.frame(rasterToPoints(resistance_rast)) %>%
                setNames(c("x", "y", "resist")),
              aes(x = x, y = y, fill = resist)) +
  scale_fill_viridis_c(option = "C", na.value = "transparent") +
  geom_sf(data = lakes, fill = "gray80", color = NA) +
  geom_sf(data = uganda, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = lcp_sf, color = "dark blue", size = 0.4, alpha = 0.8) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Least-Cost Paths on Resistance Surface",
       x = "Longitude", y = "Latitude", fill = "Resistance") 

```