---
title: "09_EEMS_results"
author: "Norah Saarman"
date: "2026-03-05"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EEMS input:

**Path:** `/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/input`

EEMS_params.ini
EEMS.coord
EEMS.ind2deme
EEMS.outer
EEMS.sites

### Basic params file
datapath = /uufs/chpc.utah.edu/common/home/u6036559/git/uganda-tsetse-LG/input/EEMS
mcmcpath = /uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run3
nIndiv = 2374
nSites = 11
diploid = 1
nDemes = 200
numMCMCIter = 300000
numBurnIter = 100000
numThinIter = 999
seed = 234


## EEMS results:

**Run 1 Path:** `/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run1`
200 demes, 49 observed/populated demes, seed = 1772671234

**Run 2 Path:** `/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run2`
200 demes, 49 observed/populated demes, seed = 2

**Run 3 Path:** `/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run2`
200 demes, 49 observed/populated demes, seed = 234

Install packages for plotting output in R:
remotes::install_github("dipetkov/rEEMSplots2")

# Load packages:
```{r packages}
library(reemsplots2)
library(fields)
library(RColorBrewer)
library(maps)
library(viridis)
library(sf)
library(viridis)
library(dplyr)
library(terra)
library(sf)
library(classInt)
library(ggplot2)
```

# Set directories
```{r set_dir}

data_dir  <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/data"
input_dir <- "../input"
results_dir <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results"

run1 <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run1"
run2 <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run2"
run3 <- "/uufs/chpc.utah.edu/common/home/saarman-group1/uganda-tsetse-LG/results/EEMS_results/run3"
```

# Make plots of the standard surfaces

## set up colors
```{r plot_setup}
# Load shape files with sf
lakes   <- st_read(file.path(data_dir, "raw/ne_10m_lakes.shp"), quiet = TRUE)
uganda  <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf", continent = "Africa")

# make sure everything is same CRS as the EEMS plot (usually WGS84)
lakes_sf  <- st_transform(lakes, 4326)
uganda_sf <- st_transform(uganda, 4326)

# Custom palette #1 based on Bishop et al.
connectivity_colors <- (colorRampPalette(c("#c5c3a7","#ffe60c","#c79620", "#934334","#802839", "#6f0f40"))(100)) #"#d6ce7c"
suitability_colors  <- colorRampPalette(c("lightgray", "#7fabb9","#3c94c8","#595379","#54306a"))(100)  #"#617189"

```

## reemsplots2 package plots with automated functions
```{r reemsplots2}
library(reemsplots2)

# Make plots with automated functions colors = connectivity
plots1m <- make_eems_plots(
  mcmcpath   = run1,
  longlat    = TRUE,
  eems_colors = connectivity_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)
plots2m <- make_eems_plots(
  mcmcpath   = run2,
  longlat    = TRUE,
  eems_colors = connectivity_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)
plots3m <- make_eems_plots(
  mcmcpath   = run3,
  longlat    = TRUE,
  eems_colors = connectivity_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)

# Make plots with automated functions colors = suitability
plots1q <- make_eems_plots(
  mcmcpath   = run1,
  longlat    = TRUE,
  eems_colors = suitability_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)
plots2q <- make_eems_plots(
  mcmcpath   = run2,
  longlat    = TRUE,
  eems_colors = suitability_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)
plots3q <- make_eems_plots(
  mcmcpath   = run3,
  longlat    = TRUE,
  eems_colors = suitability_colors,
  add_outline = TRUE,
  add_demes   = TRUE
)
```

## (i) posterior mean migration surface
```{r migration, fig.width=7, fig.height=10}
plots1m$mrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)

plots2m$mrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)

plots3m$mrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)
```

## (ii) posterior mean diversity surface
```{r diversity, fig.width=7, fig.height=10}
plots1q$qrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)

plots2q$qrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)

plots3q$qrates01 +
  geom_sf(data = lakes_sf, fill = "black", color = NA, inherit.aes = FALSE) +
  geom_sf(data = uganda_sf, fill = NA, color = "black", linewidth = 0.25, inherit.aes = FALSE)
```

## (iii) fitted versus observed distance between demes
```{r distance, fig.width=7, fig.height=7}
plots1q$rdist01
plots2q$rdist01
plots3q$rdist01
```

## (iv) log posterior MCMC chain convergence
```{r mcmc, fig.width=7, fig.height=7}
plots1q$pilogl01
plots2q$pilogl01
plots3q$pilogl01
```
